---
title             : "DateLife: leveraging databases and analytical tools to reveal the dated Tree of Life"
shorttitle        : "DateLife: revealing the dated Tree of Life"

author:
  - name          : "Luna L. Sánchez Reyes"
    affiliation   : "1,2"
    corresponding : yes    # Define only one corresponding author
    address       : ""
    email         : "sanchez.reyes.luna@gmail.com"
    role:         # Contributorship roles (e.g., CRediT, https://casrai.org/credit/)
      - Data curation
      - Investigation
      - Software
      - Visualization
      - Validation
      - Writing - Original Draft Preparation
      - Writing - Review & Editing
  - name          : "Emily Jane McTavish"
    affiliation   : "1"
    role:
      - Resources
      - Software
      - Writing - Review & Editing
  - name          : "Brian O'Meara"
    affiliation   : "2"
    role:
      - Conceptualization
      - Funding acquisition
      - Methodology
      - Resources
      - Software
      - Supervision
      - Writing - Review & Editing

affiliation:
  - id            : "1"
    institution   : "University of California, Merced"
  - id            : "2"
    institution   : "University of Tennessee, Knoxville"

authornote: |
  School of Natural Sciences, University of California, Merced, Science and Engineering Building 1.

  Department of Ecology and Evolutionary Biology, University of Tennessee, Knoxville, 425 Hesler Biology Building, Knoxville, TN 37996, USA.

abstract: |
  Time of evolutionary origin is fundamental for research in the natural sciences, as well as for education, science communication and policy.
  Despite an increased availability of fossil and molecular data, and time-efficient analytical techniques, achieving a high-quality reconstruction of time of evolutionary origin as a phylogenetic tree with branch lengths proportional to absolute time (chronogram), is still a difficult and time-consuming task for a majority of interested parties.
  Yet, the amount of published chronograms has increased significantly in the past two decades, and a non-negligeable proportion of these data have been steadily accumulating in public, open databases such as TreeBASE and Open Tree
  of Life, exposing a wealth of expertly-curated and peer-reviewed data on time of evolutionary origin in a programatic and reusable way, for a large quantity and diversity of organisms.
  This trend results from intensive and localized efforts for improving data sharing practices, as well as incentivizing open science in biology. Despite these trends, accessibility for time data is not that good.
  R has become a widely used element of the biological data analyst toolkit. Hence,to improve accessibility of time data we developed an R package that access these data and eases interaction, reanalysis and reuse of it, incorporation of these data into the evolutionary workflow. 


  Here we present `datelife`, a service implemented as an R package and a web site
  (www.datelife.org) for increased accessibility, efficient reuse, summary and reanalysis of expert, peer-reviewed, public data on time of evolutionary origin.

  Main results:
    1. blah
    2. blah
    3. blah

  Results in a general context:
  All source and summary chronograms can be saved in formats that permit easy reuse and reanalysis. Summary and newly generated trees are potentially useful to evaluate evolutionary hypothesis in different areas of research in biology.
  How well this trees work for this purpose still needs to be tested.

  `datelife` will be useful to increase awereness on the existing variation in expert time of divergence data, and might foster exploration of the effect of alternative divergence time hypothesis on the results of analyses, nurturing a culture of more cautious interpretation of evolutionary results.

  <!-- https://tinyurl.com/ybremelq -->

keywords          : "Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations"
wordcount         : "55"

bibliography      : ["paper_references.bib"]

floatsintext      : no
figurelist        : no
tablelist         : no
footnotelist      : no
linenumbers       : yes
mask              : no
draft             : no

documentclass     : "apa6"
classoption       : "man"
output            : papaja::apa6_pdf

nocite: |
  @barker2012going, @barker2015new, @burns2014phylogenetics, @claramunt2015new, @gibb2015new, @Hedges2015, @hooper2017chromosomal, @Jetz2012, @price2014niche
---

```{r setup, include = FALSE}
library("papaja")
r_refs("paper_references.bib")
```

```{r analysis-preferences}
# Seed for random number generation
set.seed(50)
knitr::opts_chunk$set(cache.extra = knitr::rand_seed)
data(opentree_chronograms)
rpackage_available <- FALSE
availability <- ifelse(rpackage_available, "is available", "will be available")
```

# Introduction


# Description
We report how we determined our sample size, all data exclusions (if any), all manipulations, and all measures in the study. <!-- 21-word solution (Simmons, Nelson & Simonsohn, 2012; retrieved from http://ssrn.com/abstract=2160588) -->

# Benchmark

# Example
We used `r cite_r("r-references.bib")` for all our analyses.

# Discussion

# Conclusions

# Availability

`datelife` is free and open source and it can be used through its current website
<http://www.datelife.org/query/>, through its R package, and through Phylotastic's project web portal <http://phylo.cs.nmsu.edu:3000/>.
`datelife`'s website is maintained using RStudio's shiny server and the shiny package open infrastructure, as well as Docker.
`datelife`'s R package stable version `r availability`
for installation from the CRAN repository (<https://cran.r-project.org/package=datelife>)
using the command `install.packages(pkgs = "datelife")` from within R. Development versions
are available from the GitHub repository (<https://github.com/phylotastic/datelife>)
and can be installed using the command `devtools::install_github("phylotastic/datelife")`.


# Supplementary Material

Code used to generate all versions of this manuscript, the biological examples, as well as the benchmark of functionalities are available at [datelifeMS1](https://github.com/LunaSare/datelifeMS1), [datelife_examples](https://github.com/LunaSare/datelife_examples), and [datelife_benchmark](https://github.com/LunaSare/datelife_benchmark) repositories in LLSR's GitHub account.

# Funding

Funding was provided by the US National Science Foundation (NSF) grants ABI-1458603  to Datelife project and DBI-0905606 to the National Evolutionary Synthesis Center (NESCent), and the Phylotastic project Grant ABI-1458572.

# Acknowledgements

We thank colleagues from the O'Meara Lab at the University
of Tennesse Knoxville for suggestions, discussions and software testing.
The late National Evolutionary Synthesis Center (NESCent), which sponsored hackathons
that led to initial work on this project. The team that assembled `datelife`'s first proof of concept: Tracy Heath, Jonathan Eastman, Peter Midford, Joseph Brown, Matt Pennell, Mike Alfaro, and Luke Harmon.
The Open Tree of Life project that provides the open, metadata rich repository of
trees used for `datelife`.
The many scientists who publish their chronograms in an open, reusable form, and
the scientists who curate them for deposition in the Open Tree of Life repository.
The NSF for funding nearly all the above, in addition to the ABI grant that funded this project itself.

\newpage

# References

\begingroup
\setlength{\parindent}{-0.5in}
\setlength{\leftskip}{0.5in}

<div id="refs" custom-style="Bibliography"></div>
\endgroup
