---
title             : "DateLife: leveraging databases and analytical tools to reveal the dated Tree of Life"
shorttitle        : "DateLife: revealing the dated Tree of Life"

author:
  - name          : "Luna L. Sánchez Reyes"
    affiliation   : "1,2"
    corresponding : yes    # Define only one corresponding author
    address       : ""
    email         : "sanchez.reyes.luna@gmail.com"
    role:         # Contributorship roles (e.g., CRediT, https://casrai.org/credit/)
      - Data curation
      - Investigation
      - Software
      - Visualization
      - Validation
      - Writing - Original Draft Preparation
      - Writing - Review & Editing
  - name          : "Emily Jane McTavish"
    affiliation   : "1"
    role:
      - Resources
      - Software
      - Writing - Review & Editing
  - name          : "Brian O'Meara"
    affiliation   : "2"
    role:
      - Conceptualization
      - Funding acquisition
      - Methodology
      - Resources
      - Software
      - Supervision
      - Writing - Review & Editing

affiliation:
  - id            : "1"
    institution   : "University of California, Merced, USA"
  - id            : "2"
    institution   : "University of Tennessee, Knoxville, USA"

authornote: |
  School of Natural Sciences, University of California, Merced, 258 Science and Engineering Building 1, Merced, CA 95340, USA.

  Department of Ecology and Evolutionary Biology, University of Tennessee, Knoxville, 446 Hesler Biology Building, Knoxville, TN 37996, USA.



bibliography      : ["paper_references.bib"]
floatsintext      : no
figurelist        : no
tablelist         : no
footnotelist      : no
linenumbers       : yes
mask              : no
draft             : no

documentclass     : "apa6"
classoption       : "man"
output            :
  papaja::apa6_pdf:
    includes      :
      in_header: "preamble.tex"
      # after_body  : ["tables.tex",
      #                "figures-alt.tex"]

nocite: |
  @barker2013going, @barker2015new, @burns2014phylogenetics, @claramunt2015new, @gibb2015new, @Hedges2015, @hooper2017chromosomal, @Jetz2012, @price2014niche,
  @alstrom2014discovery,
  @barker2004phylogeny,
  @barker2013going,
  @barker2014mitogenomic,
  @barker2015new,
  @beresford2005african,
  @bryson2014diversification,
  @burleigh2015building,
  @burns2014phylogenetics,
  @chaves2013biogeography,
  @claramunt2015new,
  @gibb2015new,
  @hackett2008phylogenomic,
  @Jetz2012,
  @johansson2008phylogenetic,
  @kimball2019phylogenomic,
  @klicka2014comprehensive,
  @lamichhaney2015evolution,
  @lerner2011multilocus,
  @lovette2010comprehensive,
  @moyle2016tectonic,
  @odeen2011evolution,
  @oliveros2019earth,
  @packert2012horizontal,
  @parchman2007coevolution,
  @powell2014comprehensive,
  @price2014niche,
  @pulgarin2013multilocus,
  @selvatti2015paleogene,
  @tietze2013complete,
  @treplin2008molecular,
  @zuccon2012phylogenetic
---

```{r setup, include = FALSE}
library("papaja")
r_refs("paper_references.bib")
embed_figure <- FALSE
```

```{r analysis-preferences}
# Seed for random number generation
set.seed(50)
knitr::opts_chunk$set(cache.extra = knitr::rand_seed)
data(opentree_chronograms, package = "datelife")
rpackage_available <- TRUE
availability <- ifelse(rpackage_available, "is available", "will be available")
load(file = "../data/fringillidae_v2022.01.28.rda")
# to write to text the references of the studies that have at least two Fringillidae species:
write(unique(names(fringillidae_v2022.01.28$dres)), "references-fringillidae.txt")
dres <- fringillidae_v2022.01.28$dres
dquery <- fringillidae_v2022.01.28$dquery
```

\begin{center}
Abstract
\end{center}

  Chronograms –phylogenies with branch lengths proportional to time– represent key data on evolutionary time frame for the study of natural processes in many areas of biological research.
  Chronograms also provide valuable information that can be used for education, science communication, and policy decisions.
  Yet, achieving a high-quality reconstruction of a chronogram is a difficult and resource-consuming task.
  Here we present DateLife, a service implemented as an R package and an R Shiny web application available at www.datelife.org, that provides services for efficient and easy discovery, summary, reuse, and reanalysis of node age data mined from a curated database of expert, peer-reviewed, and openly available chronograms.
    The main DateLife workflow starts with one or more scientific taxon names provided by a user. Names are processed and standardized to a unified taxonomy. Then, DateLife runs a name match across its local chronogram database that is curated from Open Tree of Life's phylogenetic repository, and extracts all chronograms that contain at least two queried taxon names, along with their metadata. 
    Finally, node ages from matching chronograms are mapped upon a chosen tree topology using the congruification algorithm. Congruified node ages are used as secondary calibrations to date the chosen topology, with or without initial branch lengths, using different phylogenetic dating methods such as BLADJ, treePL, PATHd8 and MrBayes.
    We performed a cross-validation test to compare node ages resulting from a DateLife analysis (i.e, phylogenetic dating using secondary calibrations) to those from the original chronograms (i.e, obtained with primary calibrations), and found that DateLife's node age estimates appear generally correlated to those from the original chronograms, with the largest variation in ages occurring around topologically deeper nodes.
  Software such as DateLife can help increase awareness of the existing variation in alternative hypothesis of evolutionary time frame for the same organisms, and can support exploration of the effect of alternative chronogram hypotheses on downstream analyses, providing a framework for a more informed interpretation of evolutionary results.

_Keywords_: Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations.

Word count: `r wordcountaddin::word_count()`

\newpage 
<!-- # Introduction -->
<!-- If you're aiming for syst bio you don't want to start so general Readers know what chronograms and phylogenies are!.-->
Chronograms --phylogenies with branch lengths proportional to time-- provide key data on evolutionary time frame for the study of natural processes in many areas of biological research, such as comparative analysis [@harvey1991comparative; @freckleton2002phylogenetic], developmental biology [@laubichler2009form; @delsuc2018phylogenomic], conservation biology and ecology [@Felsenstein1985a; @Webb2000], historical biogeography [@posadas2006historical], and species diversification [@magallon2001absolute; @Morlon2014].

Building a chronogram is not an easy task.
It requires obtaining and curating a homology hypothesis to construct a phylogeny, selecting and placing appropriate calibrations on the phylogeny using independent age data points from the fossil record or other dated events, and inferring a full dated tree. All of this entails specialized biological training, taxonomic domain knowledge, and a non-negligible amount of research time, computational resources and funding.

Here we present the DateLife project which has the main goal of extracting and exposing age data from published chronograms, making age data readily accessible to the wider community for reuse and reanalysis, research, teaching, and science communication and policy.
<!--  R has become mainstream in the biological-data analyst toolkit. Hence, to improve accessibility of time of evolutionary origin data we developed an R package that facilitates interaction, reanalysis and reuse of it, as well as incorporation of these data into the evolutionary analysis workflow. -->
DateLife's core software application is available as an R package [@sanchez2022], and as an online Rshiny interactive website at www.datelife.org. It features key elements for scientific reproducibility, such as a curated, versioned, open and fully public chronogram database [@mctavish2015phylesystem] that stores data in a computer-readable format [@vos2012nexml]; automated and programmatic ways of accessing and downloading the data in a computer-readable format also [@Stoltzfus2013]; and methods to summarize and compare the data.

\begin{center}
\textsc{Description}
\end{center}

DateLife's core software applications are implemented in the R package `datelife`. Its current stable version -- v`r packageVersion("datelife")`, is available from The Comprehensive R Archive Network (CRAN) repository [@sanchez2022], and relies on functionalities from other biological R packages:
ape [@paradis2004],
bold [@Chamberlain2018],
geiger [@pennell2014geiger],
msa [@bodenhofer2015msa],
paleotree [@Bapst2012a],
phyloch [@Heibl2008],
phylocomr [@Ooms2018],
phytools [@Revell2012],
rotl [@Michonneau2016], and
taxize [@Chamberlain2013; @Chamberlain2018].
Figure \ref{fig:workflow} provides a graphical summary of the three main steps of the DateLife workflow: creating a search query, searching a database, and summarizing results from the search.

```{r fig1, child= if (embed_figure) "fig-workflow.Rmd"}
```

\begin{center}
\emph{Creating a Search Query}
\end{center}

DateLife starts by processing an input consisting of at least one taxon scientific name. Multiple input names can be provided as a comma separated character string or as tip labels on a tree.
If the input is a tree, it can be provided as a classic newick character string [@archie1986newick], or as a "phylo" R object [@paradis2004].
The input tree is not required to have branch lengths, and its topology is used in the summary steps described in the next section.

DateLife processes input scientific names using a Taxonomic Name Resolution Service (TNRS), which increases the probability of correctly finding the queried taxon names in the chronogram database. TNRS detects, corrects and standardizes name misspellings and typos, variant spellings and authorities, and nomenclatural synonyms to a single taxonomic standard [@Boyle2013]. TNRS also allows to correctly choose between homonyms, by considering other taxa provided as input to infer the taxonomic context of the homonym. DateLife implements TNRS using the Open Tree of Life (OpenTree) unified Taxonomy [OTT, @opentreeAPIs; @rees2017automated] as standard, storing taxonomic identification numbers (OTT ids) for further processing and analysis. Other taxonomies currently supported by DateLife are the National Center of Biotechnology Information (NCBI) taxonomic database [@schoch2020ncbi], the Global Biodiversity Information Facility (GBIF) taxonomic backbone [@gbif2022taxonomy], and the Interim Register of Marine and Non-marine Genera (IRMNG) database [@rees2017irmng].
<!-- We are using OpenTree's TNRS [BCO: yes], Boyle described TNRS first, right?  [BCO: earliest I can find, but feels like there should be much older work]-->

Besides binomial species names, DateLife accepts scientific names from any inclusive taxonomic group (e.g., genus, family, tribe), as well as subspecific taxonomic variants (e.g., subspecies, variants, strains). If a taxon name belongs to an inclusive taxonomic group, DateLife has two alternative behaviors defined by the "get species from taxon" flag. If the flag is active, DateLife retrieves all species names within the taxonomic group from the standard taxonomy of choice, and adds them to the search query. In this case, subspecific variants are excluded.
If the flag is inactive, DateLife excludes any taxon names above the species level from the search query.
Species and subspecific variant names are processed and searched as provided by the user.
The processed taxon names are saved as an R object of a newly defined class, \texttt{datelifeQuery}, that is used in the following steps. This object contains the input names standardized to a taxonomy of choice (OTT by default), the corresponding OTT id numbers, and the topology of an input tree, if one was provided.

\begin{center}
\emph{Searching a Chronogram Database}
\end{center}

At the time of writing of this manuscript 
<!-- (Jun 22, 2022), -->
(`r format(Sys.time(), "%b %d, %Y")`),
DateLife's chronogram database latest version consist of `r format(length(opentree_chronograms$trees), big.mark=",")` chronograms published in `r length(unique(opentree_chronograms$studies))` different studies. It is curated from OpenTree's phylogenetic database, the Phylesystem, which constitutes an open source of expert and peer-reviewed phylogenetic knowledge with rich metadata [@mctavish2015phylesystem], which allows automatic and reproducible assembly of our chronogram database. Datelife's chronogram database is navigable as an R data object within the `datelife` R package.

A unique feature of the Phylesystem is that any user can add new published, state-of-the-art chronograms any time, through their curator application (https://tree.opentreeoflife.org/curator). As chronograms are added to Phylesystem, they can be incorporated into the chronogram database of the `datelife` R package, which is currently manually updated as new chronogram data is added to Phylesystem. The updated database is assigned a new version number, followed by a package release on CRAN.
Users can directly implement `datelife` functions to trigger an update of the local chronogram database, to incorporate any new chronograms to their DateLife analysis before an official database update is released on CRAN. <!--BCO: What is immediately? Also, it's not just uploading -- they have to include metadata on what brlen mean, maybe? indicate the ingroup/outgroup, do TNRS within OToL, right? Maybe this has improved. I remember being frustrated trying to load in giant trees for datelife use b/c I had to click on the ingroup node.
Luna: Very true, it is not that easy, I added the curation step to hint that there are more steps than simply uploading the chronogram.-->

A DateLife search is implemented by matching processed taxon names provided by the user to tip labels in the chronogram database. Chronograms with at least two matching taxon names on their tip labels are identified and pruned down to preserve only the matched taxa.
These matching pruned chronograms are referred to as source chronograms.
Total distance (in units of millions of years) between taxon pairs within each source chronogram are stored as a patristic distance matrix (Figure \ref{fig:workflow}).
The matrix format speeds up extraction of pairwise taxon ages of any queried taxa, as opposed to searching the ancestor node of a pair of taxa in a "phylo" object or newick string.
Finally, the patristic matrices are associated to the study citation where the original chronogram was published, and stored as an R object of the newly defined class \texttt{datelifeResult}.

\begin{center}
\emph{Summarizing Search Results}
\end{center}

Summary information is extracted from the \texttt{datelifeResult} object to inform decisions for subsequent steps in the analysis workflow. Basic summary information available to the user is:

1. The matching pruned chronograms as newick strings or "phylo" objects.
2. The ages of the root of all source chronograms. These ages can correspond to the age of the most recent common ancestor (mrca) of the user's group of interest if the source chronograms have all taxa belonging to the group. If not, the root corresponds to the mrca of a subgroup withing the group of interest.
3. Study citations where original chronograms were published.
4. A report of input taxon names matches across source chronograms.
5. The source chronogram(s) with the most input taxon names.
6. Various single summary chronograms resulting from summarizing age data, generated using the methodology described next.

\begin{center}
\emph{Choosing a Topology}
\end{center}

DateLife requires a tree topology to summarize age data upon.
We recommend that users provide as input a tree topology from the literature, or one of their own making. If no topology is provided, DateLife automatically extracts one from the OpenTree synthetic tree, a phylogeny currently encompassing 2.3 million taxa across all life, assembled from 1, 239 published phylogenetic trees and OpenTree's unified Taxonomy, OTT [@opentreeoflife2019synth].
Alternatively, DateLife can combine topologies from source chronograms using a supertree approach [@Criscuolo2006].
To do this, DateLife first identifies the source chronograms that form a grove, roughly, a sufficiently overlapping set of taxa between trees, by implementing definition 2.8 for n-overlap from Ané et al. [-@ane2009groves].
If the source chronograms do not form a grove, the supertree reconstruction will fail.
In rare cases, a group of trees can have multiple groves. By default, DateLife chooses the grove with the most taxa, however, the "criterion = trees" flag allows the user to choose the grove with the most trees instead.
The result is a single summary (or supertree) topology, that combines topologies from source chronograms in a grove.

\begin{center}
\emph{Applying Secondary Calibrations}
\end{center}

Once a topology is chosen, DateLife applies the congruification method [@Eastman2013] that find nodes belonging to the same clade across source chronograms, and then extracts the corresponding node ages from patristic distance matrices stored as a `datelifeResult` object. Note that by definition, these matrices store total distance (time from tip to tip), assuming that the terminal taxa are coeval and occur at the present. Hence, node ages correspond to half the values stored in the `datelifeResult` matrices.
A table of congruified node ages that can be used as calibrations for a dating analysis is stored as a `congruifiedCalibrations` object.

For each congruent node, the pairwise distances that traverse that node are summarized into a single summary matrix using classic summary statistics (i.e., mean, median, minimum and maximum ages), and the Supermatrix Distance Method [SDM; @Criscuolo2006], which deforms patristic distance matrices by minimizing variance and then averaging them.
These single summary taxon pair age matrices are stored as summarized calibrations that can be used as secondary calibrations to date a tree topology - with or without initial branch lengths, using phylogenetic dating methods currently supported within DateLife: BLADJ [@Webb2008; @webb2005phylomatic], MrBayes [@Huelsenbeck2001; @Ronquist2003], PATHd8 [@Britton2007], and treePL [@Smith2012].

\begin{center}
\emph{Dating a Tree Topology}
\end{center}

**_With no branch lengths.--_**
When producing or obtaining a tree with branch lengths for a group of interest is not possible, DateLife can date a topology without branch lengths, obtained from OpenTree or by implementing the supertree approach described above, by implementing the Branch Length Adjuster [BLADJ; @Webb2008; @webb2005phylomatic] algorithm, which requires no initial branch lengths. 
The algorithm starts by fixing ages for nodes with calibration data upon the given topology. Then, it distributes time for nodes with no data evenly between calibrated nodes, minimizing age variance in the resulting chronogram [@Webb2008], which has proven useful for ecological analyses that require age data [@Webb2008].
When there is conflict between ages of calibrated nodes, BLADJ ignores node ages that are older than the age of a parent node.
The BLADJ algorithm requires a root age estimate to run. If there is no information on the age of the root in the chronogram database, users can provide an estimate from the literature. If none is provided, DateLife will not return a dated topology and will provide a warning message along with suggestions on how the user can provide an age for the root so that a fully dated tree can be returned.
<!-- assigns an arbitrary age to the root as 10% older than the oldest age available within the group. -->

In the absence of genetic data, we consider that BLADJ is the most agnostic way to assign ages to nodes with no available data, as it does not require any assumptions on the underlying model of branch length distribution. It is however common practice in the literature to use a birth-death model to assign ages to nodes with no genetic data [@Jetz2012; @rabosky2018inverse; @smith2018constructing]. To do so, DateLife implements MrBayes  [@Huelsenbeck2001; @Ronquist2003], using nodes with published age data as calibration priors on a fixed topology, a simple birth-death model with parameters that can be determined by the user, and no genetic data.

**_Dating a tree with branch lengths.--_**
Relative branch lengths provide key information for phylogenetic dating, specifically for nodes without any calibration data available. 
While, using initial branch lengths data is the golden standard for phylogenetic dating analyses, producing such requires assembling and curating a homology hypothesis, and choosing and implementing a method for phylogenetic inference.
DateLife implements a workflow to streamline this process by leveraging on genetic data from the Barcode of Life Data System, BOLD [@ratnasingham2007bold] to obtain markers for input taxa.
<!--DNA sequences obtained from BOLD are short (between 500-800 bp) and display little variation. -->
By default, mined genetic sequences are aligned with MUSCLE [@edgar2004muscle] using functions from the msa R package [@bodenhofer2015msa]. Alternatively, sequences can be aligned with MAFFT [@katoh2009multiple], using functions from the ape R package [@paradis2004].
The BOLD sequence alignment is then used to obtain initial branch lengths with the accelerated transformation (ACCTRAN) parsimony algorithm, which resolves ambiguous character optimization by assigning changes along branches of the tree as close to the root as possible [@agnarsson2008acctran], resulting in older nodes [@forest2005teasing]. The parsimony branch lengths are then optimized using Maximum Likelihood, given the alignment, the topology and a simple Jukes-Cantor model, producing a BOLD tree with branch lengths proportional to expected number of substitutions per site. Both parsimony and ML optimizations are done with functions from the `phangorn` package [@schliep2011phangorn]. 
The BOLD workflow is currently only available on DateLife's R package and not on the web application.

Phylogenetic dating methods supported in DateLife that incorporate branch length information from the input topology in combination with the secondary calibrations include:
PATHd8, a non-clock, rate-smoothing method to date trees [@Britton2007];
treePL [@Smith2012], a semi-parametric, rate-smoothing, penalized likelihood dating method [@sanderson2002]; 
and MrBayes [@Huelsenbeck2001; @Ronquist2003], a Bayesian inference program implementing Markov chain Monte Carlo (MCMC) methods to estimate a posterior distribution of model parameters.

\begin{center}
\emph{Visualizing Results}
\end{center}

Finally, users can save all source and summary chronograms in formats allowing for reuse and reanalysis, such as newick and the R "phylo" format. Input and summary chronograms can be visualized and compared graphically, and users can construct their own graphs using DateLife's chronogram plot generation functions available from the R package `datelifeplot` [@datelifeplot].

<!-- TODO ADD FIGURE: mock example explaining difference between using summary ages and raw ages to calibrate a topology. Then show it in the biological example. This will show what happens if a tree (phylogenetic conflict) or node (age) do not agree.-->

<!-- \newpage -->

\begin{center}
\textsc{Benchmark}
\end{center}

R package `datelife` code speed was tested on an Apple iMac
with one 3.4 GHz Intel Core i5 processor. <!--No multiprocessor tests were performed-->
We registered variation in computing time of query processing and search through the database relative to number of queried taxon names.
Query processing time increases roughly linearly with number of input taxon names, and
increases considerably if Taxonomic Name Resolution Service (TNRS) is activated.
<!-- Defining tnrs here again, bc we only talk about it two times, so it easy to forget it -->
Up to ten thousand names can be processed and searched in less than 30 minutes with the most time consuming settings.
Once names have been processed as described in methods, a name search through the chronogram database can be performed in less than a minute, even with a very large number of taxon names (Fig. \ref{fig:runtime_main}).
<!--Speed of other datelife workflows? NO, bc it reflects speed from other software mainly. But I will put a sentence about it...
Summarizing `datelife` results processing times: shows sdm speed, clustering methods speed;
generating new trees: shows bladj speed;
Adding dates processing time: reflects more mrbayes speed than anything;
get_bold_otol_tree running time: reflects bold downloading data speed.-->
<!--Speed with different number of chronograms in database? NO-->
<!--Speed of web interface and of r package in computers with different capacities? NOOOO-->

```{r fig2, child= if (embed_figure) "fig-benchmark.Rmd"}
```

`datelife`'s code performance was evaluated with a set of unit tests designed and
implemented with the R package testthat [@RCoreTeam2018] that were run both locally
with the devtools package [@RCoreTeam2018], and on a public server using the continuous integration tool of GitHub actions (<https://docs.github.com/en/actions>).
At present, unit tests cover more than 40% of `datelife`'s code (<https://codecov.io/gh/phylotastic/datelife>). <!-- unit test make code robust, verifiable, debuggable, -->
Unit testing helps identify potential issues as code is updated or, more critically, as services code relies upon may change.

\begin{center}
\textsc{Case Studies}
\end{center}

We illustrate the DateLife workflow using a family within the passeriform birds encompassing the true finches, Fringillidae, as case study. On a small example, we analysed 6 bird species, and results from each step of the workflow are shown in Fig. \ref{fig:figure2}. As a second example, we analysed 289 bird species in the family Fringillidae that are included in the NCBI taxonomy. The resulting summary chronogram is shown in Fig. \ref{fig:fringillidages}, and results from previous steps of the workflow are available as Supplementary Figures.
<!--We used `r cite_r("r-references.bib")` for all our analyses.-->

\begin{center}
\emph{A Small Example}
\end{center}

**_Creating a search query.--_**
We chose 6 bird species within the Passeriformes. The sample includes
two species of cardinals:
the black-thighed grosbeak -- *Pheucticus tibialis* and
the crimson-collared grosbeak -- *Rhodothraupis celaeno*;
three species of buntings:
the yellowhammer -- *Emberiza citrinella*,
the pine bunting --  *Emberiza leucocephalos* and
the yellow-throated bunting --  *Emberiza elegans*;
and one species of tanager, the vegetarian finch -- *Platyspiza crassirostris*.
Processing of input names found that *Emberiza elegans* is synonym for *Schoeniclus elegans* in the default reference taxonomy (OTT v3.3, June 1, 2021). For a detailed discussion on the state of the synonym, refer to Avibase [@lepage2004avibase; @lepage2014avibase; @avibase-emberiza].
Discovering this synonym allowed assigning five age data points for the parent node of *Emberiza elegans*, shown as *Schoeniclus elegans* in figure \ref{fig:figure2}A, which would not have had any data otherwise.
<!--\textcolor{red}{EJM: I don't understand what this sentence means}
LLSR: it is related to the synonym found by tnrs processing, if we had used the synonym instead of the one accepted in the database, we would not have discovered that information
LLSR: I moved it here, it makes more sense here, I think!.
-->

<!-- \newpage -->

**_Searching the database.--_**
DateLife used the processed input names to search the local chronogram database and found 9 matching chronograms in 6 different studies (Fig. \ref{fig:figure2}B). Three studies matched five input names [@barker2015new; @Hedges2015; @Jetz2012], one study matched four input names [@hooper2017chromosomal] and two studies matched two input names [@barker2013going; @burns2014phylogenetics]. No studies matched all input names. Together, source chronograms provide 28 unique age data points, covering all nodes on our chosen tree topology to date (Table \ref{tbl:table1}).

<!-- \vspace{3mm} -->

**_Summarizing search results.--_**
DateLife obtained OpenTree's synthetic tree topology for these taxa (Fig. \ref{fig:figure2}C), and congruified and mapped age data to nodes in this chosen topology (Table \ref{tbl:table1}).
The name processing step allowed including five data points for node "n4" (parent of _Schoeniclus elegans_; Fig. \ref{fig:figure2}A) that would not have had any data otherwise due to name mismatch.
Age summary statistics per node were calculated (Table \ref{tbl:table2}) and used as calibrations to date the tree topology using the BLADJ algorithm.
As expected, more inclusive nodes (e.g., node "n1") have more variance in age data than less inclusive nodes (e.g., node "n5").
Summary age data for node "n2" were excluded as final calibration because they are older than age data of the more inclusive node, "n1" (Fig. \ref{fig:figure2}C4).

<!-- \newpage -->

```{r fig-small-example, child= if (embed_figure) "fig-small-example.Rmd"}
```

```{r table1, child= if (embed_figure) "table1.Rmd"}
```

```{r table2, child= if (embed_figure) "table2.Rmd"}
```

\newpage

\begin{center}
\emph{An Example with the Family of True Finches}
\end{center}

**_Creating a query.--_**
To obtain ages for all species within the family of true finches, Fringillidae, we ran a DateLife query using the "get species from taxon" flag,
which gets all recognized species names within a named group from a taxonomy of choice.
Following the NCBI taxonomy, our DateLife query has `r length(dquery$cleaned_names)` Fringillidae species names.
<!--BCO: Feel like we need a number here, though it's good that the clade hasn't gone extinct ;-) -->
This taxon-constrained approach implies that the full DateLife analysis will be performed using a tree topology and ages available for species names from a given taxonomic group, which do not necessarily correspond to a monophyletic group. Users can change this behavior by providing all species names corresponding to a monophyletic group as input for a DateLife search, or a monophyletic tree to construct a DateLife summary.
<!--BCO: it's ambiguous: it gets a topology and age for the larger clade that includes all Fringillidae species? Or for the paraphyletic Fringillidae group?
Luna: Clarified in text, I hope!-->

**_Searching the database.--_**
Next, we used the processed species names in our DateLife query to identify chronograms with at least two Fringillidae species as tip taxa.
The DateLife search identified `r length(dres)` chronograms matching this criteria, published in `r length(unique(names(dres)))` different studies [@barker2013going; @barker2015new; @burns2014phylogenetics; @claramunt2015new; @gibb2015new; @Hedges2015; @hooper2017chromosomal; @Jetz2012; @kimball2019phylogenomic; @oliveros2019earth; @price2014niche; @roquet2014one; @uyeda2017evolution].
Once identified, DateLife pruned these matching chronograms to remove tips that do not belong to the queried taxon names, and transformed these pruned chronograms to pairwise distance matrices, revealing 1, 206 different age data points available for species within the Fringillidae (Supplementary Table S1).

**_Summarizing search results.--_**
The final step entailed congruifying and summarizing the age data available for the Fringillidae species into two single summary chronograms, using two different types of summary ages, median and SDM.
As explained in the "Description" section, a tree topology to summarize age data upon is required.
By default, DateLife uses the topology from OpenTree's synthetic tree that contains all taxa from the search query.
According to OpenTree's synthetic tree, species belonging to the family Fringillidae do not form a monophyletic group (Fig. \ref{fig:fringillidae-topologies}). Hence, a topology containing only the 289 species from the original query was extracted from Open Tree of Life’s synthetic tree v12.3 [@opentreeoflife2019synth].

<!-- \vspace{50mm} -->

```{r fig4, child= if (embed_figure) "fig-topologies.Rmd"}
```

Age data from source chronograms was congruified to OpenTree's topology (Fig. \ref{fig:fringillidae-topologies}B), reducing the age data set to 818 different data points (Supplementary Table S2). For each congruent node, age summary statistics were calculated and used as fixed secondary calibrations over the chosen tree topology, to obtain a fully dated phylogeny with the program BLADJ (Fig. \ref{fig:fringillidages}).

<!-- ????????????????? Say some things about the results!
ANSWER:
Compare median versus SDM:
Median summary chronograms are older and have wider variation in maximum ages than chronograms obtained with SDM.

-->


```{r fig5, child= if (embed_figure) "fig-fringillidae-full-example.Rmd"}
```

\begin{center}
\textsc{Cross-Validation Test}
\end{center}

<!--Data from source chronograms can be used to date tree topologies with no branch lengths, as well as trees with branch lengths as relative substitution rates-->
We performed a cross validation test of a DateLife analysis using the Fringillidae source chronograms obtained above (Supplementary Figs. S2-S20).
We used as inputs for a DateLife analysis all individual tree topologies from each of the 19 source chronograms from 13 studies, treating their node ages as unknown.
We congruified node ages extracted from chronograms from all other studies upon the individual topologies, effectively excluding original ages from each topology. Finally, average node ages per node were applied as secondary calibrations and smoothed with the BLADJ algorithm.
We found that node ages from the original studies, and ages estimated using all other age data available are largely correlated (Fig. \ref{fig:cvXY}).
For five studies, DateLife tended to underestimate ages for topologically deeper nodes (those with many descendant taxa, aka 'closer to the root') relative to the original estimate, and overestimate ages for nodes closer to the tips. Accordingly, root ages are generally older in the original study than estimated using cross-validated ages (Supplementary Fig. S1).
In general, topologically deeper nodes display the largest age variation between node ages from the original chronograms and ages summarized with DateLife.

```{r fig6, child= if (embed_figure) "fig-cross-validation.Rmd"}
```

<!--
%```{r, child = "cross-validation-pathd8.Rmd"}
%```
-->

\begin{center}
\textsc{Discussion}
\end{center}

DateLife makes state-of-the-art data on evolutionary time frame easily accessible for comparison, reuse, and reanalysis, to researchers in all areas of science and with all levels of expertise in the matter. It is an open service that does not require any expert biological knowledge from users --besides the names of the species or group they want to work with, for any of its functionality.

A total of `r format(length(unique(unname(unlist(lapply(opentree_chronograms$trees, "[[", "tip.label"))))), big.mark=",")` unique terminal taxa are represented in DateLife's database.
Incorporation of more chronograms into the database will continue to improve DateLife's services. One option to increase the number of chronograms in the DateLife database is the Dryad data repository.  Methods to automatically mine chronograms from Dryad could be designed and implemented. However, Dryad's metadata system has no information to automatically detect branch length units, and those would still need to be determined manually by a human curator.
We would like to emphasize on the importance of sharing chronogram data, including systematically curated metadata, into open repositories, such as OpenTree's Phylesystem [@mctavish2015phylesystem] for the benefit of the scientific community as a whole.

While it is known that incorporating at least some data on lineage divergence times represents a relevant improvement for testing alternative hypothesis using phylogenetic distance in ecological and conservation biology studies [@Webb2008]. Next, we discuss some particularities of using summarized node age data from a DateLife analysis that users should be aware of.

\begin{center}
\emph{Age Variation in Source Chronograms}
\end{center}

Conflict in estimated ages among alternative studies is common in the literature. See, for example, the robust ongoing debate about crown group age of angiosperms  [@sauquet2021age; @barba2018constraining; @magallon2015metacalibrated; @sanderson2001sources; @ramshaw1972time].
Source chronograms available for the same organisms have potentially been estimated implementing calibrations very differently.
In the Fringillidae example above, the chronograms from @burns2014phylogenetics were inferred using molecular substitution rate estimates across birds [@weir2008calibrating], and have much older age estimates for the same nodes than chronograms that were inferred using fossils as calibrations (Figs. \ref{fig:fringillidages}, \ref{fig:cvXY}; Supplementary Figs. S1, S5).
<!--\textcolor{red}{EJM: MORE ON the FINCH examples and the cross validation here}-->

Different calibration implementations might also imply fundamentally distinct evolutionary hypotheses [@antonelli2017supersmart].
For example, two independent researchers working on the same clade should both carefully select and justify their choices of fossil calibration placement.
Yet, if one researcher concludes that a fossil should calibrate the ingroup of a clade, while another researcher concludes that the same fossil should calibrate the outgroup of the clade, the resulting age estimates will differ, as the placement of calibrations as stem or crown group has been proven to significantly affect time of lineage divergence estimates [@sauquet2013practical].


\begin{center}
\emph{Primary vs Secondary Calibrations}
\end{center}

While most chronograms in DateLife's database are constructed using primary calibrations (molecular substitution rates or ages obtained from the fossil record or geological events), DateLife summarizes chronograms using secondary calibrations (ages coming from other chronograms).
@graur2004reading cautioned on the increased error and uncertainty in estimated ages when using secondary calibrations in dating analyses.
@schenk2016sec showed that, in simulations, divergence times inferred using secondary calibrations are significantly younger than those inferred with primary calibrations, when obtained with Bayesian inference methods, and when priors are implemented in similar ways in both analyses.
Accordingly, the scientific community seems to have more confidence in chronograms obtained from a single analysis, using fossil data as primary sources of calibrations [@schenk2016sec], and using fossils that have been widely discussed and curated as calibrations to date other trees, making sure that all data reflect a coherent evolutionary history [@sauquet2013practical], as for example done by @antonelli2017supersmart. <!-- Will this be implemented in future `datelife` versions? Probably not. -->
There have been attempts to create fossil calibration databases [@ksepka2015fossil], though these still have room to grow. <!--BCO: Polite way to say 'no new additions in four years?' LLSR: lol, yes-->
<!--\textcolor{red}{LLSR: What does our cross validation analysis show in relation to  the effect of using secondary calibrations for a dating analysis that does not make any prior assumptions on the nature of the calibrations themselves? the age difference between primary and secondary calibrations?}-->

It seems that using several (as opposed to just a few) secondary calibrations can provide sufficient information to alleviate or even neutralize potential biases [@sauquet2013practical].
Certainly, further studies are required to fully understand the effect of secondary calibrations on outputs from different tree dating methods, and on downstream analyses. It is possible that secondary calibrations can be safely used with dating methods that do not require setting priors, such as penalized likelihood [@sanderson2003r8s], with methods that do not make any assumptions on the ages and fix them to a node on a tree topology, such as BLADJ [@Webb2008; @webb2005phylomatic], or methods that summarize age data unto a tree topology.

Our cross validation analysis might provide some insight in this regard. When ages are estimated with secondary calibrations, nodes closer to the root do tend to be slightly younger than ages estimated with primary calibrations. However, nodes closer to the tip tend to be older when estimated using secondary calibrations with a dating method that does not make any prior assumptions on the nature of the calibrations themselves (Supplementary Figures S2-S20). The only exception to this was observed on cross validation results of the @burns2014phylogenetics chronogram, which displays much younger node ages when estimated using secondary calibrations (Supplementary Figs. S1, S5).

\begin{center}
\emph{Sumarizing Chronograms}
\end{center}

By default, DateLife currently summarizes all source chronograms that overlap with at least two species names. Users can exclude source chronograms if they have reasons to do so.
Strictly speaking, a good chronogram should reflect the real time of lineage divergence accurately and precisely.
To our knowledge, there are no tested measures to determine independently when a chronogram is better than another. Yet, several characteristics of the data used for dating analyses, as well as from the output chronogram itself, could be used to score the quality of source chronograms.

Some measures that have been proposed are the proportion of lineage sampling and the number of calibrations used [@magallon2010using; @magallon2015metacalibrated].
Some characteristics that are often cited in published studies as a measure of improved age estimates as compared to previously published estimates are: quality of alignment (missing data, GC content), lineage sampling (strategy and proportion), phylogenetic and dating inference method, number of fossils used as calibrations, support for nodes and ages, and magnitude of confidence intervals.

DateLife provides an opportunity to capture concordance and conflict among date estimates, which can also be used as a metric for chronogram reliability.
Its open database of chronograms allows other researchers to do such analyses themselves reproducibly, and without needing permission. Though, of course, they should follow proper citation practices, especially for the source chronogram studies.

The exercise of summarizing age data from across multiple studies provides the opportunity to work with a chronogram that reflects a unified evolutionary history for a lineage, by putting together evidence from different hypotheses.
The largest, and taxonomically broadest chronogram currently available from OpenTree was constructed summarizing age data from 2,274 published chronograms using NCBI's taxonomic tree as backbone [@Hedges2015].
<!--BCO: I just checked: TIMETREEE NOW ALLOWS SOURCE TREE DOWNLOADS! This is great news. They have a copyright (not legal in US on trees, IMHO) notice on individual trees, but they do say "To download all individual study timetrees, please send an email request to info@timetree.org." -- I think this could be great to do-->
A summarizing exercise may also amplify the effect of uncertainty and errors in source data, and blur parts of the evolutionary history of a lineage that might only be reflected in source chronograms and lost on the summary chronogram [@sauquet2021age].
<!--\textcolor{red}{EJM: ??? it's not clear to me what this last part means! What might you be losing?
LLSR: I changed loosing to blurring. But I will elaborate more on it}
Choosing source chronograms that we are going to keep and the ones that we are going to discard might be key.
\textcolor{red}{LLSR: Example}-->

<!-- Comparison of available age data for a wide range of organisms shown here suggest that discrepancy in age estimates across studies is a widespread phenomenon that requires further attention. -->

<!-- Still, even if all source chronograms have been generated by excellent standards and using similar methods, the evolutionary history they depict might be very different.  -->

\begin{center}
\emph{Effects of Phylogenetic Sampling on Downstream Analyses}
\end{center}

Analysis of species diversification using simulated and empirical phylogenies suggest that using a more completely sampled phylogeny provides estimates that are closer to the true diversification history than when analysing incompletely sampled phylogenies [@cusimano2012new; @chang2020estimating; @sun2020estimating]. 
Ideally, phylogenies should be completed using genetic data, but this is a difficult task to achieve for many biological groups.
Hence, DateLife's workflow features different ways of assigning node ages in the absence of calibrations and branch length information for certain taxa.
Completing a phylogeny using a stochastic birth-death politomy resolver and a backbone taxonomy is a common practice in scientific publications: @Jetz2012, created a chronogram of all 9, 993 bird species, where 67% had molecular data and the rest was simulated; @rabosky2018inverse created a chronogram of 31, 536 ray-finned fishes, of which only 37% had molecular data; @smith2018constructing constructed a chronogram of 353, 185 seed plants where only 23% had molecular data. These stochastically resolved chronograms provide less biased diversification rates estimated with methods that account for sampling fractions [@chang2020estimating; @cusimano2012new], but can also introduce spurious patterns of early bursts of diversification [@cusimano2010slowdowns; @sun2020estimating]. 

Taxonomy based politomy resolvers also introduce topological differences.
The study of macroevolutionary processes largely depends on an understanding of the timing of species diversification events, and different phylogenetic and chronogram hypothesis can provide very different overviews of the macroevolutionary history of a biological group. 
For example, alternative topologies in chronograms from the same biological group can infer very different species diversification patterns [@rabosky2015no; @title2016macrophylogenies].
<!-- There is evidence that different chronograms will result in very different macroevolutionary histories inferred from them [@title2016macrophylogenies]. However, it is unknown how using very different age data could affect hypothesis testing in other areas of research and more effort is needed in that direction -->
Similarly there are worries that patterns of morphological evolution cannot be accurately inferred with these types of phylogenies, as any patterns would be erased by randomization [@rabosky2015no]. We note that the same applies for geographical and morphological dependent diversification analysis. Hence, we suggest that these types of randomly resolved phylogenies can be used as null or neutral models, representing the case of diversification independent of traits and geographical scenario.

While stochastic politomy resolvers have been useful to advance research, notably, risks come with this practice.
Taken to the extreme, one could generate a fully resolved, calibrated tree of all modern and extinct taxa using a single taxonomy, a single calibration, and assigning branch lengths following a birth-death diversification model. Clearly, this can lead to a misrepresentation of the true evolutionary history.
We urge DateLife users to follow the example of the large tree papers cited above, by carefully considering the statistical assumptions being made, potential biases, and assessing the consistency of the results with prior work.


\begin{center}
\textsc{Conclusions}
\end{center}

Knowledge of the evolutionary time frame of organisms is key to many research areas: trait evolution, species diversification, biogeography, macroecology and more. It is also crucial for education, science communication and policy, but generating chronograms is difficult, especially for those who want to use phylogenies but who are not systematists, or do not have the time to acquire and develop the necessary knowledge and skills to construct them on their own. Importantly, years of primarily publicly funded research have resulted in vast amounts of chronograms that are already available on scientific publications, but hidden to the public and scientific community for reuse.

<!--Mention again uses shown here and outline potential uses not shown here-->
The DateLife project allows for easy and fast summary of public and state-of-the-art data on time of lineage divergence.
It provides a straightforward way to get an informed idea on the state of knowledge of the time frame of evolution of different regions of the tree of life, and allows identifying regions that require more research, or that have conflicting information.
It is available as an R package, and as a web-based R shiny application at www.datelife.org
<!-- It can also summarize public data, or use it to generate new chronograms based on trees provided by the user. -->
Both summary and newly generated trees are useful to evaluate evolutionary hypotheses in different areas of research. The DateLife project should improve awareness of the existing variation in expert time of divergence data, and foster exploration of the effect of alternative divergence time hypothesis on the results of analyses, nurturing a culture of more cautious interpretation of evolutionary results.


\begin{center}
\textsc{Availability}
\end{center}

The DateLife software is free and open source. It can be used online through its R shiny web application hosted at <http://www.datelife.org>, and locally through the `datelife` R package, downloadable from Zenodo (<https://doi.org/10.5281/zenodo.593938>) and the CRAN repository [@sanchez2022].
DateLife's web application is maintained using RStudio's shiny server and the shiny package open infrastructure, as well as Docker and OpenTree's infrastructure (dates.opentreeoflife.org/datelife).
`datelife`'s stable version `r availability` for installation from CRAN repository using the command `install.packages(pkgs = "datelife")` from within R.
Development versions are available from DateLife's GitHub repository (<https://github.com/phylotastic/datelife>) and can be installed using the command `devtools::install_github("phylotastic/datelife")`.


\begin{center}
\textsc{Supplementary Material}
\end{center}

Supplementary material, including code, biological examples, benchmark results, data files and online-only appendices, can be found in the Dryad data repository (<https://doi.org/10.5061/dryad.cnp5hqc6w>), as well as in the Zenodo stable repositories that host the reproducible manuscript (<https://doi.org/10.5281/zenodo.7435094>), the biological examples (<https://doi.org/10.5281/zenodo.7435101>), and the software benchmark (<https://doi.org/10.5281/zenodo.7435106>). The corresponding development versions hosted on GitHub can be found at <https://github.com/LunaSare/datelifeMS1#readme>, <https://github.com/LunaSare/datelife_examples#readme>, and <https://github.com/LunaSare/datelife_benchmark#readme>. 

\begin{center}
\textsc{Funding}
\end{center}

Funding was provided by the US National Science Foundation (NSF) grants ABI-1458603 to the DateLife project; DBI-0905606 to the National Evolutionary Synthesis Center (NESCent); ABI-1458572 to the Phylotastic project; and ABI-1759846 to the Open Tree of Life project.

\begin{center}
\textsc{Acknowledgements}
\end{center}

The DateLife project was born as a prototype tool aiming to provide the services describe in this paper, and was initially developed over a series of hackathons at the National Evolutionary Synthesis Center, NC, USA [@Stoltzfus2013].
We thank colleagues from the O'Meara Lab at the University
of Tennessee Knoxville for suggestions, discussions and software testing.
The late National Evolutionary Synthesis Center (NESCent), which sponsored hackathons
that led to initial work on this project. The team that assembled DateLife's first proof of concept: Tracy Heath, Jonathan Eastman, Peter Midford, Joseph Brown, Matt Pennell, Mike Alfaro, and Luke Harmon.
The Open Tree of Life project that provides the open, metadata rich repository of
trees used to construct DateLife's chronogram database.
The many scientists who publish their chronograms in an open, reusable form, and the scientists who curate them for deposition in the Open Tree of Life repository.
The NSF for funding nearly all the above, in addition to the ABI grant that funded this project itself.

\newpage

# References

\begingroup
\setlength{\parindent}{-0.5in}
\setlength{\leftskip}{0.5in}

<div id="refs" custom-style="Bibliography"></div>
\endgroup


```{r fig1, child= if (!embed_figure) c("table1.Rmd", "table2.Rmd", "fig-workflow.Rmd", "fig-benchmark.Rmd", "fig-small-example.Rmd", "fig-topologies.Rmd", "fig-fringillidae-full-example.Rmd", "fig-cross-validation.Rmd")}
```
