---
title             : "DateLife: leveraging databases and analytical tools to reveal the dated Tree of Life"
shorttitle        : "DateLife: revealing the dated Tree of Life"

author:
  - name          : "Luna L. Sánchez Reyes"
    affiliation   : "1,2"
    corresponding : yes    # Define only one corresponding author
    address       : ""
    email         : "sanchez.reyes.luna@gmail.com"
    role:         # Contributorship roles (e.g., CRediT, https://casrai.org/credit/)
      - Data curation
      - Investigation
      - Software
      - Visualization
      - Validation
      - Writing - Original Draft Preparation
      - Writing - Review & Editing
  - name          : "Emily Jane McTavish"
    affiliation   : "1"
    role:
      - Resources
      - Software
      - Writing - Review & Editing
  - name          : "Brian O'Meara"
    affiliation   : "2"
    role:
      - Conceptualization
      - Funding acquisition
      - Methodology
      - Resources
      - Software
      - Supervision
      - Writing - Review & Editing

affiliation:
  - id            : "1"
    institution   : "University of California, Merced"
  - id            : "2"
    institution   : "University of Tennessee, Knoxville"

authornote: |
  School of Natural Sciences, University of California, Merced, Science and Engineering Building 1.

  Department of Ecology and Evolutionary Biology, University of Tennessee, Knoxville, 446 Hesler Biology Building, Knoxville, TN 37996, USA.

abstract: |
  Date estimates for times of evolutionary divergences are key data for research in the natural sciences. These estimates also provide valuable information for for education, science communication, and policy decisions.
  Although achieving a high-quality reconstruction of a phylogenetic tree with branch lengths proportional to absolute time (chronogram) is a difficult and time-consuming task, the increased availability of fossil and molecular data, and time-efficient analytical techniques has resulted in many recent publications of large chronograms for a large number and wide diversity of organisms.
  When these estimates are shared in public, open databases this wealth of expertly-curated and peer-reviewed data on time of evolutionary origin is exposed in a programatic and reusable way.
  Intensive and localized efforts have improved data sharing practices, as well as incentivizited open science in biology.
  Here we present DateLife, a service implemented as an R package and an R Shiny website application available at www.datelife.org, that provides functionalities for efficient and easy finding, summary, reuse, and reanalysis of expert, peer-reviewed, public data on time of evolutionary origin.
  The main DateLife workflow constructs a chronogram for any given combination of taxon names by searching a local chronogram database constructed and curated from the Open Tree of Life Phylesystem phylogenetic database, which incorporates phylogenetic data from the TreeBASE database as well.
  We implement and test methods for summarizing time data from multiple source chronograms using supertree and congruification algorithms, and using age data extracted from source chronograms as secondary calibration points to add branch lengths proportional to absolute time to a tree topology.
  DateLife will be useful to increase awereness of the existing variation in expert time of divergence data, and can foster exploration of the effect of alternative divergence time hypotheses on the results of analyses, providing a framework for a more informed interpretation of evolutionary results.

  <!--  R has become mainstream in the biological-data analyst toolkit. Hence, to improve accessibility of time of evolutionary origin data we developed an R package that facilitates interaction, reanalysis and reuse of it, as well as incorporation of these data into the evolutionary analysis workflow. -->

bibliography      : ["paper_references.bib"]
keywords          : "Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations"
wordcount         : "`r wordcountaddin::word_count()`"
floatsintext      : no
figurelist        : no
tablelist         : no
footnotelist      : no
linenumbers       : yes
mask              : no
draft             : no

documentclass     : "apa6"
classoption       : "man"
output            :
  papaja::apa6_pdf:
    includes      :
      in_header: "preamble.tex"
      after_body  : ["figures-alt.tex", "figures-supplementary.tex"]

nocite: |
  @barker2013going, @barker2015new, @burns2014phylogenetics, @claramunt2015new, @gibb2015new, @Hedges2015, @hooper2017chromosomal, @Jetz2012, @price2014niche
---

```{r setup, include = FALSE}
library("papaja")
r_refs("paper_references.bib")
```

```{r analysis-preferences}
# Seed for random number generation
set.seed(50)
knitr::opts_chunk$set(cache.extra = knitr::rand_seed)
data(opentree_chronograms, package = "datelife")
rpackage_available <- TRUE
availability <- ifelse(rpackage_available, "is available", "will be available")
load("../data/dquery.rda")
load("../data/dres.rda")
```

# Introduction
<!-- If you're aiming for syst bio you don't want to start so general Readers know what chronograms and phylogenies are!.-->
Chronograms --phylogenies with branch lengths proportional to time-- provide key data for the study of natural processes in many areas of biological research, such as developmental biology [@laubichler2009form; @delsuc2018phylogenomic], conservation biology [@Felsenstein1985a; @Webb2000], historical biogeography [@posadas2006historical], and species diversification [@magallon2001absolute; @Morlon2014].

Building a chronogram is not an easy task.
It requires obtaining and curating data to construct a phylogeny, selecting and placing appropriate calibrations on the phylogeny using independent age data points from the fossil record or other dated events, and inferring the full dated tree;
it also generally requires specialized biological training, taxonomic domain knowledge, and a non-negligible amount of research time, computational resources and funding.

Here we present the DateLife project and its core software application, available as an R package [@datelifeR], and as an online Rshiny interactive website at www.datelife.org, which captures data from published chronograms, and make these data readily accessible to users for reuse and reanalysis.
The software features key elements for scientific reproducibility, such as a versioned, open and fully public chronogram database [@mctavish2015phylesystem], age data stored in a computer readable format [@vos2012nexml], automated and programmatic ways of accessing the data [@Stoltzfus2013] and methods to summarize and compare age data.


# Description

DateLife's core software application consists of the R package `datelife`. Its latest stable version -- v`r packageVersion("datelife")`, is available from the CRAN repository [@datelifeR], and relies on functionalities from various biological R packages:
ape [@paradis2004],
bold [@Chamberlain2018],
geiger [@Harmon2008],
paleotree [@Bapst2012a],
phyloch [@Heibl2008],
phylocomr [@Ooms2018],
phytools [@Revell2012],
rotl [@Michonneau2016], and
taxize [@Chamberlain2013; @Chamberlain2018].
Figure \ref{fig:workflow} provides a graphical summary of the three main steps of the DateLife algorithm: providing an input, searching a chronogram database, and summarizing results from the search.

<!--_**Providing an input.--**_ -->

## Processing an input

DateLife starts by processing an input consisting of at least two taxon names, which can be provided as a comma separated character string or as tip labels on a tree. If the input is a tree, it can be provided as a classic newick character string [@archie1986newick], or as a "phylo" R object [@paradis2004]. The input tree is not required to have branch lengths, and its topology is used in the summary steps described below.

DateLife accepts scientific names that can belong to any inclusive taxonomic group (e.g., genus, family, tribe, etc.) or a binomial species name. Subspecies and variants are ignored. If an input taxon name belongs to an inclusive taxonomic group the algorithm has two alternative behaviors defined by the "get species from taxon" flag. If the flag is active, DateLife retrieves all species names within the inclusive taxonomic group (according to a taxonomy) and adds them to the input string. If the flag is inactive, DateLife excludes the taxon names above the species level from the input.

DateLife processes input scientific names using a Taxonomic Name Resolution Service (TNRS), which increases the probability of correctly finding the queried taxon names in the chronogram database. TNRS detects, corrects and standardizes name misspellings and typos, variant spellings and authorities, and nomenclatural synonyms to a single taxonomic standard [@Boyle2013]. DateLife implements TNRS using OpenTree's unified taxonomy as standard [@opentreeAPIs; @rees2017automated], storing OpenTree's Taxonomy identification numbers for further processing.
<!-- We are using OpenTree's TNRS [BCO: yes], Boyle described TNRS first, right?  [BCO: earliest I can find, but feels like there should be much older work]-->

The processed input taxon names are saved as an R object of a newly defined class \texttt{datelifeQuery} that is used in the following steps. This object contains the standardized names, the corresponding OpenTree taxonomic id numbers, and the topology of the input tree if any was provided.

<!--_**Searching the database.--**_  -->

## Searching the database

DateLife's chronogram database latest version consist of `r length(opentree_chronograms$trees)` chronograms published in `r length(unique(opentree_chronograms$studies))` different studies. It is curated from OpenTree's phylogenetic database, the Phylesystem, which constitutes an open source of expert and peer-reviewed phylogenetic knowledge with rich metadata [@mctavish2015phylesystem], which allows automatic and reproducible assembly of our chronogram database. Datelife's chronogram database is navigable as an R data object within the `datelife` R package.
Published chronograms can be added to Phylesystem by any user, at any time, and are immediately publicly available (https://tree.opentreeoflife.org/curator). A user may then update their local cache of DateLife's chronogram database to include new chronogram data on a following search.

A DateLife search is implemented by matching processed taxon names provided by the user to tip labels in the chronogram database. Chronograms with at least two matching taxon names on their tip labels are identified and pruned down to preserve only the matched taxa.
These matching pruned chronograms are referred to as source chronograms.
Total distance (in units of millions of years) between taxon pairs within each source chronogram are stored as a patristic distance matrix (Figure \ref{fig:workflow}).
The matrix format speeds up extraction of pairwise taxon ages of any queried taxa, as opposed to searching the ancestor node of a pair of taxa in a "phylo" object or newick string.
Finally, the patristic matrices are associated to the study citation where the original chronogram was published, and stored as an R object of the newly defined class \texttt{datelifeResult}.

<!-- _**Summarizing search results.--**_ -->

## Summarizing search results

Summary information is extracted from the \texttt{datelifeResult} object to inform decisions for subsequent steps in the analysis workflow. Basic summary information available to the user is:

1. The matching pruned chronograms as newick strings or "phylo" objects.
2. The ages of the root of all source chronograms. These ages can correspond to the age of the most recent common ancestor (mrca) of the user's group of interest if the source chronograms have all taxa belonging to the group. If not, the root corresponds to the mrca of a subgroup withing the group of interest.
3. Study citations where original chronograms were published.
4. A report of input taxon names matches across source chronograms.
5. The source chronogram(s) with the most input taxon names.
6. Various single summary chronograms resulting from summarizing age data, generated using the methodology described below.


_**Choosing a topology.--**_ 
DateLife requires a tree topology to summarize age data upon.
We recommend that users provide a tree topology as input from the literature, or one of their own making. If no topology is provided, DateLife automatically subsets one from the OpenTree synthetic tree [@opentreeoflife2019synth]. 
Alternatively, DateLife can combine topologies from source chronograms using a supertree approach.
To combine topologies from source chronograms into a single summary (or supertree) topology, the DateLife algorithm starts by identifying the source chronograms that form a grove, roughly, a sufficiently overlapping set of taxa between trees, by implementing definition 2.8 for n-overlap from Ané et al. [-@ane2009groves]. In rare cases, a group of trees can have multiple groves. By default, DateLife chooses the grove with the most taxa, however, the "criterion = trees" flag allows the user to choose the grove with the most trees instead.
If source chronograms do not form a grove, the supertree reconstruction will fail.

_**Dating the topology.--**_ 
Input topologies from OpenTree or the supertree approach described above do not include branch length estimates of any kind.
Optionally, to estimate branch lengths proportional to substitution rates on these topologies, DateLife can mine the Barcode of Life Data System, BOLD [@ratnasingham2007bold] to obtain genetic markers for the input taxa.
These markers are aligned with MUSCLE [@edgar2004muscle] (by default) or MAFFT [@katoh2009multiple]. 
This alignment can be used to estimate branch lengths on input topologies that lack branch lengths.
Currently, branch length reconstruction in DateLife is performed using parsimony and the likelihood of the phylogenetic tree given a sequence alignment is computed [@schliep2011phangorn].
While relative branch length information provides additional data for nodes without secondary date calibrations, topologies without branch lengths can also be dated.

Once a topology is chosen, DateLife applies the congruification method [@Eastman2013] to find nodes belonging to the same clade across source chronograms, and extract the corresponding node ages from the patristic distance matrices stored as `datelifeResult`. By definition, the matrices store total distance (time from tip to tip), hence, node ages correspond to half the values stored in the patristic distance matrices.
This assumes that the terminal taxa are coeval and occur at the present.
A table of congruified node ages that can be used as secondary calibrations is stored as a `congruifiedCalibrations` object.

For each congruent node, the pairwise distances that traverse that node are summarized into a single summary matrix using classic summary statistics (i.e., mean, median, minimum and maximum ages), and the Supermatrix Distance Method [SDM; @Criscuolo2006], which deforms patristic distance matrices by minimizing variance and then averaging them.
These single summary taxon pair age matrices (Summarized calibrations) can be applied as secondary calibrations to date a tree topology, using different dating methods currently supported within DateLife: MrBayes [@Huelsenbeck2001; @Ronquist2003], PATHd8 [@Britton2007], BLADJ [@Webb2008; @webb2005phylomatic], and treePL [@Smith2012].

By default, DateLife implements the Branch Length Adjuster (BLADJ) algorithm to obtain a fully dated topology. BLADJ fixes node ages that have calibration data, and distributes time between nodes with no data evenly between nodes with calibration data.
This minimizes age variance in the resulting chronogram [@Webb2008].
BLADJ does not incorporate input phylogeny branch lengths even when they are present.
When there is conflict in ages between nodes with calibration data, the algorithm ignores node ages that are older than the age of a parent node. 
BLADJ requires a root age estimate. If there is no information on the age of the root in the chronogram database, users can provide an estimate from the literature. If none is provided, DateLife assigns an arbitrary age to the root as 10% older than the oldest age available within the group.

Alternative options supported in DateLife (MrBayes, PATHD8, TreePL) incorporate branch length information from the input topology in combination with the secondary calibrations.
PATHd8 is a non-clock, rate-smoothing method [@Britton2007] to date trees.
treePL [@Smith2012], is a semi-parametric, rate-smoothing, penalized likelihood dating method
[@Sanderson2002].
The MrBayes  [@Huelsenbeck2001; @Ronquist2003] approach in datelife uses the secondary calibrations as priors on node ages. 

_**Visualizing results.--**_
Finally, users can save all source and summary chronograms in formats that permit reuse and reanalyses (newick and R "phylo" format), as well as visualize and compare results graphically, or construct their own graphs using \texttt{datelife}'s chronogram plot generation functions available from the R package `datelifeplot` [@datelifeplot].


<!-- TODO ADD FIGURE: mock example explaining difference between using summary ages and raw ages to calibrate a topology. Then show it in the biological example. This will show what happens if a tree (phylogenetic conflict) or node (age) do not agree.-->


# Benchmark

`datelife`'s code speed was tested on an Apple iMac
with one 3.4 GHz Intel Core i5 processor. <!--No multiprocessor tests were performed-->
We registered variation in computing time of query processing and search through the database relative to number of queried taxon names.
Query processing time increases roughly linearly with number of input taxon names, and
increases considerably if Taxonomic Name Resolution Service (TNRS) is activated.
<!-- Defining tnrs here again, bc we only talk about it two times, so it easy to forget it -->
Up to ten thousand names can be processed and searched in less than 30 minutes with the most time consuming settings.
Once names have been processed as described in methods, a name search through the chronogram database can be performed in less than a minute, even with a very large number of taxon names (Fig. \ref{fig:runtime_main}).
<!--Speed of other datelife workflows? NO, bc it reflects speed from other software mainly. But I will put a sentence about it...
Summarizing `datelife` results processing times: shows sdm speed, clustering methods speed;
generating new trees: shows bladj speed;
Adding dates processing time: reflects more mrbayes speed than anything;
get_bold_otol_tree running time: reflects bold downloading data speed.-->
<!--Speed with different number of chronograms in database? NO-->
<!--Speed of web interface and of r package in computers with different capacities? NOOOO-->
`datelife`'s code performance was evaluated with a set of unit tests designed and
implemented with the R package testthat [@RCoreTeam2018] that were run both locally
with the devtools package [@RCoreTeam2018], and on a public server --via
GitHub, using the continuous integration tool Travis CI (<https://travis-ci.org>). At
present, unit tests cover more than 40% of `datelife`'s code (<https://codecov.io/gh/phylotastic/datelife>). <!-- unit test make code robust, verifiable, debuggable, -->
Unit testing helps identify potential issues as code is updated or, more critically, as services code relies upon may change.


# Case studies

We illustrate the DateLife algorithm using a group within the Passeriform birds encompassing the family of true finches, Fringillidae, as case study. The first example analyses 6 bird species and shows all steps of the algorithm. The second example is an application analysing all 289 species in the family Fringillidae for which we have date informtaion.
<!--We used `r cite_r("r-references.bib")` for all our analyses.-->

## Small example

We randomly chose 6 bird species in the family Fringillidae, the true finches. The sample includes
two species of cardinals:
the black-thighed grosbeak -- *Pheucticus tibialis* and
the crimson-collared grosbeak -- *Rhodothraupis celaeno*;
three species of buntings:
the yellowhammer -- *Emberiza citrinella*,
the pine bunting --  *Emberiza leucocephalos* and
the yellow-throated bunting --  *Emberiza elegans*;
and one species of tanager, the vegetarian finch -- *Platyspiza crassirostris*.

Processing input names found that *Emberiza elegans* is synonym for *Schoeniclus elegans* in the default reference taxonomy [Open Tree of Life Taxonomy v3.3, June 1, 2021]. For a detailed discussion on the state of the synonym refer to Avibase [@lepage2004avibase; @lepage2014avibase; @avibase-emberiza]. 
DateLife used the processed input names to search the local chronogram database and found 9 matching chronograms in 6 different studies. Three studies matched five input names [@barker2015new; @Hedges2015; @Jetz2012], one study matched four input names [@hooper2017chromosomal] and two studies matched two input names [@barker2013going; @burns2014phylogenetics]. No studies matched all input names. Together, matching chronograms have 28 unique age data points. All nodes have age data.
DateLife used OpenTree's synthetic tree topology for these taxa and mapped age data to nodes in the tree.
As expected, more inclusive nodes (e.g., node "n1") have more age data than less inclusive nodes (e.g., node "n5").
The name resolution step allowed discovering five data points for node "n4" that would not have had any data otherwise due to name mismatch. 
Age summary statistics per node were calculated and used as secondary calibrations to date the tree topology using the BLADJ algorithm.
Age data for node "n2" was excluded as final calibration because it is older than age data of a more inclusive node (n1).

## Fringillidae 

To estimate ages for the taxa in the family Fringillidae, we ran a `datelife` query using the "get species from taxon" flag. 
This will get all recognised species names within the named taxon. 
The Fringillidae has species, according to the Open Tree of Life taxonomy. <!--BCO: Feel like we need a number here, though it's good that the clade hasn't gone extinct ;-) -->
??? We have date estimates for `r length(dquery$cleaned_names)` of those species.

??? There is an option to include all, right?
According to the OpenTree of life synthetic phylogeny, Fringillidae is not monophyletic, so this taxon-constrained approach returns a topology and ages for the species in that named group. <!--BCO: it's ambiguous: it gets a topology and age for the larger clade that includes all Fringillidae species? Or for the paraphyletic Fringillidae group?-->

The next step is to run a `datelife` search on this curated set of species taxon names that will find all chronograms that contain at least two species. The algorithm proceeds to prune the trees to keep matching species names on tips only, and transform the pruned trees to pairwise distance matrices.
There are `r length(dres)` chronograms containing at least two Fringillidae species, published in `r length(unique(names(dres)))` different studies [@barker2013going; @barker2015new; @burns2014phylogenetics; @claramunt2015new; @gibb2015new; @Hedges2015; @hooper2017chromosomal; @Jetz2012; @price2014niche].
The final step is to summarize the available information using two alternative types of summary chronograms, median and SDM. As explained in the "Description" section, data from source chronograms is first summarised into a single patristic distance matrix and then the available node ages are used as fixed node calibrations over a consensus tree topology, to obtain a fully dated tree with the program BLADJ (Fig. \ref{fig:fringillidages}). Median summary chronograms are older and have wider variation in maximum ages than chronograms obtained with SDM.
<!-- ????????????????? Say some things about the results!
ANSWER:
(Going from a summary distance matrix Using this method is better than using clustering algorithms  -->
<!-- and hierarchical clustering method which appears to push ages back -->

# Cross-validation test

<!--Data from source chronograms can be used to date tree topologies with no branch lengths, as well as trees with branch lengths as relative substitution rates-->
We performed a cross validation analysis of the DateLife workflow using the Fringillidae chronograms. 
We used the individual tree topologies from each of the ??? TODO CHECK NUMBER 19 source chronograms from ?? studies as inputs, treating their node ages as unknown.
We estimated dates for these topologies using the node ages from the chronograms from the other studies as calibrations and smoothing using BLADJ.

We found that node ages from original study, and ages estimated using all other age data available are correlated  \textcolor{red}{R^2 value?} (Supplementary Fig. \ref{fig:cvXY}).

For five studies, Datelife tended to underestimate ages for topologically deeper nodes (those with many descendant taxa, aka 'closer to the root') relative to the orginal estimate, and overestimate ages for nodes closer to the tips. 
Accordingly, root ages are generally older in the original study than estimated using cross-validated ages (Supplementary Fig. \ref{fig:cvXYdiffs}).


<!-- 
%```{r, child = "cross-validation-pathd8.Rmd"}
%```
-->

# Discussion

`datelife` makes state-of-the-art information on time of lineage divergence easily accessible for comparison, reuse, and reanalysis, to researchers in all areas of science and with all levels of expertise in the matter. It is an open service that does not require any expert biological knowledge from users, besides the names of the organisms or clade they want to work with, for any of its functionality.

At the time of writing of this manuscript (`r format(Sys.time(), "%b %d, %Y")`), `datelife`'s database has `r length(opentree_chronograms$trees)` chronograms, pulled entirely from OpenTree's database, the Phylesystem [@mctavish2015phylesystem]. A unique feature of OpenTree's Phylesystem is that the community can add new state-of-the-art chronograms any time. As chronograms are added to Phylesystem, they are incorporated into an updated `datelife`'s database that is assigned a new version number, followed by a package release on CRAN. `datelife`'s chronogram database is updated as new chronogram data is added to Phylesystem, at a minimum of once a month and a maximum of every 6 months. <!--BCO: Nicely done-->
Users can also upload new chronograms to OpenTree themselves, and trigger an update of their local `datelife` database to incorporate the new chronograms, to have them immediately available for analysis. <!--BCO: What is immediately? Also, it's not just uploading -- they have to include metadata on what brlen mean, maybe? indicate the ingroup/outgroup, do TNRS within OToL, right? Maybe this has improved. I remember being frustrated trying to load in giant trees for datelife use b/c I had to click on the ingroup node-->
<!--BCO; this duplicates an earlier section and feels more methods-y. Maybe move up and merge?-->


A total of `r length(unique(unname(unlist(lapply(opentree_chronograms$trees, "[[", "tip.label")))))` unique terminal taxa are represented in `datelife`'s database.
Incorporation of more chronograms into `datelife`'s database will continue to improve its services. One option to increase the of chronograms in the DateLife database is the Dryad data repository, which now holds some chronograms.  Methods to automatically mine chronograms from Dryad could be designed and implemented. However, Dryad's metadata system has no information to automatically detect branch length units, and those would still need to be determined manually by a curator.

The largest, and taxonomically broadest, summary chronogram currently available from OpenTree
was constructed using age data from 2,274 published chronograms [@Hedges2015].
<!--BCO: I just checked: TIMETREEE NOW ALLOWS SOURCE TREE DOWNLOADS! This is great news. They have a copyright (not legal in US on trees, IMHO) notice on individual trees, but they do say "To download all individual study timetrees, please send an email request to info@timetree.org." -- I think this could be great to do--> 

We would like to emphasize on the importance of sharing chronogram data,  including systematically curated metadata, into open repositories, such as OpenTree's Phylesystem [@mctavish2015phylesystem] for the benefit of the scientific community as a whole.

By default, `datelife` currently summarizes all source chronograms that overlap with at least two species names. Users can exclude source chronograms if they have reasons to do so.
Strictly speaking, the best chronogram should reflect the real time of lineage divergence accurately and precisely.
To our knowledge, there are no good measures to determine independently if a chronogram is better than another.
Some measures that have been proposed are the proportion of lineage sampling and the number of calibrations used [@magallon2010using, @magallon2015metacalibrated].
Several characteristics of the data used for dating analyses, as well as from the output chronogram itself, could be used to score quality of source chronograms.
Some characteristics that are often cited in published studies as a measure of improved age estimates as compared to previously published estimates are: quality of alignment (missing data, GC content), lineage sampling (strategy and proportion), phylogenetic and dating inference method, number of fossils used as calibrations, support for nodes and ages, and magnitude of confidence intervals.
DateLife provides an opportunity to capture concordance and conflict among date estimates, which can also be used as a metric for chronogram reliability.
Its open database of trees allows other researchers to do such analyses themselves reproducibly and without needing permission, though of course they should follow proper citation practices, especially for the source trees.

Scientists have more confidence in chronograms constructed using primary calibrations (ages obtained from the fossil or geological record) to ones constructed with secondary calibrations (ages coming from other chronograms)[@schenk2016sec]. @schenk2016sec showed that in simulations divergence times inferred with secondary calibrations are significantly younger than those inferred with primary calibrations in analyses performed with Bayesian inference methods when priors are implemented in similar ways in both analyses. However, secondary calibrations are useful in dating methods that do not require setting priors, such as penalized likelihood [@sanderson2003r8s], or as fixed ages for nodes.
Certainly, further studies are required to fully understand the effect of using secondary calibrations on time estimates and downstream analyses.

The input chronograms may have been estimated using primary fossil data or using estimates of molecular substitution rates.
These differences in approach which can deepen the already substantial variation in time estimates between lineages, as observed from the comparison of source chronograms in the Fringillidae example.
For eaxmple, the chronograms from  @burns2014phylogenetics tend to have much older date estimates for the same nodes than other inferences (Figure 5). 
This study did not use any primary calibrations, and inferred dates using an substitution rate estimates across birds \textcolor{red}{(cite Weir, J.T., Schluter, D., 2008. Calibrating the avian molecular clock. Mol. Ecol. 17,
231–2328).}
\textcolor{red}{MORE ON the FINCH examples and the cross validation here}

Conflict among alternative date estimates is common in the literature
(see, for example, the robust ongoing debate about crown group age of angiosperms [@sauquet2021age; @barba2018constraining; @magallon2015metacalibrated; @sanderson2001sources; @ramshaw1972time]. 
Using a different chronogram may return different results [@title2016macrophylogenies] particularly for studies of species diversification, or focussed on the timing of evolutionary events.
Stitching together multiple chronograms creates a larger tree that uses information from across studies, but may amplify uncertainty and errors.

Summarizing chronograms might also imply summarizing fundamentally distinct evolutionary hypotheses.
For example, two different researchers working on the same clade could both carefully select and argue for their choices of fossil calibrations.
Still, if one researcher decides a fossil will calibrate the ingroup of a clade, while another researcher uses the same one to calibrate outside the clade, the resulting age estimates will often differ substantially, as the placement of calibrations as stem or crown group is proved to deeply affect estimated times of lineage divergence [@sauquet2013practical]. Trying to summarize the resulting chronograms into a single one using simple summary statistics can erase many types of relevant information from the source chronograms. Accordingly, the prevailing view is that we should favor lineage divergence estimates obtained from a single analysis, using fossil data as primary sources of calibrations, and using fossils that have been widely discussed and curated as calibrations to date other trees, making sure that all data used in the analysis reflect a coherent evolutionary history, as for example done by [@antonelli2017supersmart]. <!-- Will this be implemented in future `datelife` versions? Probably not. -->
There have been attempts to create fossil calibration databases <!--REF TO ADD: Ksepka, D.T., J.F. Parham, J.F. Allman, M.J. Benton, M.T. Carrano, K.A. Cranston, P.C. J. Donoghue, J.J. Head, E.J. Hermsen, R.B. Irmis, W.G. Joyce, M. Kohli, K.S. Lamm, D. Leehr, J.S.L. Patané, P.D. Polly, M.J. Phillips, N.A. Smith, N.D. Smith, M. van Tuinen, J.L. Ware, R.C.M. Warnock. 2015.The Fossil Calibration Database, A New Resource for Divergence Dating. Systematic Biology: doi:10.1093/sysbio/syv025.-->, though these still have room to grow. <!--BCO: Polite way to say 'no new additions in four years?'-->
However, the exercise of summarizing different chronograms has the potential to help getting a single global evolutionary history for a lineage by putting together evidence from different hypotheses. Choosing the elements of the chronograms that we are going to keep and the ones that we are going to discard is key, since we are potentially losing important parts of the evolutionary history of a lineage that might only be reflected in source chronograms and not on the summary chronogram [@sauquet2021age]. <!-- Example -->
\textcolor{red}{??? it's not clear to me what this last part means! What might you be losing?}

<!-- still requires more research. -->
<!-- Often, there is a large discrepancy in time of lineage divergence estimates across studies from expert knowledge, that is well known and documented for different groups of organisms [e.g., angiosperms; @magallon2015metacalibrated].  -->
<!-- Comparison of available age data for a wide range of organisms shown here suggest that this is a widespread phenomenon that requires further attention. -->
<!-- In the case of the example presented here, most chronograms are new, meaning that we expect that data was generated with high quality and the software used should be very advanced. However, this is just an assumption.  -->

<!-- Still, even if all source chronograms have been generated by excellent standards and using similar methods, the evolutionary history they depict might be very different.  -->

Nonetheless, in ecology and conservation biology, incorporating at least some data on lineage divergence times represents a relevant improvement for testing alternative hypothesis using phylogenetic distance [@Webb2008].
Hence, we integrated into datelife's workflow different ways of estimating node ages in the absence of calibrations and branch length information for taxa lacking this information.
"Making up" branch lengths is an common practice in scientific publications: @Jetz2012, created a time-calibrated tree of all 9,993 bird species, where 67% had molecular data and the rest was simulated; @rabosky2018inverse created a time-calibrated tree of 31,536 ray-finned fishes, of which only 37% had molecular data; @smith2018constructing constructed a tree of 353,185 seed plants where only 23% had molecular data.
Obviously, there are risks in this practice!
Taken to the extreme, one could make a fully resolved, calibrated tree of all modern and extinct taxa using a single taxonomy and a single calibration with the polytomy resolution and branch estimation methods.
There has yet to be a thorough analysis of what can go wrong when one extends inferences beyond the data in this way, so we urge caution; we also urge readers to follow the example of many of the large tree papers cited above and make carefully consider the statistical assumptions being made, and assess the consistency of the results with prior work.


<!-- There is evidence that different chronograms will result in very different macroevolutionary histories inferred from them [@title2016macrophylogenies]. However, it is unknown how using very different age data could affect hypothesis testing in other areas of research and more effort is needed in that directions -->

# Conclusions

Divergence time information is key to many areas of evolutionary studies: trait evolution,
diversification, biogeography, macroecology and more. It is also crucial for science communication and education, but generating chronograms is difficult,
especially for those who want to use phylogenies but who are not systematists, or
do not have the time to acquire and develop the necessary knowledge and data curation skills. Moreover, years of primarily public funded research have resulted in vast amounts of chronograms that are already available on scientific publications, but hidden to the public and scientific community for reuse.

<!--Mention again uses shown here and outline potential uses not shown here-->
The `datelife` R package allows easy and fast summarization of publicly available information
on time of lineage divergence. This provides a straightforward way to get an informed idea on the state of knowledge of the time frame of evolution of different regions of the tree of life, and allows identification of regions that require more research or that have conflicting information.
It is available as an R package, or a web-based R shiny app at dates.opentreeloflife.org/datelife.
<!-- It can also summarize public data, or use it to generate new chronograms based on trees provided by the user. -->
Both summary and newly generated trees are useful to evaluate evolutionary hypotheses in different areas of research. The DateLife project helps with awareness of the existing variation in expert time of divergence data, and will foster exploration of the effect of alternative divergence time hypothesis on the results of analyses, nurturing a culture of more cautious interpretation of evolutionary results.


# Availability

`datelife` is free and open source and it can be used through its current website
<http://www.datelife.org>, through the `datelife` R package, and through Phylotastic's project web portal <https://phylo.cs.nmsu.edu/>.
`datelife`'s website is maintained using RStudio's shiny server and the shiny package open infrastructure, as well as Docker.
`datelife`'s R package stable version `r availability`
for installation from the CRAN repository (<https://cran.r-project.org/package=datelife>)
using the command `install.packages(pkgs = "datelife")` from within R. Development versions
are available from the GitHub repository (<https://github.com/phylotastic/datelife>)
and can be installed using the command `devtools::install_github("phylotastic/datelife")`.


# Supplementary Material

Code used to generate all versions of this manuscript, the biological examples, as well as the benchmark of functionalities are available at [datelifeMS1](https://github.com/LunaSare/datelifeMS1), [datelife_examples](https://github.com/LunaSare/datelife_examples), and [datelife_benchmark](https://github.com/LunaSare/datelife_benchmark) repositories in LLSR's GitHub account.

# Funding

Funding was provided by the US National Science Foundation (NSF) grants ABI-1458603 to the Datelife project; DBI-0905606 to the National Evolutionary Synthesis Center (NESCent), ABI-1458572 to the Phylotastic project, and ABI-1759846 to the Open Tree of Life project.

# Acknowledgements


The DateLife project was born as a prototype tool aiming to provide these services, and was initially developed over a series of hackathons at the National Evolutionary Synthesis Center, NC, USA [@Stoltzfus2013].
We thank colleagues from the O'Meara Lab at the University
of Tennesse Knoxville for suggestions, discussions and software testing.
The late National Evolutionary Synthesis Center (NESCent), which sponsored hackathons
that led to initial work on this project. The team that assembled `datelife`'s first proof of concept: Tracy Heath, Jonathan Eastman, Peter Midford, Joseph Brown, Matt Pennell, Mike Alfaro, and Luke Harmon.
The Open Tree of Life project that provides the open, metadata rich repository of
trees used for `datelife`.
The many scientists who publish their chronograms in an open, reusable form, and
the scientists who curate them for deposition in the Open Tree of Life repository.
The NSF for funding nearly all the above, in addition to the ABI grant that funded this project itself.

\newpage

# References

\begingroup
\setlength{\parindent}{-0.5in}
\setlength{\leftskip}{0.5in}

<div id="refs" custom-style="Bibliography"></div>
\endgroup
