---
title: "Untitled"
output: pdf_document
date: "2023-06-06"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Getting source chronograms

```{r}
load("../data/frin_sample_dsumm.rda")
frin_sample_dsumm$phylo_median
spp <- frin_sample_dsumm$phylo_median$tip.label[1:6]
spp_write <- sub("_", " ", spp)
spp_write <- spp_write[order(spp_write)]
write(spp_write, file = "../figures/fringillidae_species_sample.txt")
spp_dq <- datelife::make_datelife_query(spp)
spp_dr <- datelife::get_datelife_result(spp_dq)
```

```{r}
topology <- datelife::get_otol_synthetic_tree(ott_ids = spp_dq$ott_ids)
spp_dq$phy <- topology
spp_summ <- datelife:::summary.datelifeResult(datelife_query = spp_dq, 
                                              object = spp_dr)
```



```{r}
phylo_median <- spp_summ$phylo_median
ape::branching.times(phylo_median)
```
Node ages in `phylo_median` do not seem to represent the median



```{r}
phylo_all <- spp_summ$phylo_all
# Reorder chronogram list alphabetically:
alphabetic_order <- order(names(phylo_all))
phylo_all <- phylo_all[alphabetic_order]
names(phylo_all)
```

```{r}
names(phylo_all) <- c("Barker et al. 2013",
                      "Barker et al. 2015 - chronogram 1",
                      "Barker et al. 2015 - chronogram 2",
                      "Burns et al. 2014",
                      "Hedges et al. 2015 - chronogram 1",
                      "Hedges et al. 2015 - chronogram 2",
                      "Hooper et al. 2017",
                      "Jetz et al. 2012 - chronogram 1",
                      "Jetz et al. 2012 - chronogram 2")
```

All calibrations, even those that are not congruent:

```{r}
all_calibrations <- datelife::extract_calibrations_phylo(
  input = phylo_all,
  each = FALSE)
class(all_calibrations)
ls(all_calibrations)
nrow(all_calibrations)
```
There is a total of 28 calibrations from the 9 source chronograms.

Congruified calibrations:
```{r}
matched <- datelife:::match_all_calibrations(
  phy = phylo_median, 
  calibrations = all_calibrations)
nrow(matched$matched_calibrations)
```
All calibrations are congruent with the topology.

```{r}
calib_summ <- datelife:::summary.matchedCalibrations(
  matched$matched_calibrations)
```

`calib_summ` and `matched` objects are the same:
```{r}
nrow(matched$matched_calibrations) == nrow(calib_summ$in_phy)
colnames(matched$matched_calibrations) == colnames(calib_summ$in_phy)
```

```{r}
head(matched$matched_calibrations) 
```
```{r}
head(calib_summ$in_phy)
```

Add nodeAge column to calib_summ:

```{r}
if (all(calib_summ$in_phy$MinAge == calib_summ$in_phy$MaxAge)) {
  calib_summ$in_phy$nodeAge <- calib_summ$in_phy$MinAge
}
rr <- calib_summ$in_phy$mrca_node_number == 7
calib_summ$in_phy[rr,]
mean(calib_summ$in_phy[rr,"nodeAge"])
```


Load a color scheme for the studies:

```{r}
load("../data/color_studies.rda")
color_studies
```

The following is supposed to get the median summary matrix, but it is not.
```{r}
best_grove <- datelife::get_best_grove(spp_dr)

# Take a list of chronograms as tables and summarize age data per taxon pair into a single matrix:
median_matrix <- datelife::datelife_result_median_matrix(best_grove$best_grove)
colnames(median_matrix)

median_matrix_list <- list(summary_matrix = median_matrix)

median_matrix_table <- datelife::matrices_to_table(median_matrix_list)

# match taxon pairs to nodes on a tree topology
matched_median_ages <- datelife:::match_all_calibrations(phy = spp_summ$phylo_median, calibrations = median_matrix_table)
mm_summ <- datelife:::summary.matchedCalibrations(matched_median_ages$matched_calibrations)
mm_summ$in_phy
```


Simplifying median ages on tree plot:

```{r}
# Getting the median per node:
median_singletons <- c()
for (i in unique(mm_summ$in_phy$mrca_node_name)) {
  print(i)
  x <- median(unique(mm_summ$in_phy$nodeAge[mm_summ$in_phy$mrca_node_name == i]))
  median_singletons <- c(median_singletons, x)
}

# Detecting node names that are duplicated 
duplos <- duplicated(mm_summ$in_phy$mrca_node_name)
# Simplifying median ages:
mm_summ_sim <- mm_summ
rr <- mm_summ$in_phy$mrca_node_number == 7
mm_summ$in_phy[rr,]
mm_summ_sim$in_phy$nodeAge[!duplos] <- median_singletons
# Removing duplicates:
mm_summ_sim$in_phy <- mm_summ$in_phy[!duplos,]
# Columns MinAge and MaxAge need to be equal to nodeAge
mm_summ_sim$in_phy$MinAge <- mm_summ_sim$in_phy$MaxAge <- mm_summ_sim$in_phy$nodeAge
```

Let's do a DateLife approach for getting the median summary:

```{r}
calib_summ$in_phy
node_median <- c()
for (i in (unique(calib_summ$in_phy$mrca_node_name))) {
  rr <- calib_summ$in_phy$mrca_node_name == i
  node_median <- c(node_median, median(calib_summ$in_phy$nodeAge[rr]))
}
node_median

mm_summ_sim$in_phy$nodeMedian <- node_median
mm_summ_sim$in_phy[,c("mrca_node_name", "nodeAge", "nodeMedian")]
mm_summ_sim$in_phy$nodeAge <- node_median

```
```{r}
topology$node.label <- mm_summ_sim$in_phy$mrca_node_name
plot(topology)
ape::nodelabels(text = topology$node.label)
phylo_median_new <- datelife::make_bladj_tree(tree = topology, 
                          nodenames = as.character(topology$node.label), 
                          nodeages = mm_summ_sim$in_phy$nodeMedian)
plot(phylo_median_new)
ape::branching.times(phylo_median_new)
```



Plotting parameters:

```{r}
mm_summ_sim$in_phy$reference <- rep("Used as calibration", length(mm_summ_sim$in_phy$reference))
mm_summ_sim$in_phy$reference[2] <- "Not used"
gplots::col2hex("purple") # "#A020F070", 
color_median <- rep("purple", 2) # mustard yellow "#7A0403FF"

names(color_median) <- unique(mm_summ_sim$in_phy$reference)

point_colors <- list(a1 = color_studies, a2 = color_median)
all_ages <- list(a1 = calib_summ, a2 = mm_summ_sim)

points_median <- stats::setNames(c("/", "*"), names(color_median))
point_type <- list(a1 =  19, a2 = points_median)
median_cex <- c(4.5, 6)
names(median_cex) <- names(color_median)
# point_cex <- list(a1 =  3, a2 = 4.5)
point_cex <- list(a1 =  3, a2 = median_cex)


leg_color<- list(a1 = color_studies, a2 = color_median)

leg_pch <- list(a1 = 19, a2 = c("/", "*"))
leg_cex <- c(1.1, 1.2)
leg_text <- lapply(leg_color, names)
leg_title <- c("Taxon pair node ages from source chronograms", "Median of node ages")
leg_box <- c(TRUE, TRUE)
leg_pt_cex <- list(a1 = 1.7, a2 = c(1.7, 2.5))
```



Plot:

```{r}
# source("../R/plot_node_ages_v2.1.R")
devtools::load_all(path = "../../datelifeplot")
source("../R/plot_node_ages_v2.2.R")
library(ape)
```

```{r}
size <- 2
svg(file = "../figures/figure-small-example/all_ages.svg",
    height = 4.5 * size,
    width = 6 * size)
plot_node_ages_2.2(chronogram = phylo_median_new,
               title = "",
               plot_type = "phyloch",
               time_depth = 45,
               pos_axis = 2,
               cex_axis = 2,
               cex_axislabels = 2,
               axis_label = "",
               edge_width = 3,
               cex_tiplabels = 2.3,
               matched_ages  = all_ages,
               pch_type = point_type,
               pch_color = point_colors,
               pch_cex = point_cex,
               bars_lwd = 23,
               legend_cex = leg_cex,
               legend_pt_cex = leg_pt_cex,
               legend_text = leg_text,
               legend_pch = leg_pch,
               legend_color = leg_color,
               legend_title = leg_title,
               legend_box = leg_box,
               legend_x = c(-6, -2),
               legend_y = c(6.2, 3.8),
               mai1 = 1, mai2 = 1, mai3 = 0.3, mai4 = 4,
               omi1 = 1, omi2 = 0.3, omi3 = 0.3, omi4 = 0.3,
               label.offset = 1
               )
dev.off()
```     
