---
title: "Congruification"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Congruification}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
devtools::load_all("../../datelifeplot")
devtools::load_all("../../datelife")
load("../data/fringillidae_v0.6.2.rda")
```

Function congruify calibrations:

## Extract all calibrations from chronograms as a data.frame if none are provided

This congruifies chronograms agianst themselves

```{r}
chronograms <- fringillidae_v0.6.2$dsumm$phylo_all
extracted_calibrations <- extract_calibrations_phylo(input = chronograms,
                                                     each = TRUE)
ls(extracted_calibrations)
colnames(extracted_calibrations[[1]])
```

## Match target tree (`phy`) nodelabels to calibrations data.frame

So we have all taxon pairs and the corresponding node labels from `phy`

```{r}
phy <- fringillidae_v0.6.2$dsumm$phylo_median
if (is.null(phy$edge.length)) {
    message("... Adding temporary branch lengths to 'phy' for congruification.")
    phy <- ape::compute.brlen(phy)
}
class(phy) <- "phylo"
phy
```
Self congruification for target tree (`phy`):

```{r}
phy_calibrations <- suppressWarnings(
                        geiger::congruify.phylo(reference = phy,
                                                target = phy,
                                                scale = NA,
                                                ncores = 1)$calibrations)
colnames(phy_calibrations)
```

```{r}
head(phy_calibrations)
```

```{r}
phy_matched <- mrca_calibrations(phy = phy,
                                 calibrations = phy_calibrations)
class(phy_matched)
length(phy_matched)
ls(phy_matched)
colnames(phy_matched$matched_calibrations)
nrow(phy_matched$matched_calibrations)
head(phy_matched$matched_calibrations)
phy_matched$matched_phy
```

## Congruify source chronograms to `phy_matched$matched_phy` and themselves
Function to
Congruify a source chronogram with the target and itself
return the calibrations data.frame only
  
```{r}
run_congruification_self <- function(phy,
                                      chronograms,
                                       index) {
    self <- chronograms[[index]]
    phy$tip.label <- sub(" ", "_", phy$tip.label)
    self$tip.label <- sub(" ", "_", self$tip.label)
    class(self) <- "phylo"
    cc_phy <- suppressWarnings(geiger::congruify.phylo(reference = phy,
                               target = self,
                               scale = NA,
                               ncores = 1))
    cc_self <- suppressWarnings(geiger::congruify.phylo(reference = self,
                               target = self,
                               scale = NA,
                               ncores = 1))
    list(phy = cc_phy, self = cc_self)
}
```

## Congruify them all:
```{r}
congruified_self <- lapply(seq(chronograms),
                            function(i) {
                              run_congruification_self(phy = phy_matched$matched_phy,
                                                  chronograms,
                                                  index = i)
                            })
  names(congruified_self) <- names(chronograms)
xx <- data.frame(congruent_calibrations = sapply(congruified_self, function(x) nrow(x$phy$calibrations)), 
           source_Nnode = sapply(congruified_self, function(x) nrow(x$self$calibrations)))

study_names <- c("Barker et al. 2013",  # 1
                      "Barker et al. 2015 - chronogram 1",  # 2
                      "Barker et al. 2015 - chronogram 2",  # 3
                      "Burns et al. 2014",  # 4
                      "Claramunt et al. 2015",  # 5
                      "Gibb et al. 2015", # 6
                      "Hedges et al. 2015 - chronogram 1",  # 7
                      "Hedges et al. 2015 - chronogram 2",  # 8
                      "Hooper et al. 2017",  # 9
                      "Jetz et al. 2012 - chronogram 1",  # 10
                      "Jetz et al. 2012 - chronogram 2",  # 11
                      "Kimball et al. 2019 - chronogram 1", # 12
                      "Kimball et al. 2019 - chronogram 2", # 13
                      "Oliveros et al. 2019", # 14
                      "Price et al. 2014 - chronogram 1", # 15
                      "Price et al. 2014 - chronogram 2", # 16
                      "Roquet et al. 2014 - chronogram 1", # 17
                      "Roquet et al. 2014 - chronogram 2", # 18 
                      "Uyeda et al 2017") # 19
row.names(xx) <- study_names
xx
```
Check that the order of tagret and self does not matter:

```{r}
# same as `run_congruification_self` but using chronograms as reference
run_congruification_self2 <- function(phy,
                                      chronograms,
                                       index) {
    self <- chronograms[[index]]
    phy$tip.label <- sub(" ", "_", phy$tip.label)
    self$tip.label <- sub(" ", "_", self$tip.label)
    class(self) <- "phylo"
    cc_phy <- suppressWarnings(geiger::congruify.phylo(reference = self,
                               target = phy,
                               scale = NA,
                               ncores = 1))
    cc_self <- suppressWarnings(geiger::congruify.phylo(reference = self,
                               target = self,
                               scale = NA,
                               ncores = 1))
    list(phy = cc_phy, self = cc_self)
}
congruified_self2 <- lapply(seq(chronograms),
                            function(i) {
                              run_congruification_self2(phy = phy_matched$matched_phy,
                                                  chronograms,
                                                  index = i)
                            })
  names(congruified_self) <- names(chronograms)
yy <- data.frame(congruent_calibrations = sapply(congruified_self, function(x) nrow(x$phy$calibrations)), 
           source_Nnode = sapply(congruified_self, function(x) nrow(x$self$calibrations)))

study_names <- c("Barker et al. 2013",  # 1
                      "Barker et al. 2015 - chronogram 1",  # 2
                      "Barker et al. 2015 - chronogram 2",  # 3
                      "Burns et al. 2014",  # 4
                      "Claramunt et al. 2015",  # 5
                      "Gibb et al. 2015", # 6
                      "Hedges et al. 2015 - chronogram 1",  # 7
                      "Hedges et al. 2015 - chronogram 2",  # 8
                      "Hooper et al. 2017",  # 9
                      "Jetz et al. 2012 - chronogram 1",  # 10
                      "Jetz et al. 2012 - chronogram 2",  # 11
                      "Kimball et al. 2019 - chronogram 1", # 12
                      "Kimball et al. 2019 - chronogram 2", # 13
                      "Oliveros et al. 2019", # 14
                      "Price et al. 2014 - chronogram 1", # 15
                      "Price et al. 2014 - chronogram 2", # 16
                      "Roquet et al. 2014 - chronogram 1", # 17
                      "Roquet et al. 2014 - chronogram 2", # 18 
                      "Uyeda et al 2017") # 19
row.names(yy) <- study_names
yy
```



## Cross congruify source chronograms and target tree??

I think we do not need a cross congruification step. 

We need the following info as a data.frame:
- mrca_node_number on target tree (NA if it is not congruent)
- taxon pair names
- age node from source chronogram
- citation from source chronogram

Add this to `run_congruification_self` function:

```{r}
index <- 1
run_congruification_self <- function(phy,
                                      chronograms,
                                       index) {
    self <- chronograms[[index]]
    phy$tip.label <- sub(" ", "_", phy$tip.label)
    self$tip.label <- sub(" ", "_", self$tip.label)
    class(self) <- "phylo"
    cc_phy <- suppressWarnings(geiger::congruify.phylo(reference = self,
                               target = phy,
                               scale = NA,
                               ncores = 1))
    
    mrcas <- mrca_calibrations(phy = cc_phy$target, 
                      calibrations = cc_phy$calibrations)
    calibs_matched <- mrcas$matched_calibrations
    calibs_matched$study <- rep(names(chronograms)[index], nrow(calibs_matched))
    return(calibs_matched)
}

```
Renamed to `congruify_and_mrca`.
It is used to get a table of congruified ages that can be used as calibrations

```{r}
# Cross congruify source chronograms and target tree
fix_congruification <- function(congruified_self,
                                  index = 1,
                                  extracted_calibrations) {
    source_chronogram <- congruified_self[[index]]$self$reference
    target_tree <- congruified_self[[index]]$phy$reference
    xx <- target_tree$node.label
     # if number of congruified nodes in target tree is the same as node number from source chronogram
    if (length(xx[ xx!= ""]) == source_chronogram$Nnode) {
        # take extracted calibrations data frame as is
        calibs <- extracted_calibrations[[index]]
        calibs$congruent <- rep(TRUE, nrow(calibs))
        calibs_matched <- mrca_calibrations(phy = source_chronogram,
                                            calibrations = calibs)
        # TODO: check that the following works:
        # col_names <- colnames(calibs_matched$matched_calibrations)
        # ii <- col_names %in% c("mrca_node_number", "mrca_node_name")
        # col_names[ii] <- c("source_mrca_node_number", "source_mrca_node_name")
        # colnames(calibs_matched$matched_calibrations) <- col_names
      } else {
        # case for index = 11, 10, 8, 7
        # take congruified calibrations data frame
        # TODO: needs checking, as the number of congruent nodes in target tree is higher than nodes available in output$calibrations, meaning, that some valid node data is being dropped from final results for some reason
        # first, get most complete calibrations data.frame
        # it is usually the one extracted from target tree
        calibs <- congruified_self[[index]]$phy$calibrations
        calibs_matched <- mrca_calibrations(phy = source_chronogram,
                                            calibrations = calibs)
        # use actual ages from source chronogram
        mrca_node_name <- calibs_matched$matched_calibrations$mrca_node_name
        source_bt <- ape::branching.times(calibs_matched$matched_phy)
        source_ages <- source_bt[mrca_node_name]
        calibs$MaxAge <- calibs$MinAge <- unname(source_ages)
        # add congruent column
        calibs$congruent <- rep(TRUE, nrow(calibs))
        # add source node numbers and names
        calibs$source_mrca_node_name <- mrca_node_name
        calibs$source_mrca_node_number <- as.numeric(sub("n", "", mrca_node_name))
        # add nodes that were not congruent
        nn <- is.na(match(names(source_bt), mrca_node_name))
        nn_ages <- ape::branching.times(calibs_matched$matched_phy)[nn]
        nn_node_name <- names(nn_ages)
        nn_node_number <- as.numeric(sub("n", "", nn_node_name))

        non_congruent <- data.frame(MinAge = unname(nn_ages),
                                    MaxAge = unname(nn_ages),
                                    congruent = rep(FALSE, length(nn_ages)),
                                    source_mrca_node_number = nn_node_number,
                                    source_mrca_node_name = nn_node_name)
        calibs <- data.table::rbindlist(list(calibs, non_congruent),
                                        fill = TRUE)
      }
    calibs$reference <- rep(names(congruified_self)[index], nrow(calibs))
    return(calibs)
  }
  # TODO: fix error in cross validation data set when running the following:
  fixed_congruification <- lapply(seq(congruified_self),
                                  function(i) {
                                    tryCatch(fix_congruification(congruified_self,
                                                        index = i,
                                                        extracted_calibrations),
                                              error = function(e) NA) } )
  errored <- is.na(fixed_congruification)
  if (any(errored)) {
    message("fixing congruification errored for:\n",
              paste(which(errored), "--", names(congruified_self)[errored], "\n"))
  }
  ##############################################################################
  ##############################################################################
  # Merge calibrations data frames into a single one
  # and match to phy nodes
  ##############################################################################
  ##############################################################################
  calibrations <-   data.table::rbindlist(fixed_congruification[!errored], fill = TRUE)
  calibrations_results <- mrca_calibrations(phy = phy,
                                            calibrations = calibrations)
  return(structure(calibrations_results,
                   class = "congruifiedCalibrations"))
  
```


