---
title: "Plots for new figure 1"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Plots for new figure 1}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
## Getting source chronograms

```{r setup}
# library(datelifeMS1)
devtools::load_all("../../datelifeplot")
devtools::load_all()
frin_sample_dsumm$phylo_median
spp <- frin_sample_dsumm$phylo_median$tip.label[1:6]
spp_dq <- datelife::make_datelife_query(spp)
spp_dr <- datelife::get_datelife_result(spp_dq)
```

As you can see in figure X, there is one chronogram that has inconsistent ages for one node relative to other node ages. It is Study 6, Burns et al.
To produce a figure that would showcase the datelife workflow in a clear and straightforward way, we decided to keep only 3 source chronograms:

Study 1, and
Study 3 - Chronogram 1 and 2.

These studies correspond to indices 1, 4 and 5 from the datelife result, so we are going to prune those out:

```{r}
spp_dr_fig1 <- spp_dr[c(1,4,5)]
names(spp_dr_fig1)
names(spp_dr)[c(1,4,5)] == names(spp_dr_fig1)
```


```{r}
class(spp_dr_fig1) <- "datelifeResult"
spp_summ <- datelife:::summary.datelifeResult(datelife_query = spp_dq, object = spp_dr_fig1)
spp_summ$phylo_all

mock_spp_names <- gsub(" ", "_", spp_summ$phylo_median$tip.label)
names(mock_spp_names) <- LETTERS[1:6]
mock_spp_names

mock_study_names <- names(spp_summ$phylo_all)
iis <- match(mock_study_names, unique(mock_study_names))
names(mock_study_names) <- paste("Study", iis)
names(mock_study_names)
ffs <- table(names(mock_study_names))

is_dupl <- names(mock_study_names) %in% names(ffs)[ffs > 1]

# Add "Chronogram X" to duplicated studies:

for (i in names(ffs)[ffs > 1]) {
  dds <- i == names(mock_study_names)
  chrono_names <- paste("- Chronogram", seq(ffs[[i]]))
  names(mock_study_names)[dds] <- paste(names(mock_study_names)[dds],chrono_names )
}
```

### Rename source chronograms:

```{r}
phylo_all_renamed <- c()
for (phy in spp_summ$phylo_all){
  is_in_tree <- match(gsub(" ", "_", phy$tip.label), mock_spp_names)
  phy$tip.label <- names(mock_spp_names)[is_in_tree]
  phylo_all_renamed <- c(phylo_all_renamed, list(phy))
}
class(phylo_all_renamed) <- "multiPhylo"
names(phylo_all_renamed) <- names(mock_study_names)
```

### Plotting ALL source chronograms

I combined source chronogram plots with ppt.
```{r}
# as pdf:
plot_max_time <- 30
datelifeplot::plot_phylo_all(chronograms = phylo_all_renamed,
                             write = "pdf",
                             folder_name = "../figures/figure1/source_chronograms",
                             file_name = "phylo_all",
                             plot_type = "phyloch",
                             max_depth = plot_max_time,
                             edge.width = 3,
                             cex_tiplabels = 2,
                             cex_axislabel = 1.5,
                             cex_axis = 2,
                             cex_title = 2.5,
                             pos_title = 3,
                             pos_axis = 2,
                             axis_label = "",
                             units = NULL,
                             plot_height = 250,
                             plot_width = 5)
```
## Get calibrations

Rename median chronogram tip labels:

```{r}
phylo_median <- spp_summ$phylo_median
is_in_tree <- match(gsub(" ", "_", phylo_median$tip.label), mock_spp_names)
phylo_median$tip.label <- names(mock_spp_names)[is_in_tree]
```

Get calibrations:

```{r}
all_calibrations <- datelife::extract_calibrations_phylo(
  input = phylo_all_renamed,
  each = FALSE)
matched <- datelife::match_all_calibrations(
  phy = phylo_median, 
  calibrations = all_calibrations)
calib_summ <- datelife:::summary.matchedCalibrations(
  matched$matched_calibrations)

```
## Phylo median tree as a phylogram (all branches equal length)

```{r}
# visualize median branching times:
ape::branching.times(phylo_median)
# force equal branch lengths, with root age = 1
pp <- ape::compute.brlen(phylo_median, power = 1)
# visualize the new branch lengths:
ape::branching.times(pp)
pp$edge.length
plot(pp)
ape::axisPhylo(side = 1)
# to make branch lengths proportional to a certain age we have o multiply them by that age, in this case it's the total plot_max_time
pp$edge.length <- pp$edge.length*45 
# I decided to make the two oldest branches shorter:
pp$edge.length[pp$edge.length == 27] <- 9
```

Plot ages from source chronograms on top of opentree synth tree:

```{r}
# Get references in matched calibrations:
studies <- as.character(unique(calib_summ$in_phy$reference))
# order references alphabetically:
studies <- studies[order(studies)]
#create a vector of colors with study names:
color_studies <- viridis::turbo(n = 9)
color_studies <- color_studies[c(2,8,5)]
names(color_studies) <- studies

plot_max_time <- 30

svg(filename = "../figures/figure1/opentree_source_ages.svg",
    height = 5,
    width = 5)
datelifeplot::plot_node_ages(pp,
               title = c(""),
               cex_title = 2,
               plot_type = "phyloch",
               time_depth = plot_max_time,
               pos_axis = 2,
               cex_axis = 1.3,
               edge.width = 3,
               cex_tiplabels = 2,
               calibration_summary = calib_summ,
               pch = 18,
               color_pch = color_studies,
               cex_pch = 2,
               lwd_bars = 10,
               cex_legend = 1,
               axis_label = "",
               x_legend = 0,
               y_legend = 7.5,
               mai2 = 0
               )
ape::nodelabels(text = pp$node.label, frame = "circle", bg = "white")
dev.off()
```
## Plot tree topology with node labels and no time axis

```{r}
plot_max_time <- 30
svg(filename = "../figures/figure1/opentree_topology.svg",
    height = 5,
    width = 5)
datelifeplot::plot_chronogram(pp,
                         title = c("Open Tree of Life synthetic",
                                   "tree topology"),
                         time_depth = plot_max_time,
                         time_axis = TRUE,
                         axis_label = "",
                         edge.width = 3,
                         cex_tiplabels = 2,
                         cex_axislabel = 1.5,
                         cex_axis = 2,
                         cex_title = 2,
                         pos_title = 3,
                         units = NULL
                             )
ape::nodelabels(text = pp$node.label, frame = "circle", bg = "white")
dev.off()
```


## Get pairwise median age points and plot them:

```{r}

names(spp_dr_fig1)
spp_dr_fig1[[1]]

best_grove <- datelife::get_best_grove(spp_dr_fig1)

median_matrix <- datelife::datelife_result_median_matrix(best_grove$best_grove)
colnames(median_matrix)

median_matrix_list <- list(summary_matrix = median_matrix)

median_matrix_table <- datelife::matrices_to_table(median_matrix_list)

matched_median_ages <- datelife::match_all_calibrations(phy = spp_summ$phylo_median, calibrations = median_matrix_table)

mm_summ <- datelife:::summary.matchedCalibrations(matched_median_ages$matched_calibrations)

mm_summ$in_phy

## rename

mm_summ$in_phy$taxonA

is_in_taxonA <- match(gsub(" ", "_", mm_summ$in_phy$taxonA), mock_spp_names)
mm_summ$in_phy$taxonA <- names(mock_spp_names)[is_in_taxonA]

is_in_taxonB <- match(gsub(" ", "_", mm_summ$in_phy$taxonB), mock_spp_names)
mm_summ$in_phy$taxonB <- names(mock_spp_names)[is_in_taxonB]
```


Table:

```{r}
colnames(mm_summ$in_phy)
pretty <- mm_summ$in_phy[,c("taxonA", "taxonB", "nodeAge", "mrca_node_number")]
pretty
# Eliminate duplicated taxon pairs:
## Find the duplicated pairs
taxon_pairs <- c()
for (i in seq(nrow(pretty))) {
  taxon_pair <- as.character(pretty[i, c("taxonA", "taxonB")])
  taxon_pair <- taxon_pair[order(taxon_pair)]
  taxon_pair <- paste(taxon_pair, collapse = "")
  taxon_pairs <- c(taxon_pairs, taxon_pair)
}
## Eliminate them:
pretty <- pretty[!duplicated(taxon_pairs),]

# Reorder ROWS by node number and taxon names
pretty <- pretty[order(pretty$mrca_node_number, pretty$taxonA, pretty$taxonB),]
# Eliminate rownames
rownames(pretty) <- NULL
# Reorder COLUMNS
pretty <- pretty[, c("mrca_node_number", "taxonA", "taxonB", "nodeAge")]
# Modify column names to be informative
colnames(pretty)[4] <- "Median Pairwise Age"
colnames(pretty)[1] <- "Node Name"
# Change node numbers my node names
pretty$`Node Name` <- paste0("n", pretty$`Node Name`-6)

# Print table as pdf
pdf("../figures/figure1/median_table.pdf")       # Export PDF
gridExtra::grid.table(pretty) # if you do not want row numbers set rows = NULL
dev.off()
```



```{r}
color_reference <- "blue"

mm_summ$in_phy$reference <- rep("Median pairwise \nnode ages\n", nrow(mm_summ$in_phy))
length(unique(mm_summ$in_phy$nodeAge))
# there are 8 different median pairwise ages

names(color_reference) <- c("Median pairwise \nnode ages\n")

svg(filename = "../figures/figure1/median_ages.svg",
    height = 5,
    width = 5.5)
datelifeplot::plot_node_ages(phylo_median,
               title = c("Median Summary Chronogram"),
               pos_title = 3,
               cex_title = 2,
               plot_type = "phyloch",
               time_depth = plot_max_time,
               pos_axis = 2,
               cex_axis = 1.3,
               edge.width = 3,
               cex_tiplabels = 2,
               calibration_summary = mm_summ,
               pch = "x",
               color_pch = color_reference,
               cex_pch = 2,
               lwd_bars = 15,
               cex_legend =1.4,
               x_legend = 0,
               y_legend = 7,
               mai2 = 0,
               axis_label = ""
               )
dev.off()
```
             
# SDM tree

```{r}
phylo_sdm <- spp_summ$phylo_sdm
is_in_tree <- match(gsub(" ", "_", phylo_sdm$tip.label), mock_spp_names)
phylo_sdm$tip.label <- names(mock_spp_names)[is_in_tree]

best_grove <- datelife::get_best_grove(spp_dr_fig1)

sdm_matrix <- datelife::datelife_result_sdm_matrix(best_grove$best_grove)
colnames(sdm_matrix)

sdm_matrix_list <- list(summary_matrix = sdm_matrix)

sdm_matrix_table <- datelife::matrices_to_table(sdm_matrix_list)

matched_sdm_ages <- datelife::match_all_calibrations(phy = spp_summ$phylo_sdm, calibrations = sdm_matrix_table)

sdm_summ <- datelife:::summary.matchedCalibrations(matched_sdm_ages$matched_calibrations)

sdm_summ$in_phy

## rename

sdm_summ$in_phy$taxonA

is_in_taxonA <- match(gsub(" ", "_", sdm_summ$in_phy$taxonA), mock_spp_names)
mm_summ$in_phy$taxonA <- names(mock_spp_names)[is_in_taxonA]

is_in_taxonB <- match(gsub(" ", "_", sdm_summ$in_phy$taxonB), mock_spp_names)
mm_summ$in_phy$taxonB <- names(mock_spp_names)[is_in_taxonB]
```


Plot:

```{r}
color_reference <- "purple"

sdm_summ$in_phy$reference <- rep("SDM pairwise \nnode ages\n", nrow(sdm_summ$in_phy))

names(color_reference) <- c("SDM pairwise \nnode ages\n")

svg(filename = "../figures/figure1/sdm_ages.svg",
    height = 5,
    width = 7)
datelifeplot::plot_node_ages(phylo_sdm,
               title = c("SDM Summary Chronogram"),
               cex_title = 2,
               plot_type = "phyloch",
               time_depth = plot_max_time,
               pos_axis = 2,
               cex_axis = 1.3,
               edge.width = 3,
               cex_tiplabels = 2,
               calibration_summary = sdm_summ,
               pch = "|",
               color_pch = color_reference,
               cex_pch = 2,
               lwd_bars = 15,
               cex_legend = 1.5,
               x_legend = 0,
               y_legend = 6
               )
dev.off()
```
