---
title: "figure-fringillidae"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{figure-fringillidae}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
devtools::load_all("../../datelifeplot")
writing_dir <- "../figures/figure-fringillidae/"
```

## Get a `datelifeQuery` object

```{r}
frin_dquery <- datelife::make_datelife_query(input = "fringillidae",
                                             use_tnrs = TRUE,
                                             get_spp_from_taxon = TRUE)
```

## Get a `datelifeResult` object

```{r, eval=FALSE}
frin_dres = datelife::get_datelife_result(input = frin_dquery, cache = opentree_chronograms)
frin_dres_file <- paste0("frin_dres_", Sys.Date())
usethis::use_data(frin_dres, name = frin_dres_file, overwrite = TRUE)
```

```{r, include = FALSE}
load(file = "../data/frin_dres.rda")
```

## Get a `datelifeSummary` object
```{r}
frin_dsumm <- datelife:::summary.datelifeResult(datelife_query = frin_dquery, object = frin_dres)

# usethis::use_data(frin_dsumm, overwrite = TRUE)
```
## Save objects and load them

```{r}

# usethis::use_data(frin_sample_dquery, overwrite = TRUE)

```
Make an object with all source chronograms, and reorder chronogram list alphabetically:

```{r}
phylo_all <- frin_dsumm$phylo_all
names(phylo_all)
alphabetic_order <- order(names(phylo_all))
phylo_all <- phylo_all[alphabetic_order]
names(phylo_all)
```
We will rename source chronograms to have a shorter study name:

```{r}

names(phylo_all) <- c("Barker et al. 2013",  # 1
                      "Barker et al. 2015 - chronogram 1",  # 2
                      "Barker et al. 2015 - chronogram 2",  # 3
                      "Burns et al. 2015",  # 4
                      "Claramunt et al. 2015",  # 5
                      "Gibb et al. 2015", # 6
                      "Hedges et al. 2015 - chronogram 1",  # 7
                      "Hedges et al. 2015 - chronogram 2",  # 8
                      "Hooper et al. 2017",  # 9
                      "Jetz et al. 2012 - chronogram 1",  # 10
                      "Jetz et al. 2012 - chronogram 2",  # 11
                      "Kimball et al. 2019 - chronogram 1", # 12
                      "Kimball et al. 2019 - chronogram 2", # 13
                      "Oliveros et al. 2019", # 14
                      "Price et al. 2014 - chronogram 1", # 15
                      "Price et al. 2014 - chronogram 2", # 16
                      "Roquet et al. 2014 - chronogram 1", # 17
                      "Roquet et al. 2014 - chronogram 2", # 18 
                      "Uyeda et al 2017") # 19
                      

```

### Plotting source chronograms

getting oldest age of all trees in a `multiPhylo ` object:

```{r}
max(unlist(lapply(phylo_all, ape::branching.times)))

plot_max_time <- 45
```

Plot as pdf:

```{r}
datelifeplot::plot_phylo_all(chronograms = phylo_all,
                             folder_name = paste0(writing_dir, "/source_phylo_all"),
                             plot_height = 890,
                             plot_width = 600,
                             mai1 = 1,
                             mai2 = 0,
                             mai3 = 1,
                             mai4 = 5,
                             omi1 = 1,
                             omi2 = 0,
                             omi3 = 0,
                             omi4 = 0,
                             write = "pdf",
                             plot_type = "phyloch",
                             max_depth = plot_max_time,
                             edge.width = 2,
                             cex_tiplabels = 2,
                             cex_axislabel = 1.5,
                             cex_axis = 2,
                             cex_title = 2.5,
                             pos_title = -2.8,
                             pos_axis = 2,
                             axis_label = "")
```


## Get calibrations

Get calibrations for the topology in `frin_dsumm$phylo_median`

```{r}
phylo_median <- frin_dsumm$phylo_median
all_calibrations <- datelife::extract_calibrations_phylo(
  input = phylo_all,
  each = FALSE)
matched <- datelife::match_all_calibrations(
  phy = phylo_median, 
  calibrations = all_calibrations)
calib_summ <- datelife:::summary.matchedCalibrations(
  matched$matched_calibrations)
nrow(calib_summ$in_phy)
```

### Table of matched calibrations

Make it pretty.

```{r}
colnames(calib_summ$in_phy)
# keep columns of interest:
pretty_calibs <- calib_summ$in_phy[,c("taxonA", "taxonB", "MaxAge", "mrca_node_number", "reference")]
pretty_calibs
# Eliminate duplicated taxon pairs:
## Find the duplicated pairs within studies:
taxon_pairs_ref <- c()
for (i in seq(nrow(pretty_calibs))) {
  tax_pair_ref <- as.character(pretty_calibs[i, c("taxonA", "taxonB", "reference")])
  tax_pair_ref <- tax_pair_ref[order(tax_pair_ref)]
  tax_pair_ref <- paste(tax_pair_ref, collapse = "")
  taxon_pairs_ref <- c(taxon_pairs_ref, tax_pair_ref)
}
## Eliminate duplicated rows:
if (any(duplicated(taxon_pairs_ref))) {
  pretty_calibs <- pretty_calibs[!duplicated(taxon_pairs_ref),]
}

# Reorder ROWS by node number, reference, age and taxon names
pretty_calibs <- pretty_calibs[order(pretty_calibs$mrca_node_number, pretty_calibs$MaxAge, pretty_calibs$reference, pretty_calibs$taxonA, pretty_calibs$taxonB),]
# Eliminate rownames
rownames(pretty_calibs) <- NULL
# Reorder COLUMNS
pretty_calibs <- pretty_calibs[, c("mrca_node_number", "taxonA", "taxonB", "MaxAge", "reference")]
# Modify column names to be informative
colnames(pretty_calibs) <- c("Node Name", "taxon A", "taxon B", "Node Age", "Study chronogram")
# Change node numbers my node names
pretty_calibs$`Node Name` <- paste0("n", pretty_calibs$`Node Name`-6)

# substitute underscore by space:
pretty_calibs$`taxon A` <- sub("_", " ", pretty_calibs$`taxon A`)
pretty_calibs$`taxon B` <- sub("_", " ", pretty_calibs$`taxon B`)

# order taxa alphabetically by column
for (i in seq(nrow(pretty_calibs))) {
  original <- as.character(pretty_calibs[i, c("taxon A", "taxon B")])
  reordered <- original[order(original)]
  pretty_calibs[i, c("taxon A", "taxon B")] <- reordered
}


pretty_calibs
```

Print table as pdf.
Trick to print in multiple pages from https://stackoverflow.com/questions/15937131/print-to-pdf-file-using-grid-table-in-r-too-many-rows-to-fit-on-one-page

```{r}
## set the theme to have italics in species names:
number_of_rows <- nrow(pretty_calibs)
tt1 <- gridExtra::ttheme_default(core=list(fg_params=list(fontface=c(rep("plain", number_of_rows), rep("italic", number_of_rows*2), rep("plain", number_of_rows*2)))))

# Export PDF
pdf(paste0(writing_dir, "/calibs_table.pdf"), width = 10)
gridExtra::grid.table(pretty_calibs, theme = tt1) # if you do not want row numbers set rows = NULL
dev.off()
```

### Plot matched calibrations from source chronograms.

Choose a color scheme for the studies:

```{r}
# Get references in matched calibrations:
studies <- as.character(unique(calib_summ$in_phy$reference))


# order references alphabetically:
studies <- studies[order(studies)]
studies
```

Visualize plotting colors:

```{r}
#create a vector of colors with study names:
number_of_colors <- length(studies) * 10


xx <- viridis::turbo(n = number_of_colors)


plot(-1:number_of_colors + 1, -1:number_of_colors + 1, type = "n")
text(1:number_of_colors, 1:number_of_colors, xx, col = xx)

# xx <- viridis::turbo(n = 20)
# plot(-2:length(xx)+1, -2:length(xx)+1, type = "n")
# text(1:(length(xx)), 1:length(xx), paste(1:length(xx),"=",xx), col = xx)

```

Choose plotting colors:

```{r}
#create a vector of colors with study names:
# color_studies <- xx[c(12,
#                       3,4,
#                       15,
#                       18,17,
#                       20,
#                       8,9)]
# 

length(xx) / length(unique(names(frin_dsumm$phylo_all)))
color_studies <- c("blue", 
                   xx[15:16],
                   xx[1],
                   "purple", 
                   "orchid",
                   xx[65:66],
                   "pink",
                   xx[c( 95:96, 115:116)],
                   "skyblue",
                   xx[c(145:146, 175:176, 190)])
names(color_studies) <- studies

```


Now, we can plot:

```{r}
svg(filename = paste0(writing_dir, "/node_ages_matched.svg"),
    height = 30,
    width =  15)
datelifeplot::plot_node_ages(phylo_median,
               title = "Median Summary Chronogram",
               plot_type = "phyloch",
               time_depth = plot_max_time,
               calibration_summary = calib_summ,
               pch = 19,
               color_pch = color_studies,
               cex_pch = 1,
               lwd_bars = 10,
               cex_legend = 0.7,
               x_legend = 0 ,
               y_legend = 270,
               mai1 = 1, 
               mai2 = 0,
               mai3 = 0.5,
               mai4 = 3.73,
               omi1 = 2,
               omi2 = 1,
               omi3 = 0.5,
               omi4 = 0,
               edge.width = 2,
               cex_tiplabels = 0.5,
               cex_axislabel = 1,
               cex_axis = 1,
               cex_title = 2,
               pos_title = -1,
               pos_axis = 2
               )
dev.off()
```
## Get pairwise median age points

```{r}
best_grove <- datelife::get_best_grove(frin_dres)

# Take a list of chronograms as tables and summarize age data per taxon pair into a single matrix:
median_matrix <- datelife::datelife_result_median_matrix(best_grove$best_grove)
colnames(median_matrix)

median_matrix_list <- list(summary_matrix = median_matrix)

median_matrix_table <- datelife::matrices_to_table(median_matrix_list)

# removing NAs:

median_matrix_table <- median_matrix_table[!is.na(median_matrix_table$taxonA),]

# match taxon pairs to nodes on a tree topology

matched_median_ages <- datelife::match_all_calibrations(phy = phylo_median, calibrations = median_matrix_table)

mm_summ <- datelife:::summary.matchedCalibrations(matched_median_ages$matched_calibrations)

nrow(mm_summ$in_phy)
```


### Plotting source ages AND summary ages on the same dated tree

Set colors and data:

```{r}
color_median <- "purple"
mm_summ$in_phy$reference <- rep("Pairwsie median of node ages", length(mm_summ$in_phy$reference))
names(color_median) <- mm_summ$in_phy$reference[1]

point_colors <- list(a1 = color_studies, a2 = color_median)
all_ages <- list(a1 = calib_summ, a2 = mm_summ)
point_type <- list(a1 = "/", a2 = "•")

leg_color<- c(color_studies, color_median)
# leg_pch <- c(rep("|", length(color_studies)), "•")
leg_pch <- rep(19, length(leg_color))
leg_text <- names(leg_color)
```

Plot:

```{r}
svg(filename = paste0(writing_dir, "/median_and_calibration_ages.svg"),
    height = 30,
    width = 15)
datelifeplot::plot_node_ages2(phylo_median,
               title = c("Median Summary Chronogram"),
               pos_title = 3,
               cex_title = 2,
               plot_type = "phyloch",
               time_depth = plot_max_time,
               pos_axis = 2,
               cex_axis = 1,
               edge.width = 3,
               cex_tiplabels = 0.5,
               matched_ages  = all_ages,
               pch = point_type,
               pch_color = point_colors,
               pch_cex = 2,
               bars_lwd = 10,
               bars_color = "#DCDCDC50",
               legend_cex = 0.8,
               legend_x = -5,
               legend_y = 280,
               legend_text = leg_text,
               legend_pch = leg_pch,
               legend_color = leg_color,
               mai1 = 1,
               mai2 = 0.8,
               mai4 = 3.6,
               axis_label = ""
               )
dev.off()
```        

Simplifying median ages on tree plot:

```{r}
# Getting the median per node:
median_singletons <- c()
for (i in unique(mm_summ$in_phy$mrca_node_name)) {
  print(i)
  x <- median(unique(mm_summ$in_phy$nodeAge[mm_summ$in_phy$mrca_node_name == i]))
  median_singletons <- c(median_singletons, x)
}

ape::branching.times(phylo_median)

# Detecting node names that are duplicated 
duplos <- duplicated(mm_summ$in_phy$mrca_node_name)
# Simplifying median ages:
mm_summ$in_phy$nodeAge[!duplos] <- median_singletons
# Removing duplicates:
mm_summ$in_phy <- mm_summ$in_phy[!duplos,]
# Columns MinAge and MaxAge need to be equal to nodeAge
mm_summ$in_phy$MinAge <- mm_summ$in_phy$MaxAge <- mm_summ$in_phy$nodeAge

```

Plotting parameters:

```{r}
mm_summ$in_phy$reference <- rep("Used as calibration", length(mm_summ$in_phy$reference))
# mm_summ$in_phy$reference[2] <- "Not used"
gplots::col2hex("purple") # "#A020F0"
color_median <- c("#A020F0")  # purple with no transparency

names(color_median) <- unique(mm_summ$in_phy$reference)

point_colors <- list(a1 = color_studies, a2 = color_median)
all_ages <- list(a1 = calib_summ, a2 = mm_summ)
point_type <- list(a1 = 19, a2 = "/")


leg_color<- list(a1 = color_studies, a2 = color_median)
leg_pch <- list(a1 = 19, a2 = "/")
leg_text <- lapply(leg_color, names)
leg_title <- c("Taxon pair node ages from studies", "Median of node ages")
leg_box <- c(TRUE, TRUE)
```

Plot again:
```{r}
pdf(file = paste0(writing_dir, "median_and_calibration_ages_simple.pdf"),
    height = 11.75*2,
    width = 8.25*2)
datelifeplot::plot_node_ages2(phylo_median,
               title = c("Median Summary Chronogram"),
               pos_title = -5,
               cex_title = 2,
               plot_type = "phyloch",
               time_depth = plot_max_time,
               pos_axis = 2,
               cex_axis = 1,
               edge.width = 2,
               cex_tiplabels = 0.6,
               matched_ages  = all_ages,
               pch = point_type,
               pch_color = point_colors,
               pch_transparency = 90,
               pch_cex = 1,
               bars_lwd = 10,
               bars_color = "#DCDCDC50",
               legend_cex = c(0.9, 0.9),
               legend_x = c(-10, -10),
               legend_y = c(190, 150),
               legend_text = leg_text,
               legend_pch = leg_pch,
               legend_color = leg_color,
               legend_title = leg_title,
               legend_box = leg_box,
               mai1 = 0,
               mai2 = 2.5,
               mai3 = 1,
               mai4 = 3.6,
               omi1 = 0,
               omi2 = 0,
               omi3 = 0,
               axis_label = ""
               )
dev.off()
```      

