---
title: "figure-cross-validation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{figure-cross-validation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(datelifeMS1)
devtools::load_all("../../datelifeplot")
devtools::load_all("../../datelife")
writing_dir <- "../figures/figure-cross-validation/"

# Based on function use_othercals3 from https://github.com/LunaSare/datelife_examples/blob/master/R/functions_data.R#L36
use_other_calibrations <- function(trees, other_calibrations, ...){
    res <- lapply(seq(trees), function(i){
      phy <- trees[[i]]
      phy$edge.length <- NULL
      print(i)
        xx <- suppressMessages(suppressWarnings(datelife::use_calibrations_bladj(phy,
            calibrations = other_calibrations[[i]], ...)))
        return(xx)
    })
    class(res) <- "multiPhylo"
    names(res) <- names(trees)
    res
}

# Function based on https://github.com/LunaSare/datelife_examples/blob/master/R/functions_data.R#L3
# get calibrations form all other chronograms not belonging to the same study
extract_other_calibrations <- function(phyloall){
    # dd <- duplicated(names(phyloall))
    res <- lapply(seq(phyloall), function(i){
        dd <- names(phyloall) %in% names(phyloall)[i]
        datelife::extract_calibrations_phylo(phyloall[!dd], each = FALSE)
    })
    names(res) <- names(phyloall)
    res
}
```

## Cross validation of Fringillidae source chronograms.

We are using the object `frin_dsumm` and getting all source chronograms, and reordering the chronogram list alphabetically:

```{r}
phylo_all <- frin_dsumm$phylo_all
names(phylo_all)
alphabetic_order <- order(names(phylo_all))
phylo_all <- phylo_all[alphabetic_order]
```

Next, get calibrations for all other chronograms:

```{r}
frin_other_calibs <- extract_other_calibrations(phylo_all)
class(frin_other_calibs)
length(frin_other_calibs)
names(frin_other_calibs) <- names(phylo_all)
```

Date the topology based on other calibrations:

```{r}
cross_validation <- use_other_calibrations(trees = phylo_all, other_calibrations = frin_other_calibs)

# target = the tree topology that is going to be dated
# reference = the chronogram with dates to time-scale `target`
cc1 <- geiger::congruify.phylo(reference = phylo_all[[1]],
                        target = phylo_median,
                        scale = NA,
                        ncores = 1)
ls(cc1)
cc1_alt <- geiger::congruify.phylo(reference = phylo_median,
                        target = phylo_all[[1]],
                        scale = NA,
                        ncores = 1)$calibrations

cc_all <- geiger::congruify.phylo(reference = c(list(phylo_median), phylo_all),
                        target = phylo_median,
                        scale = NA,
                        ncores = 1)
names(cc_all) <- c("phylo_median", names(phylo_all))
unname(which(is.na(cc_all)))
cc_all[is.na(cc_all)]
```

Congruification of all chronograms at the same time returns `NA` for some chronograms (`r paste0(names(cc)[is.na(cc)], collapse, = ", ")`).

So, I am testing congruification one-on-one, on chronograms with index = 5, 6, 12, 13, 14, 15, 19:

```{r}
# Function
# Congruify a source chronogram with the target tree and itself
run_congruification <- function(phylo_median, phylo_all, index) {
  
    self <- phylo_all[[index]]
    phylo_median$tip.label <- sub(" ", "_", phylo_median$tip.label)
    self$tip.label <- sub(" ", "_", self$tip.label)
    cc <- geiger::congruify.phylo(reference = list(median = phylo_median, 
                                                   self = self),
                            target = self,
                            scale = NA,
                            ncores = 1)
    names(cc) <- c("median", "self")
    cc
}
```


Test: congruification one-on-one

```{r}
names(phylo_all)[5]

cc5 <- run_congruification(phylo_median, phylo_all, index = 5)

cc5$median$reference$node.label[cc5$median$reference$node.label != ""]
cc5$self$reference$node.label
cc5$self$target$hash
cc5$median$target$hash


pdf(file = paste0(writing_dir, "cc5_self_reference.pdf"))
ape::plot.phylo(cc5$self$reference)
ape::nodelabels(cc5$self$reference$node.label, adj = 0)
dev.off()

pdf(file = paste0(writing_dir, "cc5_median_reference.pdf"),
    height = 25)
ape::plot.phylo(cc5$median$reference, cex = 0.5)
ape::nodelabels(cc5$median$reference$node.label, adj = 0, cex = 0.5)
dev.off()

cc15 <- run_congruification(phylo_median, phylo_all, index = 15)

cc15$median$reference$node.label[cc15$median$reference$node.label != ""]
cc15$self$reference$node.label
cc15$self$target$hash
cc15$median$target$hash
```

Resut: The chronograms do not return NA when congruified individually.

Next test: Check that hashes from congruification do correspond to the same node in target and reference
```{r}
# Function:
# Plot congruified trees with hash labels as node labels
# @description plots an output of run_congruification
plot_congruification <- function(cc, 
                                 id = "",
                                 title_text = "",
                                 writing_dir, 
                                 height_self = 7, 
                                 height_target = 25) {
  phylo <- cc$self$reference
  pho <- phylo_height_omi(phy = phylo)
  pdf(file = paste0(writing_dir, 
                    deparse(substitute(cc)),
                    id,
                    "_congruification_reference2self.pdf"),
      height = pho$height/72)
  ape::plot.phylo(phylo)
  title(main = c(title_text, "source chronogram to self"))
  ape::nodelabels(phylo$node.label, adj = 0)
  dev.off()

  phylo <- cc$median$reference
  pho <- phylo_height_omi(phy = phylo)
  pdf(file = paste0(writing_dir,
                    deparse(substitute(cc)),
                    id,
                    "_congruification_reference2target.pdf"),
      height = pho$height/72)
  ape::plot.phylo(phylo, cex = 0.5)
  title(main = c(title_text, "source chronogram to topology of interest"))
  ape::nodelabels(phylo$node.label, adj = 0, cex = 0.5)
  dev.off() 
}
```

```{r}
# run congruification and plot for all source chronograms

for (i in seq(phylo_all)) {
  cc <- c()
  title_text <- paste(i, "--", names(phylo_all)[i])
  cat(title_text, "\n")
  cc <- suppressWarnings(run_congruification(phylo_median, phylo_all, index = i))
  plot_congruification(cc = cc, 
                       id = ifelse(i<10, paste0(0, i), i), 
                       writing_dir = writing_dir,
                       title_text = title_text)
}
```
Conclusion: It seems like the hash is coherent on some chronograms (e.g., `chronograms[12]`), but not all (e.g., `chronograms[19]`, hash 6191e33766a4bcc213ce01bd8914ed80)

Create a report of congruified nodes and hashes:

```{r}
filename <- paste0(writing_dir, "congruified_node_labels.txt")
separator <- paste(rep("#", 50), collapse = "")
write("", 
      file = filename)
for (i in seq(phylo_all)) {

  title_text <- paste(i, "--", names(phylo_all)[i])

  write(paste0(separator, "\n\n", title_text, "\n"), 
        file = paste0(writing_dir, "congruified_node_labels.txt"), 
        append = TRUE)
  
  cc <- c()
  cc <- suppressWarnings(run_congruification(phylo_median, phylo_all, index = i))

  xx <- cc$median$reference$node.label[cc$median$reference$node.label != ""]
  write(paste0("\nTarget topology to source chronogram node labels -- n = ", 
               length(xx),
               "\n"), 
        file = filename, 
        append = TRUE)
  write(xx, 
        file = filename, 
        append = TRUE)
  
  xx <- cc$self$reference$node.label
  write(paste0("\nSource chronogram to self node labels -- n = ", 
               length(xx),
               "\n"),
        file = filename, 
        append = TRUE)
  write(xx, 
        file = filename, 
        append = TRUE)
  
  xx <- cc$median$target$hash
  write(paste0("\nTarget to target hash -- n = ", 
               length(xx),
               "\n"), 
        file = filename, 
        append = TRUE)
  write(xx, 
        file = filename, 
        append = TRUE)
  
  xx <- cc$self$target$hash
  write(paste0("\nSource chronogram to target hash -- n = ", 
               length(xx),
               "\n"), 
        file = filename, 
        append = TRUE)
  write(xx, 
        file = filename, 
        append = TRUE)
  write("\n", 
        file = filename, 
        append = TRUE)
}
```
Conclusion: Hash labels can be different between source chronograms (references), even if they correspond to the same node in target topology.

However, hashed nodes do correspond to the same clade between source chronograms (references).

Then, we can use taxon pairs from hashed nodes in source chronograms (references) to identify corresponding node labels on target.


Question: What part from congruification output do we need?

```{r}

cc_all <- vector(mode = "list")

for (i in seq(phylo_all)) {
  cat(i, "--", names(phylo_all)[i], "\n")
  cc <- c()
  cc <- suppressWarnings(run_congruification(phylo_median, phylo_all, index = i))
  cc_all <- c(cc_all, list(cc))
}
names(cc_all) <- names(phylo_all)
ls(cc_all)
ls(cc_all[[19]])
ls(cc_all[[19]]$median)
cc_all[[19]]$median$reference
cc_all[[19]]$median$calibrations
cc_all[[19]]$self$calibrations
# get congruified node labels:
# hashed nodes are always in cc$reference (not in cc$target)
for (i in seq(cc_all)){
    cat(names(cc_all[i]), "\n")
    cc <- cc_all[[i]]
    
    xx <- cc$self$reference$node.label
    cat("hashed source chronogram nodes = ",
        length(xx[xx != ""]),
        "\n")
    cat("source chronogram calibrations table row number = ", 
        length(cc$self$calibrations$MRCA),
        "\n\n")
    
    xx <- cc$median$reference$node.label
    cat("hashed median tree nodes = ",
        length(xx[xx != ""]),
        "\n")
    cat("median tree calibrations table row number = ", 
        length(cc$median$calibrations$MRCA),
        "\n\n\n")
}
```
Conclusion: Number of rows in calibrations object is different that hashed nodes in target or reference.


```
# Test that hash names are congruent in target and reference
for (i in seq(nn)) {
  xx <- match(nn[[1]], nn[[i]])
  print(i)
  print(xx)
  yy <- nn[[i]][xx]
  print(yy)
}
```

Question: Why are `output$calibrations` tables different between self and median??
I thought they should be the same...

```{r}
cc19 <- geiger::congruify.phylo(reference = phylo_all[[19]],
                        target = phylo_median,
                        scale = NA,
                        ncores = 1)
cc19 <- geiger::congruify.phylo(reference = phylo_median,
                        target = phylo_all[[19]],
                        scale = NA,
                        ncores = 1)
cc19 <- geiger::congruify.phylo(reference = phylo_all[[19]],
                        target = phylo_all[[19]],
                        scale = NA,
                        ncores = 1)
ls(cc19)
cc19$calibrations
cc19$reference$node.label[cc19$reference$node.label != ""]
```
Conclusion: `output$calibrations` table does not have all nodes that have been congruified... the horror!!!

Important to take into account when using `geiger::congruify.phylo()`

- Hashed node labels in `output$reference$node.label` seem to be correct and to be complete.
- however, these hashed node labels are not ordered correctly in `output$reference$node.label`.
- Some congruent nodes are dropped from `output$calibrations` because they do not contain all taxa (see function `geiger:::.build_calibrations()`.


TODO:

- match target tree node labels to extracted calibrations from source chronograms
- if number of hashed nodes in target tree is equal to row number of `output$calibrations`, take `output$calibrations` as is.
- if not, get absent hashes from`output$calibrations`
- get node label from target tree corresponding to absent hashes
- add them to `output$calibrations` data.frame

```{r}
# Cross congruify source chronograms and target tree to source_chronograms
run_congruification_self <- function(phy, chronograms, index) {
  self <- chronograms[[index]]
  phy$tip.label <- sub(" ", "_", phy$tip.label)
  self$tip.label <- sub(" ", "_", self$tip.label)
  cc <- geiger::congruify.phylo(reference = list(phy = phy,
                                                 self = self),
                          target = self,
                          scale = NA,
                          ncores = 1)
  names(cc) <- c("phy", "self")
  cc
}
congruified_self <- lapply(seq(phylo_all), 
                          function(i) run_congruification_self(phylo_median, 
                                                               phylo_all, 
                                                               index = i))
names(congruified_self) <- names(phylo_all)

ls(congruified_self[[19]])
congruified_self[[1]]$phy$calibrations
congruified_self[[1]]$self$calibrations

run_congruification_phy <- function(phy, chronograms, index) {
  self <- chronograms[[index]]
  phy$tip.label <- sub(" ", "_", phy$tip.label)
  self$tip.label <- sub(" ", "_", self$tip.label)
  cc <- geiger::congruify.phylo(reference = list(phy = phy,
                                                 self = self),
                          target = phy,
                          scale = NA,
                          ncores = 1)
  names(cc) <- c("phy", "self")
  cc
}

phy_calibrations <- suppressWarnings(
                      geiger::congruify.phylo(reference = phylo_median,
                                              target = phylo_median,
                                              scale = NA,
                                              ncores = 1)$calibrations)
phy_matched <- mrca_calibrations(phy = phylo_median, 
                                 calibrations = phy_calibrations)

  
congruified_all_phy <- lapply(seq(phylo_all), 
                          function(i) run_congruification_phy(phylo_median, 
                                                               phylo_all, 
                                                               index = i))
names(congruified_all_phy) <- names(phylo_all)


# Cross congruify source chronograms and target tree
index = 8

  fix_congruification <- function(congruified_self,
                                  index = 1,
                                  extracted_calibrations) {
    source_chronogram <- congruified_self[[index]]$self$reference
    target_tree <- congruified_self[[index]]$phy$reference
    xx <- target_tree$node.label
     # if number of congruified nodes in target tree is the same as node number from source chronogram
    if (length(xx[ xx!= ""]) == source_chronogram$Nnode) {
        # take extracted calibrations data frame as is
        calibs <- extracted_calibrations[[index]]
        calibs$congruent <- rep(TRUE, nrow(calibs))
        calibs_matched <- mrca_calibrations(phy = source_chronogram,
                                            calibrations = calibs)
        # TODO: check that the following works:
        # col_names <- colnames(calibs_matched$matched_calibrations)
        # ii <- col_names %in% c("mrca_node_number", "mrca_node_name")
        # col_names[ii] <- c("source_mrca_node_number", "source_mrca_node_name")
        # colnames(calibs_matched$matched_calibrations) <- col_names
      } else {
        # case for index = 11, 10, 8, 7
        # take congruified calibrations data frame
        # TODO: needs checking, as the number of congruent nodes in target tree is higher than nodes available in output$calibrations, meaning, that some valid node data is being dropped from final results for some reason
        # first, get most complete calibrations data.frame
        # it is usually the one extracted from target tree
        calibs <- congruified_self[[index]]$phy$calibrations
        calibs_matched <- mrca_calibrations(phy = source_chronogram,
                                            calibrations = calibs)
        # use actual ages from source chronogram
        mrca_node_name <- calibs_matched$matched_calibrations$mrca_node_name
        source_bt <- ape::branching.times(calibs_matched$matched_phy)
        source_ages <- source_bt[mrca_node_name]
        calibs$MaxAge <- calibs$MinAge <- unname(source_ages)
        # add congruent column
        calibs$congruent <- rep(TRUE, nrow(calibs))
        # add source node numbers and names
        calibs$source_mrca_node_name <- mrca_node_name
        calibs$source_mrca_node_number <- as.numeric(sub("n", "", mrca_node_name))
        # add nodes that were not congruent
        nn <- is.na(match(names(source_bt), mrca_node_name))
        nn_ages <- ape::branching.times(calibs_matched$matched_phy)[nn]
        nn_node_name <- names(nn_ages)
        nn_node_number <- as.numeric(sub("n", "", nn_node_name))

        non_congruent <- data.frame(MinAge = unname(nn_ages),
                                    MaxAge = unname(nn_ages),
                                    congruent = rep(FALSE, length(nn_ages)),
                                    source_mrca_node_number = nn_node_number,
                                    source_mrca_node_name = nn_node_name)
        calibs <- data.table::rbindlist(list(calibs, non_congruent),
                                        fill = TRUE)
      }
    calibs$reference <- rep(names(congruified_self)[index], nrow(calibs))
    return(calibs)
  }

fix_congruification(congruified_self, index = 8, extracted_calibrations)

fixed_congruification <- lapply(seq(congruified_self), 
                                  function(i) {
                                    fix_congruification(congruified_self,
                                                        index = i,
                                                        extracted_calibrations)})

calibrations <-   data.table::rbindlist(fixed_congruification, fill = TRUE)

calibrations_results <- mrca_calibrations(phy = phylo_median,
                                            calibrations = calibrations)
structure(calibrations_results,
                   class = "congruifiedCalibrations")
```

```{r}
extracted_calibrations <- extract_calibrations_phylo(input = phylo_all,
                                                         each = TRUE)
fixed_congruification <- lapply(seq(congruified_self), 
                                  function(i) {
                                    fix_congruification(congruified_self,
                                                        index = i,
                                                        extracted_calibrations)})
names(fixed_congruification) <- names(phylo_all)
fixed_congruification[[7]]
colnames(calibs_matched$matched_calibrations)

```
test that MRCA from congruification corresponds to the same node obtained with phytools::find.mrca()


