---
title: "figure-fringillidae-mean"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{figure-fringillidae-mean}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
devtools::load_all("../../datelifeplot")
devtools::load_all("../../datelife")
writing_dir <- "../figures/figure-fringillidae/"
```


```{r}
load("../data/fringillidae_v0.6.2.rda")
```

## Get all ages from source chronograms

Get all ages from source chronograms (calibrations) for the topology in `dsumm$phylo_median`.
First, get all calibrations from all source chronograms:

```{r}
phylo_median <- fringillidae_v0.6.2$dsumm$phylo_median
all_calibrations <- datelife::extract_calibrations_phylo(
  input = phylo_all,
  each = FALSE)
nrow(all_calibrations) # 1206
```

Now, get mrca of all_calibrations. 
We can do this with `match_all_calibrations` or `mrca_calibrations`.
```{r}
matched <- datelife::match_all_calibrations(
  phy = phylo_median, 
  calibrations = all_calibrations)
ls(matched)
nrow(matched$matched_calibrations) # 1206
```

```{r}
calib_summ <- datelife:::summary.matchedCalibrations(
  matched$matched_calibrations)
nrow(calib_summ$in_phy) # 1206
```


### Congruify ages and get matching mrca on target tree

Original workflow is going from a summary matrix/ patristic distance matrix to a phylo object. Does not congruify.
- Converts source chronograms to patristic distance matrices
- Combines them into a single summary matrix
- Creates a table of taxon pair summary ages
- Matches taxon pair summary ages to nodes on target tree and stores them as data frame
- Calculates summary statistics of taxon pair summary ages
- Uses these summarized ages as secondary calibrations on target tree topology with BLADJ.

Alternative workflow:
- congruifies source chronograms to target tree topology
- matches taxon pair ages that can be used as calibrations to nodes on target tree and stores as a data frame of ages
- Calculates summary statistics of congruified_and_mrca'd ages
- Uses age summary statistics per node as secondary calibrations on target tree topology with BLADJ (and others).

summary_matrix_to_phylo to get_summary_chronograms


### Plot matched calibrations from source chronograms.

Choose a color scheme for the studies:

```{r}
# Get references in matched calibrations:
studies <- as.character(unique(calib_summ$in_phy$reference))


# order references alphabetically:
studies <- studies[order(studies)]
studies
```

Visualize plotting colors:

```{r}
#create a vector of colors with study names:
number_of_colors <- length(studies) * 10


xx <- viridis::turbo(n = number_of_colors)


plot(-1:number_of_colors + 1, -1:number_of_colors + 1, type = "n")
text(1:number_of_colors, 1:number_of_colors, xx, col = xx)

# xx <- viridis::turbo(n = 20)
# plot(-2:length(xx)+1, -2:length(xx)+1, type = "n")
# text(1:(length(xx)), 1:length(xx), paste(1:length(xx),"=",xx), col = xx)

```

Choose plotting colors:

```{r}
#create a vector of colors with study names:
# color_studies <- xx[c(12,
#                       3,4,
#                       15,
#                       18,17,
#                       20,
#                       8,9)]
# 

length(xx) / length(unique(names(phylo_all)))
color_studies <- c("blue", 
                   xx[c(25, 26)],
                   xx[1],
                   "deeppink", 
                   "orchid",
                   xx[c(65, 66)],
                   "pink",
                   xx[c(95, 96)], 
                   xx[c(121, 122)],
                   "skyblue",
                   xx[c(141, 142)], 
                   xx[c(165, 166)], 
                   xx[190])
names(color_studies) <- studies
frin_color_studies <- color_studies
usethis::use_data(frin_color_studies, overwrite = TRUE)
```

# Get summary chronograms from the median matrix

```{r}
all_summary_chronograms <- datelife::summary_matrix_to_phylo_all(median_matrix,
                                      fringillidae_v0.6.2$dquery,
                                      target_tree = fringillidae_v0.6.2$dsumm$phylo_sdm,
                                      total_distance = TRUE)
```

# Congruify calibrations to topology first

```{r}
phylo_median <- fringillidae_v0.6.2$dsumm$phylo_median
phylo_all <- fringillidae_v0.6.2$dsumm$phylo_all
all_calibrations <- datelife::extract_calibrations_phylo(
  input = phylo_all,
  each = FALSE)

congruified_calibs <- datelife::congruify_calibrations(phy = phylo_median,
                                 chronograms = phylo_all,
                                 calibrations = all_calibrations)

ls(congruified_calibs)

ls(congruified_calibs$matched_calibrations)
head(congruified_calibs$matched_calibrations)
```

# Then match?




