---
title: "figure of small example"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{figure2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
writing_dir <- "../figures/figure-small-example/"
```
## Getting source chronograms

```{r}
# library(datelifeMS1)
devtools::load_all("../../datelifeplot")
devtools::load_all("../../datelife")
load("../data/frin_sample_dsumm.rda")
frin_sample_dsumm$phylo_median
spp <- frin_sample_dsumm$phylo_median$tip.label[1:6]
spp_write <- sub("_", " ", spp)
spp_write <- spp_write[order(spp_write)]
write(spp_write, file = "../figures/fringillidae_species_sample.txt")
spp_dq <- datelife::make_datelife_query(spp)
spp_dr <- datelife::get_datelife_result(spp_dq)
```

```{r}
spp_summ <- datelife:::summary.datelifeResult(datelife_query = spp_dq, object = spp_dr)
```


```{r}
phylo_median <- spp_summ$phylo_median
phylo_all <- spp_summ$phylo_all
# Reorder chronogram list alphabetically:
alphabetic_order <- order(names(phylo_all))
phylo_all <- phylo_all[alphabetic_order]
names(phylo_all)
```

We will rename source chronograms to have a shorter study name:

```{r}

names(phylo_all) <- c("Barker et al. 2013",
                      "Barker et al. 2015 - chronogram 1",
                      "Barker et al. 2015 - chronogram 2",
                      "Burns et al. 2014",
                      "Hedges et al. 2015 - chronogram 1",
                      "Hedges et al. 2015 - chronogram 2",
                      "Hooper et al. 2017",
                      "Jetz et al. 2012 - chronogram 1",
                      "Jetz et al. 2012 - chronogram 2")
```

### Plotting chronograms

Plot one chronogram as test.

Create settings and try classic `ape::plot.phylo`:

```{r}
plot_max_time <- 45
chronogram <- phylo_all[[1]]
chronogram$root.edge <- plot_max_time - max(ape::branching.times(chronogram))

ape::plot.phylo(chronogram,
                cex = 1, #edge.width = 2,
                label.offset = 0.5,
                x.lim = c(0, plot_max_time),
                root.edge = TRUE,
                plot = TRUE, edge.width = 2)
ape::axisPhylo(line = 1)
```

Plot with `datelifeplot`:
```{r}
datelifeplot::plot_phylo(chronogram,
                         title = names(phylo_all)[1],
                         plot_type = "phyloch",
                         time_depth = plot_max_time,
                         write = "pdf",
                         file_name = "../figures/figure-small-example/test.pdf",
                         cex_tiplabels = 2,
                         cex_axislabel = 1.5,
                         cex_title = 2,
                         pos_title = 2.2,
                         pos_axis = 2)
```
[!Source chronogram test]("../figures/figure-small-example/test.pdf")


### Plotting ALL source chronograms

We know that chronogram 9 has the oldest ages.
Get the maximum age:
```{r}
max(ape::branching.times(phylo_all[[9]]))  # 44.29586
plot_max_time <- 45
```
Another way of getting oldest age of all trees in a `multiPhylo ` object:

```{r}
max(unlist(lapply(phylo_all, ape::branching.times)))
```

Plot as pdf:

```{r}
datelifeplot::plot_phylo_all(chronograms = phylo_all,
                             folder_name = "../figures/figure-small-example/source_phylo_all",
                             plot_height = 440,
                             plot_width = 600,
                             mai1 = 1,
                             mai2 = 0,
                             mai3 = 1,
                             mai4 = 3.72,
                             omi1 = 1,
                             omi2 = 0,
                             omi3 = 0,
                             omi4 = 0,
                             write = "pdf",
                             plot_type = "phyloch",
                             max_depth = plot_max_time,
                             edge.width = 2,
                             cex_tiplabels = 2,
                             cex_axislabel = 1.5,
                             cex_axis = 2,
                             cex_title = 2.5,
                             pos_title = -2.8,
                             pos_axis = 2,
                             axis_label = "")
```


I combined source chronogram plots with ppt (files in `figures/figure-small-example/source_phylo_all*.pptx`), because combining with `magick` was not
doing what I needed. This is the code I tried:

```
for (i in seq(phylo_all)) {
  ii <- magick::image_read(path = paste0("phylo_all/phylo_all_", i, ".pdf"))
  assign(x = paste0("phy", i), value = ii)
}
img <- c(phy1, phy2, phy3, phy4, phy5, phy6, phy7, phy8, phy9)
img <- magick::image_scale(img, "300x")
magick::image_info(img)

pdf("stacked.pdf")
magick::image_append(magick::image_scale(img, "200x"))
dev.off()
```

## Get calibrations (uncongruified are the same as congruified)

All calibrations, even those that are not congruent:

```{r}
all_calibrations <- datelife::extract_calibrations_phylo(
  input = phylo_all,
  each = FALSE)
class(all_calibrations)
ls(all_calibrations)
nrow(all_calibrations)
```

Find mrca of taxon pairs in `all_calibrations` on a given tree topology (`phylo_median` in this case):

```{r}
matched <- datelife::match_all_calibrations(
  phy = phylo_median, 
  calibrations = all_calibrations)

class(matched)
ls(matched)
nrow(matched$matched_calibrations)
```

Summarize matched calibrations

```{r}
calib_summ <- datelife:::summary.matchedCalibrations(
  matched$matched_calibrations)
class(calib_summ)
ls(calib_summ)
nrow(calib_summ$in_phy)
```
### Table of matched calibrations

Make it pretty.

#### Keep columns of interest:

```{r}
colnames(calib_summ$in_phy)
pretty_calibs <- calib_summ$in_phy[,c("taxonA", "taxonB", "MaxAge", "mrca_node_number", "reference")]
nrow(pretty_calibs)
```
#### Eliminate duplicated taxon pairs:

```{r}
## Find the duplicated pairs within studies:
## Note: take into account age too, they might com from different studies.
taxon_pairs_ref <- c()
for (i in seq(nrow(pretty_calibs))) {
  tax_pair_ref <- as.character(pretty_calibs[i, c("taxonA", "taxonB", "reference","MaxAge")])
  tax_pair_ref <- tax_pair_ref[order(tax_pair_ref)]
  tax_pair_ref <- paste(tax_pair_ref, collapse = "")
  taxon_pairs_ref <- c(taxon_pairs_ref, tax_pair_ref)
}
pretty_calibs[8:9,]
sum(duplicated(taxon_pairs_ref))
```

```{r}
## Eliminate duplicated rows if any:
if (any(duplicated(taxon_pairs_ref))) {
  pretty_calibs <- pretty_calibs[!duplicated(taxon_pairs_ref),]
}
```

```{r}
# Reorder ROWS by node number, reference, age and taxon names
pretty_calibs <- pretty_calibs[order(pretty_calibs$mrca_node_number, pretty_calibs$MaxAge, pretty_calibs$reference, pretty_calibs$taxonA, pretty_calibs$taxonB),]

# Eliminate rownames
rownames(pretty_calibs) <- NULL

# Reorder COLUMNS
pretty_calibs <- pretty_calibs[, c("mrca_node_number", "taxonA", "taxonB", "MaxAge", "reference")]

# Modify column names to be informative
colnames(pretty_calibs) <- c("Node Name", "taxon A", "taxon B", "Node Age", "Study chronogram")

# Change node numbers by node names
pretty_calibs$`Node Name` <- paste0("n", pretty_calibs$`Node Name`-6)

# substitute underscore by space:
pretty_calibs$`taxon A` <- sub("_", " ", pretty_calibs$`taxon A`)
pretty_calibs$`taxon B` <- sub("_", " ", pretty_calibs$`taxon B`)

# order taxa alphabetically by column
for (i in seq(nrow(pretty_calibs))) {
  original <- as.character(pretty_calibs[i, c("taxon A", "taxon B")])
  reordered <- original[order(original)]
  pretty_calibs[i, c("taxon A", "taxon B")] <- reordered
}

nrow(pretty_calibs)
```

Print table as pdf and png
Trick to print in multiple pages from https://stackoverflow.com/questions/15937131/print-to-pdf-file-using-grid-table-in-r-too-many-rows-to-fit-on-one-page

```{r}
## set the theme to have italics in species names:
number_of_rows <- nrow(pretty_calibs)
tt1 <- gridExtra::ttheme_default(core=list(fg_params=list(fontface=c(rep("plain", number_of_rows), rep("italic", number_of_rows*2), rep("plain", number_of_rows*2)))))

# Export PDF
png(paste0("../tables/table-fringillidae-small-example.png"), 
    width = 10, 
    height = 8, 
    units = "in",
    res = 300)
# pdf(paste0("../tables/table-fringillidae-small-example.pdf"), 
#      width = 10, 
#      height = 8)
gridExtra::grid.table(pretty_calibs, theme = tt1) # if you do not want row numbers set rows = NULL
dev.off()
```

Save table as TEX, but it is not formatted pretty:

```{r}
df <- knitr::kable(pretty_calibs, 
                   format = 'latex',
                   row.names = TRUE,
                   align = "c",
                   caption = "Ages of congruified nodes.",
                   label = "table1")
writeLines(df, '../manuscript/table-fringillidae-small-example.tex')
```


Write table as CSV:
```{r}
write.csv(pretty_calibs, paste0("../tables/table-fringillidae-small-example.csv"))
```

### Congruified

Congruify source chronograms with a chosen tree topology:

```{r}
sc1 <- datelife::congruify_and_mrca.phylo(phy = phylo_median,
                             source_chronogram = phylo_all[[1]],
                             study = names(phylo_all)[1])
class(sc1) <- "data.frame"
sc1

congruified_calibs <- congruify_and_mrca.multiPhylo(phy = phylo_median,
                              source_chronograms = phylo_all)
nrow(congruified_calibs)
```

Table of summary statistics of congruified ages. This will be Table number 2 in paper, methods from https://statisticsglobe.com/summary-statistics-by-group-in-r

```{r}
colnames(pretty_calibs)
only_calibs <- pretty_calibs[,c("Node Name", "Node Age")]
colnames(only_calibs) <- c("node", "age")
# only_calibs$`Node Name` <- factor(only_calibs$`Node Name`)
class(only_calibs$`node`)
summary(only_calibs)
```

Creating summary table with a loop to add variance and sd:

```{r}
min_ages <- q1 <- median_ages <- mean_ages <- q3 <- max_ages <- sd_ages <- var_ages <- c()
for (node in unique(only_calibs$node)) {
  rowsies <- only_calibs$node %in% node
  min_ages <- c(min_ages, min(only_calibs$age[rowsies]))
  q1 <- c(q1, quantile(only_calibs$age[rowsies], 0.25))
  median_ages <- c(median_ages, median(only_calibs$age[rowsies]))
  mean_ages <- c(mean_ages, mean(only_calibs$age[rowsies]))
  q3 <- c(q3, quantile(only_calibs$age[rowsies], 0.75))
  max_ages <- c(max_ages, max(only_calibs$age[rowsies]))
  sd_ages <- c(sd_ages, sd(only_calibs$age[rowsies]))
  var_ages <- c(var_ages, var(only_calibs$age[rowsies]))
}
summary_table <- data.frame(unique(only_calibs$node),
                            min_ages,
                            q1,
                            median_ages,
                            mean_ages,
                            q3,
                            max_ages,
                            var_ages,
                            sd_ages)
colnames(summary_table) <- c("Node Name",
                             "Min Age",
                             "Q1",
                             "Median Age",
                             "Mean Age",
                             "Q3",
                             "Max Age",
                             "Variance",
                             "SD")
```


Print the summary table as pdf:

```{r}
# Export PDF
png(paste0("../tables/table-fringillidae-small-example-summary.png"), 
    width = 9.2, 
    height = 2, 
    units = "in",
    res = 300)
# pdf(paste0("../tables/table-fringillidae-small-example.pdf"), 
#      width = 10, 
#      height = 8)
gridExtra::grid.table(summary_table, 
                      theme = tt1,
                      rows = NULL) # I do not want row numbers
dev.off()

```

Write summary table as CSV:
```{r}
write.csv(summary_table, paste0("../tables/table-fringillidae-small-example-summary.csv"))
```

### Plot matched calibrations from source chronograms.

Choose a color scheme for the studies:

```{r}
# Get references in matched calibrations:
studies <- as.character(unique(calib_summ$in_phy$reference))

# order references alphabetically:
studies <- studies[order(studies)]
studies

```

Visualize plotting colors:

```{r}
#create a vector of colors with study names:
xx <- viridis::turbo(n = length(studies))

plot(-1:10, -1:10, type = "n")
text(1:9, 1:9, xx, col = xx)

xx <- viridis::turbo(n = 20)
plot(-2:length(xx)+1, -2:length(xx)+1, type = "n")
text(1:(length(xx)), 1:length(xx), paste(1:length(xx),"=",xx), col = xx)

```

Choose plotting colors:

```{r}
#create a vector of colors with study names:
color_studies <- xx[c(12,
                      3,4,
                      15,
                      18,17,
                      20,
                      8,9)]
names(color_studies) <- studies
usethis::use_data(color_studies)
```

Now, we can plot:

```{r}
svg(filename = "../figures/figure-small-example/node_ages_matched.svg",
    height = 10,
    width = 7)
datelifeplot::plot_node_ages(phylo_median,
               title = "Median Summary Chronogram",
               plot_type = "phyloch",
               time_depth = plot_max_time,
               calibration_summary = calib_summ,
               pch = 19,
               color_pch = color_studies,
               cex_pch = 2,
               lwd_bars = 15,
               cex_legend = 0.7,
               x_legend = 2,
               y_legend = 6,
               mai1 = 1, 
               mai2 = 0,
               mai3 = 0.5,
               mai4 = 3.73,
               omi1 = 2,
               omi2 = 0,
               omi3 = 0.5,
               omi4 = 0,
               edge.width = 2,
               cex_tiplabels = 2,
               cex_axislabel = 1.5,
               cex_axis = 2,
               cex_title = 2,
               pos_title = -1,
               pos_axis = 2
               )
dev.off()
```


## Get pairwise median age points

```{r}
names(spp_dr)
spp_dr[[1]]

best_grove <- datelife::get_best_grove(spp_dr)

# Take a list of chronograms as tables and summarize age data per taxon pair into a single matrix:
median_matrix <- datelife::datelife_result_median_matrix(best_grove$best_grove)
colnames(median_matrix)

median_matrix_list <- list(summary_matrix = median_matrix)

median_matrix_table <- datelife::matrices_to_table(median_matrix_list)

# match taxon pairs to nodes on a tree topology
matched_median_ages <- datelife::match_all_calibrations(phy = spp_summ$phylo_median, calibrations = median_matrix_table)

mm_summ <- datelife:::summary.matchedCalibrations(matched_median_ages$matched_calibrations)

mm_summ$in_phy
```

### Table of median of node ages

Make it pretty first:

```{r}
colnames(mm_summ$in_phy)
pretty_median <- mm_summ$in_phy[,c("taxonA", "taxonB", "nodeAge", "mrca_node_number")]
pretty_median
# Because patristic matrices are symmetrical, going from a patristic matrix to a table creates a duplicate for all taxon pairs
# Eliminate duplicated taxon pairs:
## Find the duplicated pairs
taxon_pairs <- c()
for (i in seq(nrow(pretty_median))) {
  taxon_pair <- as.character(pretty_median[i, c("taxonA", "taxonB")])
  taxon_pair <- taxon_pair[order(taxon_pair)]
  taxon_pair <- paste(taxon_pair, collapse = "")
  taxon_pairs <- c(taxon_pairs, taxon_pair)
}
# Duplicated pairs should sum to 30/2 = 15
sum(duplicated(taxon_pairs))
## Eliminate them:
pretty_median <- pretty_median[!duplicated(taxon_pairs),]

# Reorder ROWS by node number and taxon names
pretty_median <- pretty_median[order(pretty_median$mrca_node_number, pretty_median$taxonA, pretty_median$taxonB),]
# Eliminate rownames
rownames(pretty_median) <- NULL
# Reorder COLUMNS
pretty_median <- pretty_median[, c("mrca_node_number", "taxonA", "taxonB", "nodeAge")]
# Modify column names to be informative
colnames(pretty_median) <- c("Node Name", "taxon A", "taxon B", "Pairwise Median Age")
# Change node numbers to node names
pretty_median$`Node Name` <- paste0("n", pretty_median$`Node Name`-6)

# substitute underscore by space:
pretty_median$`taxon A` <- sub("_", " ", pretty_median$`taxon A`)
pretty_median$`taxon B` <- sub("_", " ", pretty_median$`taxon B`)

# Add a column of node median age
## Calculate the node median age per node:
node_median <- c()
for (i in unique(pretty_median$`Node Name`)) {
  print(i)
  x <- median(pretty_median$`Pairwise Median Age`[pretty_median$`Node Name` == i])
  print(x)
  node_median <- c(node_median, rep(x, sum(pretty_median$`Node Name` == i)))
}
## Create a column
pretty_median$`Node Median Age` <- node_median
# Look atthe final result!
pretty_median
```

Print table as pdf
```{r}
## set the theme to have italics in species names:
tt1 <- gridExtra::ttheme_default(core=list(fg_params=list(fontface=c(rep("plain", 15), rep("italic", 30), rep("plain", 30)))))

# Export PDF
pdf("../figures/figure-small-example/median_table.pdf", width = 10)
gridExtra::grid.table(pretty_median, theme = tt1) # if you do not want row numbers set rows = NULL
dev.off()
```

Simplifying pretty_median table.
Let's explore the grid and gridExtra package (code from https://cyancey95.github.io/CompBiol_2018/gridExtraPresentation.html):

```{r}
g <- lapply(1:11, 
            function(i) { 
              grid::grobTree(grid::rectGrob(gp = grid::gpar(fill = i, alpha = .5)),
                 grid::textGrob(i))
            }
    )
pattern <- rbind(c(1,1,2,2,3),
                 c(1,1,2,2,4),
                 c(1,1,5,5,6),
                 c(7,8,9,10,11))
gridExtra::grid.arrange(grobs = g, layout_matrix = pattern)
```

```{r}
# if you do not want row numbers set rows = NULL

# Detecting node names that are duplicated 
duplos <- duplicated(pretty_median$`Node Name`)
# Making node names pretty:
pretty_median$`Node Name`[!duplos]
pretty_median$`Node Name`[duplos] <- ""
# putting "n1" node name in the middle
pretty_median$`Node Name`[1] <- ""
pretty_median$`Node Name`[5] <- "n1"

# Simplifying node median ages:
pretty_median$`Node Median Age`[duplos] <- ""
pretty_median$`Node Median Age`[!duplos] <- node_median[!duplos]
pretty_median$`Node Median Age`[1] <- ""
pretty_median$`Node Median Age`[5] <- node_median[!duplos][1]
```

Plotting:

Setting up a color theme

```{r}
tt1 <- gridExtra::ttheme_default(core=list(
  fg_params=list(fontface = c(rep("plain", 15), 
                              rep("italic", 30), 
                              rep("plain", 30))),
  bg_params = list(fill = c(rep(c("grey95"), length.out=9), 
                            rep(c("grey90"), length.out=2),
                            "grey95",
                            rep(c("grey90"), length.out=2),
                            "grey95",
                            rep(c("grey95", "grey90"), length.out=15),
                            rep(c("grey95", "grey90"), length.out=15),
                            rep(c("grey95", "grey90"), length.out=15),
                            rep(c("grey95"), length.out=9), 
                            rep(c("grey90"), length.out=2),
                            "grey95",
                            rep(c("grey90"), length.out=2),
                            "grey95"))))
```

Writing the table out:

```{r}
pdf("../figures/figure-small-example/median_table_simple.pdf", width = 10)       # Export PDF
gridExtra::grid.table(pretty_median, theme = tt1) # if you do not want row numbers set rows = NULL
dev.off()
```
             
### Plotting source ages AND summary ages on the same dated tree

Set colors and data:

```{r}
color_median <- "purple" # mustard yellow "#7A0403FF"
mm_summ$in_phy$reference <- rep("Pairwsie median of node ages", length(mm_summ$in_phy$reference))
names(color_median) <- mm_summ$in_phy$reference[1]

point_colors <- list(a1 = color_studies, a2 = color_median)
all_ages <- list(a1 = calib_summ, a2 = mm_summ)
point_type <- list(a1 = "/", a2 = "•")

leg_color<- c(color_studies, color_median)
# leg_pch <- c(rep("|", length(color_studies)), "•")
leg_pch <- rep(19, length(leg_color))
leg_text <- names(leg_color)
```

Plot:

```{r}
svg(filename = "../figures/figure-small-example/median_and_calibration_ages.svg",
    height = 8.5 ,
    width = 10.5)
datelifeplot::plot_node_ages2(phylo_median,
               title = c("Median Summary Chronogram"),
               pos_title = 3,
               cex_title = 2,
               plot_type = "phyloch",
               time_depth = plot_max_time,
               pos_axis = 2,
               cex_axis = 1.5,
               edge.width = 3,
               cex_tiplabels = 2,
               matched_ages  = all_ages,
               pch = point_type,
               pch_color = point_colors,
               pch_cex = 3,
               bars_lwd = 15,
               legend_cex = 0.9,
               legend_x = -5,
               legend_y = 6.2,
               legend_text = leg_text,
               legend_pch = leg_pch,
               legend_color = leg_color,
               mai1 = 1,
               mai2 = 0.8,
               mai4 = 3.6,
               axis_label = ""
               )
dev.off()
```        

Simplifying median ages on tree plot:

```{r}
# Getting the median per node:
median_singletons <- c()
for (i in unique(mm_summ$in_phy$mrca_node_name)) {
  print(i)
  x <- median(unique(mm_summ$in_phy$nodeAge[mm_summ$in_phy$mrca_node_name == i]))
  median_singletons <- c(median_singletons, x)
}

ape::branching.times(phylo_median)

# Detecting node names that are duplicated 
duplos <- duplicated(mm_summ$in_phy$mrca_node_name)
# Simplifying median ages:
mm_summ$in_phy$nodeAge[!duplos] <- median_singletons
# Removing duplicates:
mm_summ$in_phy <- mm_summ$in_phy[!duplos,]
# Columns MinAge and MaxAge need to be equal to nodeAge
mm_summ$in_phy$MinAge <- mm_summ$in_phy$MaxAge <- mm_summ$in_phy$nodeAge

```

Plotting parameters:

```{r}
mm_summ$in_phy$reference <- rep("Used as calibration", length(mm_summ$in_phy$reference))
mm_summ$in_phy$reference[2] <- "Not used"
gplots::col2hex("purple") # "#A020F0"
color_median <- c("#A020F070", "purple") # mustard yellow "#7A0403FF"

names(color_median) <- unique(mm_summ$in_phy$reference)

point_colors <- list(a1 = color_studies, a2 = color_median)
all_ages <- list(a1 = calib_summ, a2 = mm_summ)
point_type <- list(a1 = "/", a2 = c(19, 8, 19, 19, 19))


leg_color<- list(a1 = color_studies, a2 = color_median)
# leg_pch <- c(rep("|", length(color_studies)), "•")
leg_pch <- list(a1 = 19, a2 = c(19, 8))
leg_text <- lapply(leg_color, names)
leg_title <- c("Taxon pair node ages from studies", "Median of node ages")
leg_box <- c(TRUE, TRUE)
```

Plot again:
```{r}
pdf(file = "../figures/figure-small-example/median_and_calibration_ages_simple.pdf",
    height = 8.5,
    width = 10.5)
datelifeplot::plot_node_ages2(phylo_median,
               title = c("Median Summary Chronogram"),
               pos_title = 3,
               cex_title = 2,
               plot_type = "phyloch",
               time_depth = plot_max_time,
               pos_axis = 2,
               cex_axis = 1.5,
               edge.width = 3,
               cex_tiplabels = 2,
               matched_ages  = all_ages,
               pch = point_type,
               pch_color = point_colors,
               pch_cex = 3,
               bars_lwd = 15,
               legend_cex = c(0.9, 0.9),
               legend_x = c(-5, 18),
               legend_y = c(6.2, 6.2),
               legend_text = leg_text,
               legend_pch = leg_pch,
               legend_color = leg_color,
               legend_title = leg_title,
               legend_box = leg_box,
               mai1 = 1,
               mai2 = 0.8,
               mai4 = 3.6,
               axis_label = ""
               )
dev.off()
```        

## The tree topology

Get phylo median tree as a topology (all branches of equal length)

```{r}
# visualize median branching times:
ape::branching.times(phylo_median)
# force equal branch lengths, with root age = 1
pp <- ape::compute.brlen(phylo_median, power = 1)
# visualize the new branch lengths:
ape::branching.times(pp)
pp$edge.length
plot(pp)
ape::axisPhylo(side = 1)
# to make branch lengths proportional to a certain age we have o multiply them by that age, in this case it's the total plot_max_time
pp$edge.length <- pp$edge.length*plot_max_time 
```


Plot median tree topology with node labels and no time axis

```{r}
svg(filename = "../figures/figure-small-example/median_tree_topology.svg",
    height = 5,
    width = 6,
    bg = "transparent")
datelifeplot::plot_chronogram(pp,
                         title = c("Tree topology"),
                         time_axis = FALSE,
                         axis_label = "",
                         edge.width = 3,
                         cex_tiplabels = 2,
                         cex_axislabel = 1.5,
                         cex_axis = 2,
                         cex_title = 2,
                         pos_title = 3,
                         mai4 = 3.6)
ape::nodelabels(text = pp$node.label, frame = "circle", bg = "white", cex = 1.5)
dev.off()
```


