---
title: "figure-cross-validation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{figure-cross-validation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(datelifeMS1)
devtools::load_all("../../datelifeplot")
devtools::load_all("../../datelife")
writing_dir <- "../figures/figure-cross-validation/"

# Based on function use_othercals3 from https://github.com/LunaSare/datelife_examples/blob/master/R/functions_data.R#L36
use_other_calibrations <- function(trees, other_calibrations, ...){
    res <- lapply(seq(trees), function(i){
      phy <- trees[[i]]
      phy$edge.length <- NULL
      print(i)
        xx <- suppressMessages(suppressWarnings(datelife::use_calibrations_bladj(phy,
            calibrations = other_calibrations[[i]], ...)))
        return(xx)
    })
    class(res) <- "multiPhylo"
    names(res) <- names(trees)
    res
}

# Function based on https://github.com/LunaSare/datelife_examples/blob/master/R/functions_data.R#L3
# get calibrations form all other chronograms not belonging to the same study
extract_other_calibrations <- function(phyloall){
    # dd <- duplicated(names(phyloall))
    res <- lapply(seq(phyloall), function(i){
        dd <- names(phyloall) %in% names(phyloall)[i]
        datelife::extract_calibrations_phylo(phyloall[!dd], each = FALSE)
    })
    names(res) <- names(phyloall)
    res
}
```

## Cross validation of Fringillidae source chronograms.

We are using the object `frin_dsumm` and getting all source chronograms, and reordering the chronogram list alphabetically:

```{r}
phylo_all <- frin_dsumm$phylo_all
names(phylo_all)
alphabetic_order <- order(names(phylo_all))
phylo_all <- phylo_all[alphabetic_order]
```

Next, get calibrations for all other chronograms:

```{r}
frin_other_calibs <- extract_other_calibrations(phylo_all)
class(frin_other_calibs)
length(frin_other_calibs)
names(frin_other_calibs) <- names(phylo_all)
```

Date the topology based on other calibrations:

```{r}
cross_validation <- use_other_calibrations(trees = phylo_all, other_calibrations = frin_other_calibs)

# target = the tree topology that is going to be dated
# reference = the chronogram with dates to time-scale `target`
cc <- geiger::congruify.phylo(reference = phylo_all[[1]],
                        target = phylo_median,
                        scale = NA,
                        ncores = 1)
cc <- geiger::congruify.phylo(reference = phylo_all[[1]],
                        target = phylo_median,
                        scale = NA,
                        ncores = 1)$calibrations

cc <- geiger::congruify.phylo(reference = c(list(phylo_median), phylo_all),
                        target = phylo_median,
                        scale = NA,
                        ncores = 1)
names(cc) <- c("phylo_median", names(phylo_all))
cc[is.na(cc)]
unname(which(is.na(cc)))
```

Congruification of all chronograms at the same time returns `NA` for some chronograms (`r paste0(names(cc)[is.na(cc)], collapse, = ", ")`).

So, I am testing congruification one on one:

```{r}
# Function
# Congruify a source chronogram with the target and itself
run_congruification <- function(phylo_median, phylo_all, index) {
    cc <- geiger::congruify.phylo(reference = list(median = phylo_median, 
                                                    self = phylo_all[[index]]),
                            target = phylo_all[[index]],
                            scale = NA,
                            ncores = 1)
    names(cc) <- c("median", "self")
    cc
}
```


Test: Check that hashes from congruification do correspond to the same node in target and reference

```{r}
names(phylo_all)[5]

cc5 <- run_congruification(phylo_median, phylo_all, index = 5)

cc5$median$reference$node.label
cc5$self$reference$node.label
cc5$self$target$hash
cc5$median$target$hash


pdf(file = paste0(writing_dir, "cc5_self_reference.pdf"))
ape::plot.phylo(cc5$self$reference)
ape::nodelabels(cc5$self$reference$node.label, adj = 0)
dev.off()

pdf(file = paste0(writing_dir, "cc5_median_reference.pdf"),
    height = 25)
ape::plot.phylo(cc5$median$reference, cex = 0.5)
ape::nodelabels(cc5$median$reference$node.label, adj = 0, cex = 0.5)
dev.off()

# Function:
# Plot congruified trees with hash labels as node labels
# @description plots an output of run_congruification
plot_congruification <- function(cc, 
                                 id = "",
                                 writing_dir, 
                                 height_self = 7, 
                                 height_target = 25) {
  phylo <- cc$self$reference
  pho <- phylo_height_omi(phy = phylo)
  pdf(file = paste0(writing_dir, 
                    deparse(substitute(cc)),
                    id,
                    "_congruification_reference2self.pdf"),
      height = pho$height/72)
  ape::plot.phylo(phylo)
  ape::nodelabels(phylo$node.label, adj = 0)
  dev.off()

  phylo <- cc$median$reference
  pho <- phylo_height_omi(phy = phylo)
  pdf(file = paste0(writing_dir,
                    deparse(substitute(cc)), 
                    "_congruification_reference2target.pdf"),
      height = pho$height/72)
  ape::plot.phylo(phylo, cex = 0.5)
  ape::nodelabels(phylo$node.label, adj = 0, cex = 0.5)
  dev.off() 
}

# run congruification and plot for all source chronograms

for (i in seq(phylo_all)) {
  print(names(phylo_all)[i])
  cc <- run_congruification(phylo_median, phylo_all, index = i)
  plot_congruification(cc, id = i, writing_dir)
}





# get congruification hashes:
nn <- lapply(cc, function(x) {
                tryCatch(x$reference$node.label, error = function(e) NA)
})
# Test that hash names are congruent in target and reference
for (i in seq(nn)) {
  xx <- match(nn[[1]], nn[[i]])
  print(i)
  print(xx)
  yy <- nn[[i]][xx]
  print(yy)
}
```
Conclusion: Hash labels are different between reference trees, even if they correspond to the same node in target.

However, hashed nodes do correspond to the same clade between reference chronograms.
Then, we can use taxon pairs from hashed nodes to identify corresponding node labels on target.

What part from congruification output do we need?

```{r}
ls(cc19)
cc19$self$calibrations
cc19$median$calibrations
```
TODO:

test that MRCA from congruification corresponds to the same node obtained with phytools::find.mrca()


