---
title: "figure-fringillidae"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{figure-fringillidae}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
devtools::load_all("../../datelifeplot")
devtools::load_all("../../datelife")
writing_dir <- "../figures/figure-fringillidae/"
```

## Get a `datelifeQuery` object

```{r, eval = TRUE}
dquery <- datelife::make_datelife_query(input = "fringillidae",
                                             use_tnrs = TRUE,
                                             get_spp_from_taxon = TRUE,
                                             taxonomic_source = "ncbi")
# dquery <- fringillidae_v0.6.2$dquery
```

## Get a `datelifeResult` object

```{r, eval = TRUE}
dres = datelife::get_datelife_result(input = dquery, cache = opentree_chronograms)

# dres <- fringillidae_v0.6.2$dres
```

## Get a `datelifeSummary` object
```{r}
# you can add a tree topology to summarize if you do not want the Open Tree synthetic tree topology as default
# dquery$phy <- fringillidae_v0.6.2$dsumm$phylo_median

dsumm <- datelife:::summary.datelifeResult(datelife_query = dquery, object = dres)
```

## Save data objects 

```{r}
opentree_chronograms$version
```

```{r}
data_filename <- paste0("fringillidae_v", opentree_chronograms$version)
```

```{r}
assign(data_filename, value = list(dquery = dquery,
                                   dres = dres,
                                   dsumm = dsumm))
```

```{r}
ls(fringillidae_v0.6.2)
```

```{r}
usethis::use_data(fringillidae_v0.6.2, 
                  overwrite = TRUE)
```

## Load data objects 

```{r}
load(file = "../data/fringillidae_v0.6.2.rda")
```

Make an object with all source chronograms, and reorder chronogram list alphabetically:

```{r}
phylo_all <- fringillidae_v0.6.2$dsumm$phylo_all
alphabetic_order <- order(names(phylo_all))
phylo_all <- phylo_all[alphabetic_order]
names(phylo_all)
```
We will rename source chronograms to have a shorter study name:

```{r}
names(phylo_all) <- c("Barker et al. 2013",  # 1
                      "Barker et al. 2015 - chronogram 1",  # 2
                      "Barker et al. 2015 - chronogram 2",  # 3
                      "Burns et al. 2014",  # 4
                      "Claramunt et al. 2015",  # 5
                      "Gibb et al. 2015", # 6
                      "Hedges et al. 2015 - chronogram 1",  # 7
                      "Hedges et al. 2015 - chronogram 2",  # 8
                      "Hooper et al. 2017",  # 9
                      "Jetz et al. 2012 - chronogram 1",  # 10
                      "Jetz et al. 2012 - chronogram 2",  # 11
                      "Kimball et al. 2019 - chronogram 1", # 12
                      "Kimball et al. 2019 - chronogram 2", # 13
                      "Oliveros et al. 2019", # 14
                      "Price et al. 2014 - chronogram 1", # 15
                      "Price et al. 2014 - chronogram 2", # 16
                      "Roquet et al. 2014 - chronogram 1", # 17
                      "Roquet et al. 2014 - chronogram 2", # 18 
                      "Uyeda et al 2017") # 19
```

### Plotting source chronograms

Getting oldest age among all trees in a `multiPhylo` object:

```{r}
max(unlist(lapply(phylo_all, ape::branching.times)))

plot_max_time <- 45
```

Plot as pdf:

```{r}
datelifeplot::plot_phylo_all(chronograms = phylo_all,
                             folder_name = paste0(writing_dir, "/source_phylo_all"),
                             plot_height = 890,
                             plot_width = 600,
                             mai1 = 1,
                             mai2 = 0,
                             mai3 = 1,
                             mai4 = 5,
                             omi1 = 1,
                             omi2 = 0,
                             omi3 = 0,
                             omi4 = 0,
                             write = "pdf",
                             plot_type = "phyloch",
                             max_depth = plot_max_time,
                             edge.width = 2,
                             cex_tiplabels = 2,
                             cex_axislabel = 1.5,
                             cex_axis = 2,
                             cex_title = 2.5,
                             pos_title = -2.8,
                             pos_axis = 2,
                             axis_label = "")
```


## Get calibrations

### Uncongruified

Get calibrations for the topology in `dsumm$phylo_median`

First, get all calibrations per chronogram against themselves:
```{r}
phylo_median <- fringillidae_v0.6.2$dsumm$phylo_median
all_calibrations <- datelife::extract_calibrations_phylo(
  input = phylo_all,
  each = FALSE)
nrow(all_calibrations) # 1206
```

Then match node pairs to topology.
This will use all node ages as calibrations regardless if the node is monophyletic in the target tree (topology to add dates to) and source tree (chronogram from original study).

```{r}
matched <- datelife::match_all_calibrations(
  phy = phylo_median, 
  calibrations = all_calibrations)
ls(matched)
nrow(matched$matched_calibrations) # 1206
```

```{r}
calib_summ <- datelife:::summary.matchedCalibrations(
  matched$matched_calibrations)
nrow(calib_summ$in_phy) # 1206
```

#### Table of matched calibrations

Make it pretty.

```{r}
source("../R/make_pretty_tables.R")
my_calibrations <- calib_summ$in_phy
number_of_tips <- min(my_calibrations$mrca_node_number)-1
my_calibrations$mrca_node_name <- paste0("n", (my_calibrations$mrca_node_number - number_of_tips))
head(my_calibrations$mrca_node_name)

pretty_calibs <- make_pretty_table(some_table = my_calibrations)

my_calibrations[35:36,]
nrow(pretty_calibs)
head(pretty_calibs)
```

Print table as pdf.
Trick to print in multiple pages from https://stackoverflow.com/questions/15937131/print-to-pdf-file-using-grid-table-in-r-too-many-rows-to-fit-on-one-page

```{r}
## set the theme to have italics in species names:
number_of_rows <- nrow(pretty_calibs)
tt1 <- gridExtra::ttheme_default(core=list(fg_params=list(fontface=c(rep("plain", number_of_rows), rep("italic", number_of_rows*2), rep("plain", number_of_rows*2)))))
```

# Export PDF

```{r}
pdf("../tables/table-fringillidae-all.pdf", 
    width = 12.5,
    height = 336 )
gridExtra::grid.table(pretty_calibs, theme = tt1) # if you do not want row numbers set rows = NULL
dev.off()
```

### Congruified

Congruify source chronograms with a chosen tree topology:

```{r}
sc1 <- datelife::congruify_and_mrca.phylo(phy = phylo_median,
                             source_chronogram = phylo_all[[1]],
                             study = names(phylo_all)[1])
class(sc1) <- "data.frame"
sc1

congruified_calibs <- congruify_and_mrca.multiPhylo(phy = phylo_median,
                              source_chronograms = phylo_all)
nrow(congruified_calibs)
```

```{r}
# source("../R/make_pretty_tables.R")
my_calibrations <- congruified_calibs
number_of_tips <- min(congruified_calibs$mrca_node_number)-1
my_calibrations$mrca_node_name <- paste0("n", (my_calibrations$mrca_node_number - number_of_tips))
head(my_calibrations)
nrow(my_calibrations)
some_table <- as.data.frame.data.frame(my_calibrations)
class(some_table)
pretty_calibs <- make_pretty_table(some_table = my_calibrations,
                                   duplicated_criteria = c("taxonA", "taxonB", "study","MaxAge")
)

my_calibrations[35:36,]
nrow(pretty_calibs)
head(pretty_calibs)
```

Table of summary statistics of congruified ages. This will be Table supplementary table in paper, methods from https://statisticsglobe.com/summary-statistics-by-group-in-r

```{r}
colnames(pretty_calibs)
only_calibs <- pretty_calibs[,c("Node Name", "Node Age")]
colnames(only_calibs) <- c("node", "age")
# only_calibs$`Node Name` <- factor(only_calibs$`Node Name`)
class(only_calibs$`node`)
summary(only_calibs)
```


### Plot matched calibrations from source chronograms.

Choose a color scheme for the studies:

```{r}
# Get references in matched calibrations:
studies <- as.character(unique(calib_summ$in_phy$reference))


# order references alphabetically:
studies <- studies[order(studies)]
studies
```

Visualize plotting colors:

```{r}
#create a vector of colors with study names:
number_of_colors <- length(studies) * 10


xx <- viridis::turbo(n = number_of_colors)


plot(-1:number_of_colors + 1, -1:number_of_colors + 1, type = "n")
text(1:number_of_colors, 1:number_of_colors, xx, col = xx)

# xx <- viridis::turbo(n = 20)
# plot(-2:length(xx)+1, -2:length(xx)+1, type = "n")
# text(1:(length(xx)), 1:length(xx), paste(1:length(xx),"=",xx), col = xx)

```

Choose plotting colors:

```{r}
#create a vector of colors with study names:
# color_studies <- xx[c(12,
#                       3,4,
#                       15,
#                       18,17,
#                       20,
#                       8,9)]
# 

length(xx) / length(unique(names(phylo_all)))
color_studies <- c("blue", 
                   xx[c(25, 26)],
                   xx[1],
                   "deeppink", 
                   "orchid",
                   xx[c(65, 66)],
                   "pink",
                   xx[c(95, 96)], 
                   xx[c(121, 122)],
                   "skyblue",
                   xx[c(141, 142)], 
                   xx[c(165, 166)], 
                   xx[190])
names(color_studies) <- studies
frin_color_studies <- color_studies
usethis::use_data(frin_color_studies, overwrite = TRUE)
```


Now, we can plot:

```{r}
svg(filename = paste0(writing_dir, "/node_ages_matched.svg"),
    height = 30,
    width =  15)
datelifeplot::plot_node_ages(phylo_median,
               title = "Median Summary Chronogram",
               plot_type = "phyloch",
               time_depth = plot_max_time,
               calibration_summary = calib_summ,
               pch = 19,
               color_pch = color_studies,
               cex_pch = 1,
               lwd_bars = 10,
               cex_legend = 0.7,
               x_legend = 0 ,
               y_legend = 270,
               mai1 = 1, 
               mai2 = 0,
               mai3 = 0.5,
               mai4 = 3.73,
               omi1 = 2,
               omi2 = 1,
               omi3 = 0.5,
               omi4 = 0,
               edge.width = 2,
               cex_tiplabels = 0.5,
               cex_axislabel = 1,
               cex_axis = 1,
               cex_title = 2,
               pos_title = -1,
               pos_axis = 2
               )
dev.off()
```
## Get pairwise median age points

```{r}
best_grove <- datelife::get_best_grove(fringillidae_v0.6.2$dres)

# Take a list of chronograms as tables and summarize age data per taxon pair into a single matrix:
median_matrix <- datelife::datelife_result_median_matrix(best_grove$best_grove)
colnames(median_matrix)

median_matrix_list <- list(summary_matrix = median_matrix)

median_matrix_table <- datelife::matrices_to_table(median_matrix_list)

# removing NAs:

median_matrix_table <- median_matrix_table[!is.na(median_matrix_table$taxonA),]

# match taxon pairs to nodes on a tree topology

matched_median_ages <- datelife::match_all_calibrations(phy = phylo_median, calibrations = median_matrix_table)
ls(matched_median_ages)
nrow(matched_median_ages$matched_calibrations) # 58230
mm_summ <- datelife:::summary.matchedCalibrations(matched_median_ages$matched_calibrations)

nrow(mm_summ$in_phy) #229
head(rownames(mm_summ$in_phy))
```


### Plotting source ages AND summary ages on the same dated tree

Set colors and data:

```{r}
color_median <- "purple"
mm_summ$in_phy$reference <- rep("Pairwise median of node ages", length(mm_summ$in_phy$reference))
names(color_median) <- mm_summ$in_phy$reference[1]

point_colors <- list(a1 = color_studies, a2 = color_median)
all_ages <- list(a1 = calib_summ, a2 = mm_summ)
point_type <- list(a1 = "/", a2 = "•")

leg_color <- c(color_studies, color_median)
leg_pch <- rep(19, length(leg_color))
leg_text <- names(leg_color)
```

Plot:

```{r}
svg(filename = paste0(writing_dir, "/median_and_calibration_ages.svg"),
    height = 30,
    width = 15)
datelifeplot::plot_node_ages2(phylo_median,
               title = c("Median Summary Chronogram"),
               pos_title = 3,
               cex_title = 2,
               plot_type = "phyloch",
               time_depth = plot_max_time,
               pos_axis = 2,
               cex_axis = 1,
               edge.width = 3,
               cex_tiplabels = 0.5,
               matched_ages  = all_ages,
               pch = point_type,
               pch_color = point_colors,
               pch_cex = 2,
               bars_lwd = 10,
               bars_color = "#DCDCDC50",
               legend_cex = 0.8,
               legend_x = -5,
               legend_y = 200,
               legend_text = leg_text,
               legend_pch = leg_pch,
               legend_color = leg_color,
               mai1 = 1,
               mai2 = 0.8,
               mai4 = 3.6,
               axis_label = ""
               )
dev.off()
```        

Simplifying median ages on tree plot:

```{r}
# Getting the median per node:
median_singletons <- c()
for (i in unique(mm_summ$in_phy$mrca_node_name)) {
  x <- median(unique(mm_summ$in_phy$nodeAge[mm_summ$in_phy$mrca_node_name == i]))
  median_singletons <- c(median_singletons, x)
}

# Detecting node names that are duplicated 
duplos <- duplicated(mm_summ$in_phy$mrca_node_name)
# Simplifying median ages:
mm_summ$in_phy$nodeAge[!duplos] <- median_singletons
# Removing duplicates:
mm_summ$in_phy <- mm_summ$in_phy[!duplos,]
# Columns MinAge and MaxAge need to be equal to nodeAge
mm_summ$in_phy$MinAge <- mm_summ$in_phy$MaxAge <- mm_summ$in_phy$nodeAge

```

Plotting parameters:

```{r}
mm_summ$in_phy$reference <- rep("Used as calibration", length(mm_summ$in_phy$reference))
# mm_summ$in_phy$reference[2] <- "Not used"
gplots::col2hex("purple") # "#A020F0"
color_median <- c("#A020F0")  # purple with no transparency

names(color_median) <- unique(mm_summ$in_phy$reference)

point_colors <- list(a1 = color_studies, a2 = color_median)
all_ages <- list(a1 = calib_summ, a2 = mm_summ)
point_type <- list(a1 = 19, a2 = 8)


leg_color<- list(a1 = color_studies, a2 = color_median)
leg_pch <- list(a1 = 19, a2 = 8)
leg_text <- lapply(leg_color, names)
leg_title <- c("Taxon pair node ages from studies", "Median of node ages")
leg_box <- c(TRUE, TRUE)
```

Plot Median summary Chronogram again:
```{r}
phylo_median <- datelife_result_median(fringillidae_v0.6.2$dres,
                                       target_tree = fringillidae_v0.6.2$dsumm$phylo_sdm)

```
```{r}
svg(filename = paste0(writing_dir, "median_and_calibration_ages_simple.svg"),
    height = 11.75*2,
    width = 8.25*2)
ee <- datelifeplot::plot_node_ages2(phylo_median,
               title = NULL,
               pos_title = -5,
               cex_title = 2,
               plot_type = "phyloch",
               time_depth = plot_max_time,
               pos_axis = 2,
               cex_axis = 2,
               edge.width = 2,
               cex_tiplabels = 0.6,
               matched_ages  = all_ages,
               pch = point_type,
               pch_color = point_colors,
               pch_cex = 1,
               bars_lwd = 7,
               # bars_color = "#DCDCDC50", # defaulto to "#80808050"
               legend_cex = c(1.2, 1.2),
               legend_x = c(1, 1),
               legend_y = c(140, 80),
               legend_text = leg_text,
               legend_pch = leg_pch,
               legend_color = leg_color,
               legend_title = leg_title,
               legend_box = leg_box,
               mai1 = 0,
               mai2 = 0,
               mai3 = 0,
               mai4 = 1,
               omi1 = 0,
               omi2 = 0,
               omi3 = 0,
               axis_label = ""
               )
dev.off()
```      
# Get all types of summary chronograms from the median matrix

```{r}
all_summary_chronograms <- datelife::summary_matrix_to_phylo_all(median_matrix,
                                      fringillidae_v0.6.2$dquery,
                                      target_tree = fringillidae_v0.6.2$dsumm$phylo_sdm,
                                      total_distance = TRUE)
```


Now plot:

```{r}

for (i in names(all_summary_chronograms)) {
 svg(filename = paste0(writing_dir, toupper(i), "_ages.svg"),
    height = 30,
    width =  15)
datelifeplot::plot_node_ages(all_summary_chronograms[[i]],
               title = paste(toupper(i), "Summary Chronogram"),
               plot_type = "phyloch",
               time_depth = plot_max_time,
               calibration_summary = calib_summ,
               pch = 19,
               color_pch = color_studies,
               cex_pch = 1,
               lwd_bars = 10,
               cex_legend = 0.7,
               x_legend = 0 ,
               y_legend = 270,
               mai1 = 1, 
               mai2 = 0,
               mai3 = 0.5,
               mai4 = 3.73,
               omi1 = 2,
               omi2 = 1,
               omi3 = 0.5,
               omi4 = 0,
               edge.width = 2,
               cex_tiplabels = 0.5,
               cex_axislabel = 1,
               cex_axis = 1,
               cex_title = 2,
               pos_title = -1,
               pos_axis = 2
               )
  dev.off() 
}
```

