---
title: "figure-cross-validation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{figure-cross-validation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(datelifeMS1)
devtools::load_all("../../datelifeplot")
devtools::load_all("../../datelife")
writing_dir <- "../figures/figure-cross-validation/"

# Based on function use_othercals3 from https://github.com/LunaSare/datelife_examples/blob/master/R/functions_data.R#L36
use_other_calibrations <- function(trees, other_calibrations, ...){
    res <- lapply(seq(trees), function(i){
      phy <- trees[[i]]
      phy$edge.length <- NULL
      print(i)
        xx <- suppressMessages(suppressWarnings(datelife::use_calibrations_bladj(phy,
            calibrations = other_calibrations[[i]], ...)))
        return(xx)
    })
    class(res) <- "multiPhylo"
    names(res) <- names(trees)
    res
}
load(file = "../data/frin_color_studies.rda")
```

## Cross validation of Fringillidae source chronograms.

We are using the object `frin_dsumm` and getting all source chronograms, and reordering the chronogram list alphabetically:

```{r}
load(file = "../data/frin_dsumm.rda")
phylo_all <- frin_dsumm$phylo_all
alphabetic_order <- order(names(phylo_all))
phylo_all <- phylo_all[alphabetic_order]
```

We will rename source chronograms to have a shorter study name:

```{r}
names(phylo_all) <- c("Barker et al. 2013",  # 1
                      "Barker et al. 2015 - chronogram 1",  # 2
                      "Barker et al. 2015 - chronogram 2",  # 3
                      "Burns et al. 2015",  # 4
                      "Claramunt et al. 2015",  # 5
                      "Gibb et al. 2015", # 6
                      "Hedges et al. 2015 - chronogram 1",  # 7
                      "Hedges et al. 2015 - chronogram 2",  # 8
                      "Hooper et al. 2017",  # 9
                      "Jetz et al. 2012 - chronogram 1",  # 10
                      "Jetz et al. 2012 - chronogram 2",  # 11
                      "Kimball et al. 2019 - chronogram 1", # 12
                      "Kimball et al. 2019 - chronogram 2", # 13
                      "Oliveros et al. 2019", # 14
                      "Price et al. 2014 - chronogram 1", # 15
                      "Price et al. 2014 - chronogram 2", # 16
                      "Roquet et al. 2014 - chronogram 1", # 17
                      "Roquet et al. 2014 - chronogram 2", # 18 
                      "Uyeda et al. 2017") # 19
```

Next, get calibrations for all other chronograms:
```{r}
# Function based on https://github.com/LunaSare/datelife_examples/blob/master/R/functions_data.R#L3
# get calibrations form all other chronograms not belonging to the same study
congruify_other_calibrations <- function(phylo_all){
    # dd <- duplicated(names(phyloall))
    res <- lapply(seq(phylo_all), function(i){
        cat(i, " -- ", names(phylo_all)[i], "\n")
        dd <- names(phylo_all) %in% names(phylo_all)[i]
        datelife::congruify_calibrations(phy = phylo_all[dd][[1]],
                                         chronograms = phylo_all[!dd])
    })
    names(res) <- names(phylo_all)
    res
}
```

```{r}
frin_other_calibs <- congruify_other_calibrations(phylo_all)
class(frin_other_calibs)
length(frin_other_calibs)
names(frin_other_calibs) <- names(phylo_all)
frin_other_calibs[[1]]
```

Date the topology based on other calibrations:

```{r}
# Function to date source chronograms with data from other chronograms
cross_date <- function(frin_other_calibs, index) {
  cat(index, " -- ", names(frin_other_calibs)[index], "\n")
  
  calibrations <- frin_other_calibs[[index]]
  
  if (length(calibrations) == 1) {
    message("... Source chronogram has no calibrations.")
    return(NA)
  }
  
  mm <- calibrations$matched_calibrations
  mm <- mm[mm$congruent,]
  mm <- mm[!is.na(mm$mrca_node_number),]
  calibrations$matched_calibrations <- mm
  res <- use_calibrations_bladj.matchedCalibrations(                     
                         calibrations = calibrations, 
                         type = "mean")
  res
}

validated_trees <- lapply(seq(frin_other_calibs), function(i) 
  cross_date(frin_other_calibs, index = i))

names(validated_trees) <- names(frin_other_calibs)

usethis::use_data(validated_trees, overwrite = TRUE)

```

## Plots

### get color scheme for studies

```{r}
# load(file = "../data/frin_color_studies.rda")
color_studies <- frin_color_studies
```

### plot chronograms, overlapped

```{r}
index = 1
ls(validated_trees[[index]])
plot_max_time <- 45

plot_all <- function(writing_dir, 
                     index = 1, 
                     height = 11*72, 
                     phylo_all, 
                     validated_trees, 
                     plot_max_time = 45,
                     cex_tiplabels = 1,
                     mai4 = 1) {
    #
  if (length(validated_trees[[index]]) == 1) {
    return(NA)
  }
  grDevices::png(filename = paste0(writing_dir, "cross_validation_", index, ".png"),
      height = height,
      width = 5*72,
      bg = "transparent")
  plot_chronogram(chronogram = phylo_all[[index]],
                         title = "",
                         time_depth = plot_max_time,
                         time_axis = FALSE,
                         axis_label = "",
                         edge.width = 3,
                         cex_tiplabels = cex_tiplabels,
                         mai4 = mai4)

  plot_chronogram(chronogram = validated_trees[[index]],
                         title = names(validated_trees)[index],
                         time_depth = plot_max_time,
                         time_axis = TRUE,
                         axis_label = "",
                         edge.width = 3,
                         edge.color = "#77889980",
                         cex_tiplabels = cex_tiplabels,
                         tip.color = "#77889900",
                         cex_axislabel = 1,
                         cex_axis = 1,
                         cex_title = 1.5,
                         pos_title = 2,
                         mai4 = mai4)
dev.off()
}

nice_height <- c(792, 792, 792, 792, NA, 300, 792, 792, 792, 792)
nice_cex <- c(1, 0.5, 0.5, 1, NA, 1, 0.2, 0.2, 0.8, 0.5)
i = 10
for (i in 1:8) {
    plot_all(writing_dir = writing_dir,
             index = i, 
             height = nice_height[i],
             phylo_all = phylo_all,
             validated_trees = validated_trees,
             cex_tiplabels = nice_cex[i],
             mai4 = 2.37)
}
```


### x-y graphs

Get all age values:

```{r}
res_all <- vector(mode = "list")
ages_all <- vector(mode = "list")

for (index in seq(validated_trees)) {
  res <- list(cross_val = validated_trees[[index]],
              original = phylo_all[[index]])
  
  res_all <- c(res_all, list(res))

  if (!inherits(res$cross_val, "phylo")) {
    ages <- NA
  } else {
    ages <- lapply(res, ape::branching.times)
  }
  
  ages_all <- c(ages_all, list(ages))
  
}

names(ages_all) <- names(res_all) <- names(phylo_all)
```  

Get max age:

```{r}
max_age <- max(abs(unlist(ages_all)), na.rm = TRUE)
max_age <- ceiling(max_age)
max_age
```

Get all age differences:

```{r}
age_diffs <- function(ages) {
  if (!inherits(ages, "list")) {
    return(NA)
  }
  ages$cross_val - ages$original
}
  
  
age_diffs_all <- lapply(ages_all, age_diffs)

max_diff <- max(abs(unlist(age_diffs_all)), na.rm = TRUE)
max_diff <- ceiling(max_diff)
max_diff
```

Produce an x-y plot for each chronogram:

```{r}
for (index in seq(validated_trees)) {
  if (length(validated_trees[[index]]) == 1) {
    next
  }
  res <- list(cross_val = validated_trees[[index]],
              original = phylo_all[[index]])
  
  ages <- lapply(res, ape::branching.times)
  
  nodeN <- seq(length(ages$original))
  
###################################################################
# Plot original node ages vs cross-validation chronogram:
###################################################################
  filename <- paste0(writing_dir, "xy_plot", index, "_ages.pdf")
  #################################################################
  # order ages decreasingly
  #################################################################
  X <- ages$original
  ordered <- order(X, decreasing = TRUE)
  X <- X[ordered] 
  Y <- ages$cross_val[ordered]
  #################################################################
  # plot and save as pdf
  #################################################################
  pdf(file = filename,
      height = 5,
      width = 5)
  
  par_mai <- par()$mai
  par_mai[2] <- 1
  par(mai = par_mai)

  plot(x = seq(from = -max_age, to = 0), 
       y = seq(from = -max_age, to = 0),
       xlab = "Original node ages \n Time (Ma)",
       ylab = "Cross validated node ages \n Time (Ma)",
       type = "l",
       main = names(validated_trees)[index])
  
  points(-X, 
       -Y, 
       pch = 21,
       type = "p",
       col = "gray",
       bg = color_studies[index],
       cex = 1.5)
  
  dev.off()
##############################################################################
# Plot difference between original and cross-validated ages
##############################################################################
  
  filename <- paste0(writing_dir, "xy_plot", index, "_differences.pdf")
  
  diffs <- ages$original - ages$cross_val
  colors <- ifelse(diffs < 0, "salmon", "#4B0082") # #salmon and indigo
  
  pdf(file = filename,
      height = 5,
      width = 5)
  
  par_mai <- par()$mai
  par_mai[2] <- 1
  par(mai = par_mai)

    plot(nodeN, 
       ylim = c(-max_diff, max_diff),
       xlab = "Node number",
       ylab = "Age difference \n (original - cross validated age)",
       type = "n",
       main = names(validated_trees)[index])
    
  points(nodeN,
         diffs, 
         pch = 21,
         col = "gray",
         bg = colors,
         type = "p",
         cex = 1.5)
  
  abline(h = 0)
  
  dev.off()
}
```

A single graph for x-y plots.

Plot original node ages vs cross-validation chronogram:

```{r}

```

Plot difference between original and cross-validated ages

```{r}
##############################################################################
##############################################################################
# get total number of nodes congruified to chosen topology, so ages are comparable
##############################################################################
##############################################################################

node_numbers <- sapply(validated_trees[!is.na(validated_trees)], function(vt) {
  vt$calibrations$source_mrca_node_number
})

node_numbers <- unique(unlist(node_numbers)[!sapply(unlist(node_numbers), is.na)])

max(node_numbers)

validated_trees[[1]]$calibrations$mrca_node_number

##############################################################################
##############################################################################
# get node numbers from congruified and matched node ages
##############################################################################
##############################################################################


node_names_all <- validated_trees[[1]]$calibrations$source_mrca_node_name

node_names_source <- validated_trees[[1]]$calibrations$mrca_node_name


nodeN <- seq(length(ages$original))

filename <- paste0(writing_dir, "xy_plot_0.pdf")

pdf(file = filename,
    height = 5,
    width = 5)
plot(nodeN, 
     ylim = c(-max(abs(diffs)), max(abs(diffs))),
     xlab = "Node number",
     ylab = "Age difference \n (original - cross validated age)",
     type = "n")
  
for (i in seq(validated_trees)) {

  index <- i
  
  if (length(validated_trees[[index]]) == 1) {
    next
  }

  points(nodeN,
         diffs, 
         pch = 19,
         col = colors,
         type = "p")
  abline(h = 0)
  dev.off()

}
```
