---
title: "Fringillidae Topologies"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fringillidae Topologies}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

```{r setup}
devtools::load_all("../../datelifeplot")
devtools::load_all("../../datelife")
writing_dir <- "../figures/fringillidae-topologies/"
load(file = "../data/fringillidae_v0.6.2.rda")
```

```{r, eval = TRUE}
rotl::tnrs_match_names("fringillidae")
```
```{r, eval = TRUE}
rotl::tol_node_info(839319)
```


```{r}
rotl::tol_mrca(ott_ids = fringillidae_v0.6.2$dquery$ott_ids)
```


```{r, eval = TRUE}
fringillidae_monophyletic <- rotl::tol_subtree(node_id = "mrcaott4083ott370807",
                                               label_format = "name")

fringillidae_monophyletic
```

```{r}
fringillidae_v0.6.2$dquery$phy
  
fringillidae_paraphyletic <- rotl::tol_induced_subtree(ott_ids = fringillidae_v0.6.2$dquery$ott_ids,
                                          label_format = "name")

fringillidae_paraphyletic
```


Get species names belonging to the Fringillidae according to the NCBI taxonomy:

```{r}
frin_species_names <- fringillidae_v0.6.2$dquery$cleaned_names
frin_species_names <- gsub(" ", "_", frin_species_names)

frin_species_ids <- fringillidae_v0.6.2$dquery$ott_ids
```

Some NCBI species are not in OpenTree monophyletic tree:

```{r}
species_in_mono <- frin_species_names %in% fringillidae_monophyletic$tip.label
unname(frin_species_names[!species_in_mono])

ids_in_mono <- frin_species_ids %in% gsub("ott", "", fringillidae_monophyletic$tip.label)

sum(ids_in_mono)
```

Just one NCBI species is not in OpenTree induced (paraphyletic) tree

```{r}
species_in_para <- frin_species_names %in% fringillidae_paraphyletic$tip.label

unname(frin_species_names[!species_in_para])
```

Tip colors for plotting the tree:

```{r}
tip_colors <- rep("black", ape::Ntip(fringillidae_monophyletic))
in_mono <- fringillidae_monophyletic$tip.label %in% frin_species_names
sum(in_mono)
tip_colors[in_mono] <- "red"
```

Edge colors:

```{r}
# Get edge ancestor positions

# create a vector of all tip numbers:
tip_numbers <- seq(ape::Ntip(fringillidae_monophyletic))

# find row position of tip numbers that are in tree: 
row_position_tips <- fringillidae_monophyletic$edge[,2] %in% tip_numbers[in_mono]
sum(row_position_tips)

length(row_position_tips) == nrow(fringillidae_monophyletic$edge)

edge_colors <- rep("black", nrow(fringillidae_monophyletic$edge))
edge_colors[row_position_tips] <- "red"


edge_numbers <- fringillidae_monophyletic$edge[row_position_tips,1]

row_position_edge1 <- fringillidae_monophyletic$edge[,2] %in% edge_numbers
length(row_position_edge1)
sum(row_position_edge1)

edge_colors[row_position_edge1] <- "red"
```

Make it a function

```{r}
phy <- fringillidae_monophyletic
tip_labels <- frin_species_names
get_edge_positions <- function(phy, tip_labels) {
  # get tip_labels positions in vector phy$tip.label
  phy$tip.label <- gsub(" ", "_", phy$tip.label)
  tip_labels <- gsub(" ", "_", tip_labels)
  tip_labels_in_phy <- phy$tip.label %in% tip_labels
  # sum(tip_labels_in_phy) == length(tip_labels)
  
  # create a boolean vector of length == to nrow(phy$edge)
  edges <- rep(FALSE, nrow(phy$edge))
  # get row positions
  descendant_numbers <- seq(ape::Ntip(phy))[tip_labels_in_phy]
  ancestor_row_position <- phy$edge[,2] %in% descendant_numbers
  edges[ancestor_row_position] <- TRUE
  while (sum(ancestor_row_position) > 0) {
    descendant_numbers <- phy$edge[ancestor_row_position,1]
    ancestor_row_position <- phy$edge[,2] %in% descendant_numbers
    edges[ancestor_row_position] <- TRUE
  }
  return(edges)
}
```

Get edge positions and create edge color vector

```{r}
ladderized <- ladderize(phy = fringillidae_monophyletic)
edge_pos <- get_edge_positions(phy = ladderized,
                               tip_labels = frin_species_names)
edge_colors_anti <- edge_colors <- rep("black", nrow(ladderized$edge))
edge_colors[edge_pos] <- "red"

edge_colors_anti[!edge_pos] <- anti_color

# Tried to control edge width with a vector, so tip edges were thinner but inner edges were thicker.
# It did not work
edge_width <- rep(1, nrow(ladderized$edge))
edge_width[seq(ape::Ntip(ladderized))] <- 0.1
```


```{r}
for (i in c("fan", "radial", "phylogram")) {
  pdf(file = paste0(writing_dir, 
                    "fringillidae_monophyletic_",
                    i, ".pdf"),
      height = ifelse("phylogram" %in% i, 40, 6))
  par(mai = c(0,0,0,0))
  ape::plot.phylo(ladderized, 
                  type = i,
                  cex = ifelse("phylogram" %in% i, 0.1, 0.05),
                  tip.color = tip_colors,
                  label.offset =ifelse("phylogram" %in% i, 5, 0.001),
                  edge.color = edge_colors,
                  edge.width = 1 #ifelse("phylogram" %in% i, 1, 0.3)
       )
  dev.off()  
}
```

Monophyletic with other colors:

Tip and edge colors for plotting the tree:

```{r}
tip_colors_anti <- rep("black", ape::Ntip(ladderized))
in_mono <- ladderized$tip.label %in% frin_species_names
sum(in_mono)

edge_pos <- get_edge_positions(phy = ladderized,
                               tip_labels = frin_species_names)
edge_colors_anti <- rep("black", nrow(ladderized$edge))
```

```{r}
color_loop <- c("#2F4F4F40", "#FFE4B550", "#DDA0DD50")
names(color_loop) <- c("grey", "mocassin", "plum")
for (i in seq(color_loop)) {
  anti_color <- color_loop[i]
  tip_colors_anti[!in_mono] <- anti_color
  edge_colors_anti[!edge_pos] <- anti_color
  pdf(file = paste0(writing_dir, 
                  "fringillidae_monophyletic_fan_anti_",
                  names(color_loop)[i], 
                  ".pdf"),
    height = 6)
  par(mai = c(0,0,0,0))
  ape::plot.phylo(ladderized,
                  type = "fan",
                  cex = 0.05,
                  tip.color = tip_colors_anti,
                  label.offset = 0.001,
                  edge.color = edge_colors_anti,
                  edge.width = 1)
  dev.off()

}

```

Fringillidae induced subtree

```{r}
pdf(file = paste0(writing_dir, 
                  "fringillidae_induced_phylogram.pdf"),
    height = 6,
    width = 3)
par(mai = c(0,0,0,0))
ape::plot.phylo(ape::ladderize(fringillidae_paraphyletic),
                type = "phylogram",
                cex = 0.1)
dev.off()
```
