---
title: "Fringillidae Open Tree Taxonomy"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fringillidae Open Tree Taxonomy}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(datelife)
```



`datelife` function `get_ott_children()` to get children at a certain taxonomic rank returns 289 species within the family Fringillidae:

```{r}
spp1 <- get_ott_children(ott_id = 839319)

length(spp1$Fringillidae$ott_id) # returns 289 species

# spp2 <- get_ott_children(ott_id = 839319, invalid = "")
# length(spp2$Fringillidae$ott_id)

```

But the taxonomic subtree returns 632 children within family Fringillidae.
Not all of those are species, though.
```{r}
taxon_names <- rotl::taxonomy_subtree(ott_id = 839319, 
                               output_format = "taxa", 
                               label_format = "name")
ls(taxon_names)
length(taxon_names$tip_label) #632
head(taxon_names$tip_label)
tail(taxon_names$tip_label, 20)

# Some are genus names that might not belong to Fringillidae, anyways:
tnrs_match(input = "Trichas")
rotl::is_in_tree(ott_ids = 5341190) # FALSE
grep("Trichas", taxon_names$tip_label) # index 626

# google search for "trichas finch" returned the binomial specific geothlypis trichas:
tnrs_match(input = "Geothlypis trichas")
rotl::is_in_tree(ott_ids = 998127) # TRUE

grep("Geothlypis_trichas", taxon_names$tip_label) # it is not in taxonomic subtree
grep("Geothlypis", taxon_names$tip_label) # neither this one

taxon_info <- rotl::taxonomy_taxon_info(ott_ids = 998127, include_lineage = TRUE) # Geothlypis trichas belongs to family Parulidae, not Fringillidae

paste0(sapply(taxon_info$`998127`$lineage, "[", "rank"),
       " = ",
       sapply(taxon_info$`998127`$lineage, "[", "name"))
```

What ranks exist among the children of Fringillidae?
```{r}
# Get ott ids of all taxonomic children
ott_ids <- rotl::taxonomy_subtree(ott_id = 839319, 
                               output_format = "taxa", 
                               label_format = "id")

length(ott_ids$tip_label) # 632 ott ids
head(ott_ids$tip_label)
ott_ids_numeric <- as.numeric(gsub("ott", "", ott_ids$tip_label))

in_synth_tree <- rotl::is_in_tree(ott_ids = ott_ids_numeric)
sum(in_synth_tree) # 590 children are in synthetic tree

# Get taxonomic info for all children:
taxon_info <- rotl::taxonomy_taxon_info(ott_ids = ott_ids_numeric)

taxon_info$`913583`$rank

# Get unique ranks:
unique(unlist(sapply(taxon_info, "[", "rank")))
```

Which ones are species?
```{r}
are_species <- unlist(sapply(taxon_info, "[", "rank")) == "species"
sum(are_species) # 462 species
```

Which species are in synthetic tree?
```{r}
sum(are_species & in_synth_tree) # 452 are species and in synthetic tree

flag_number <- sapply(sapply(taxon_info, "[", "flags"), length)
no_flags <- flag_number == 0
sum(no_flags) # 389 have no flags

sum(are_species & in_synth_tree & no_flags) # 389 are species and in synthetic tree and have no flags

species_names <- unlist(sapply(taxon_info[no_flags], "[", "name"))
length(species_names) # 389

```

Checking that species from subspecies are included in the vector of species_names, revelaeed that they are excluded from the final species list:

```{r}
taxon_names$tip_label[55:75]

grep("Bucanetes_githagineus", taxon_names$tip_label) # only subspecies for this binomial are present in the taxonomic subtree

# so, when removing everything but species, that species is no longer included in the list of species names:
grep("Bucanetes_githagineus", species_names) # only subspecies for this binomial are present in the taxonomic subtree

# Hence, species from subspecies are excluded from the list of children 
```

Let's recover species from subspecies.
Which ones are subspecies?
```{r}
are_subspecies <- unlist(sapply(taxon_info, "[", "rank")) == "subspecies"
sum(are_subspecies) # 138 subspecies

subspecies <- unname(unlist(sapply(taxon_info[are_subspecies], "[", "name")))

# Get unique species names only:
split <- strsplit(subspecies, split = " ") # split character string by blank spaces
binomial <- lapply(split, "[", -3) # remove the subspecies epithet
species_from_subspecies <- unique(sapply(binomial, paste0, collapse = " ")) 
length(species_from_subspecies) # 67 species extracted from subspecies

# are any species from subspecies in the species_names vector?
is.na(match(species_from_subspecies, species_names)) # nope, all species from subspecies are new!

# Get ott_ids of species from subspecies
spp_from_sub_ott_ids <- rotl::tnrs_match_names(species_from_subspecies)$ott_id

# Get their taxon info
spp_from_sub_taxon_info <- rotl::taxonomy_taxon_info(spp_from_sub_ott_ids)

is_in_synth_sub <- rotl::is_in_tree(spp_from_sub_ott_ids) 
sum(is_in_synth_sub) # 66 species from sub are in synth tree
```

Put together all data from species and species from subspecies

```{r}
ott_ids_all <- c(ott_ids_numeric, spp_from_sub_ott_ids)
length(ott_ids_all) # 699
taxon_info_all <- c(taxon_info, spp_from_sub_taxon_info)
length(taxon_info_all) # 699

# check that order of ott_ids_all is the same as taxon_info_all
all(ott_ids_all == unname(unlist(sapply(taxon_info_all, "[", "ott_id"))))

is_in_synth_all <- rotl::is_in_tree(ott_ids = ott_ids_all)
sum(is_in_synth_all) # 656 are in synth

are_species_all <- unlist(sapply(taxon_info_all, "[", "rank")) == "species"
sum(are_species_all) #  529 are species

spp_in_synth <- is_in_synth_all & are_species_all
sum(spp_in_synth) # 518 are species and are in synth tree

species_names_all <- unlist(sapply(taxon_info_all[spp_in_synth], "[", "name"))
head(species_names_all)
species_rank_all <- unlist(sapply(taxon_info_all[spp_in_synth], "[", "rank"))
all(species_rank_all == "species") # they are all species
species_ott_ids_all <- unlist(sapply(taxon_info_all[spp_in_synth], "[", "ott_id"))
spp_all_taxon_info <- rotl::taxonomy_taxon_info(ott_ids = species_ott_ids_all, include_lineage = TRUE)

# spp_all_taxon_info$`328908`$lineage

spp_all_lineage <- sapply(spp_all_taxon_info, "[", "lineage")

# spp_all_lineage$`328908.lineage`[[1]]

family <- sapply(seq(spp_all_lineage), function(i) {
  rank_names <- unlist(sapply(spp_all_lineage[[i]], "[", "rank"))
  length(spp_all_lineage[[i]]) == length(rank_names)
  names(spp_all_lineage[[i]]) <- rank_names
  spp_all_lineage[[i]]$family$name
  })

all(family == "Fringillidae")
sum(family == "Fringillidae") # 518
```

Let's make it a function:

```{r}
# Get all species belonging to a taxon from the Open Tree of Life Taxonomy
get_opentree_species <- function(taxon_name, ott_id, synth_tree_only = TRUE) {
  if (missing(ott_id)) {
    taxon_tnrs <- rotl::tnrs_match_names(names = taxon_name)
    ott_id <- taxon_tnrs$ott_id
    message("Getting species for OTT taxon '", taxon_tnrs$unique_name, "' with OTT id number = ", taxon_tnrs$ott_id, ".")
  }
  children_names <- rotl::taxonomy_subtree(ott_id = ott_id, 
                               output_format = "taxa", 
                               label_format = "name")$tip_label
  children_ott_ids <- rotl::taxonomy_subtree(ott_id = ott_id, 
                               output_format = "taxa", 
                               label_format = "id")
  children_ott_ids_numeric <- as.numeric(gsub("ott", "", children_ott_ids$tip_label))
  message("... Obtaining taxonomic info for all children within this taxon might take a while. Appreciate your patience.")
  taxon_info <- rotl::taxonomy_taxon_info(ott_ids = children_ott_ids_numeric)
  # Get unique ranks:
  message("... OTT children for this taxon belong to taxonomic ranks: ", paste0(unique(unlist(sapply(taxon_info, "[", "rank"))), collapse = ", "), ".")
  # Get subspecies
  are_subspecies <- unlist(sapply(taxon_info, "[", "rank")) == "subspecies"
  message("... There are ", sum(are_subspecies), " subspecies.")
  if(sum(are_subspecies) > 0) {
     subspecies <- unname(unlist(sapply(taxon_info[are_subspecies], "[", "name")))
    # Get unique species names from subspecies list:
    split <- strsplit(subspecies, split = " ") # split character string by blank spaces
    binomial <- lapply(split, "[", -3) # remove the subspecies epithet
    species_from_subspecies <- unique(sapply(binomial, paste0, collapse = " ")) 
    length(species_from_subspecies) 
    # Get ott_ids of species from subspecies
    spp_from_sub_ott_ids <- rotl::tnrs_match_names(species_from_subspecies)$ott_id
    # Get their taxon info
       message("... Obtaining taxonomic info for missing species within the taxon might take a while. Appreciate your patience.")
    spp_from_sub_taxon_info <- rotl::taxonomy_taxon_info(spp_from_sub_ott_ids)
    #Unify all taxon_info
    taxon_info <- c(taxon_info, spp_from_sub_taxon_info)   
  }
  ott_ids_all <-  unname(unlist(sapply(taxon_info, "[", "ott_id")))
  is_in_synth_all <- rotl::is_in_tree(ott_ids = ott_ids_all)
  are_species_all <- unlist(sapply(taxon_info, "[", "rank")) == "species"
  message("... There are ", sum(are_species_all), " species.")
  spp_in_synth <- is_in_synth_all & are_species_all
  message("... There are ", sum(spp_in_synth), " species of '", taxon_tnrs$unique_name, "' in the OpenTree synthetic tree.")
  return_species <- unlist(ifelse(synth_tree_only, 
                           list(spp_in_synth), 
                           list(are_species_all)))
  species_names_all <- unlist(sapply(taxon_info[return_species], "[", "name")) 
  return(list(species_names = species_names_all, ott_ids = ott_ids_all))
}
```

Test:
```{r}
anatidae_spp <- get_opentree_species(taxon_name = "anatidae")

anatidae_spp$species_names
```
