---
title: "node-age-comparison"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{node-age-comparison}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
devtools::install_github("phylotastic/datelife")

```

## Data for node age comparison

We will use the `datelifeQuery`, `datelifeResult` and `datelifeSummary` objects that we created in "Fringillidae.Rmd". These objects are saved in the data folder and can be loaded with the function `data()`:

```{r}
data("dquery", package = "datelifeMS1")
data("dres", package = "datelifeMS1")
data("dsumm", package = "datelifeMS1")
```

## Extract all available calibrations

We can extract calibrations directly from the chronograms summarized from the `datelifeResult` object `dres` that are stored in `dsumm$phylo_all`. 

TODO: Create a print method for a phylo_all datelifeResult that summarizes number of studies and number of chronograms available per study.


Let's explore them:

```{r}
dsumm$phylo_all
```
We have 13 chronograms in there. Let's get the reference for the study they were originally published on:

```{r}
names(dsumm$phylo_all)
```

Reading through the references, we can see that there are some duplicated ones. This means that we have several chronograms from each of those studies.
We can identify them with the function `duplicated()`:

```{r}
names(dsumm$phylo_all)[duplicated(names(dsumm$phylo_all))]
```


We will use the function `datelife::extract_calibrations_phylo`.




```{r}
all_calibrations <- datelife::extract_calibrations_phylo(input = dsumm$phylo_all,
                                               each = FALSE)
class(all_calibrations)
names(all_calibrations)
nrow(all_calibrations)
```
Our `all_calibrations` object is a `data.frame` of `r nrow(all_calibrations)` rows and 6 columns: 
 - `$MRCA` 
 - `$MaxAge`
 - `$MinAge`
 - `$taxonA`
 - `$taxonB`
 - `$reference`

<!--
```{r}
length(all_calibrations$MRCA)
length(unique(all_calibrations$MRCA))
all_calibrations$MRCA[duplicated(all_calibrations$MRCA)]


calibs_each <- datelife::extract_calibrations_phylo(input = dsumm$phylo_all,
                                               each = TRUE)

# length(calibs_each)
# class(calibs_each)
# names(calibs_each)
# # sapply(calibs_each, nrow)
# sum(sapply(calibs_each, nrow))
```

`all_calibrations` and `calibs_each` have the same information.
-->

## Choose a topology

We have several options within the `datelifeSummary` object `dsumm`:

```{r}
dsumm$phylo_median
dsumm$phylo_sdm
dsumm$phylo_biggest
```

However, none of this contain all of our `r length(dquery$cleaned_names)` input taxon names. Our best shot to get a topology with most of our input taxa is getting an Open Tree of Life subtree with `rotl::tol_induced_subtree()`. We will use the Open Tree of Life Taxonomy (OTT) identifier numbers in our `datelifeQuery` object, `dquery$ott_ids` as input:

```{r, eval FALSE}
rotl::tnrs_match_names("fringillidae")
#>   search_string  unique_name approximate_match ott_id is_synonym
#> 1  fringillidae Fringillidae             FALSE 839319      FALSE
#>   flags number_matches
#> 1                    1

rotl::is_in_tree(ott_ids = 839319)
#> [1] FALSE

# rotl::tol_subtree(ott_id = "839319")

ot <- rotl::tol_induced_subtree(ott_ids = dquery$ott_ids, label_format = "name")
ot
```

## Match calibrations to a chosen topology

Now we can match the extracted calibrations unto nodes of our chosen tree, `ot`. For this we will use the function `datelife::match_all_calibrations()`:

```{r}
matched <- datelife::match_all_calibrations(phy = dsumm$phylo_sdm, calibrations = all_calibrations)

class(matched)
names(matched)
names(matched$matched_calibrations)
nrow(matched$matched_calibrations)
class(matched$matched_calibrations)
```

Write data down as a csv file for visual inspection and archiving:

```{r}
write.csv(matched$matched_calibrations, file = "../data-raw/fringillidae-sample-matched-calibrations.csv")
```


## Summarize our `matchedCalibrations` object

We will use the function `datelife::summary.matchedCalibrations()`:

```{r}
csumm <- datelife:::summary.matchedCalibrations(matched$matched_calibrations)
names(csumm)
nrow(csumm$not_in_phy)
```
There are 12 taxon pairs that were not present in our target tree topology, as shown in `csumm$not_in_phy`.

Our calibrations of interest are in `csumm$in_phy`:

```{r}
sapply(colnames(csumm$in_phy),function(i) is.factor(csumm$in_phy[[i]]))

plot(-csumm$in_phy$MaxAge, csumm$in_phy$mrca_node_number, xlab = "Time (myrs)", ylab = "Node Number", pch = 20, col = csumm$in_phy$mrca_node_name)

plot(-csumm$in_phy$MaxAge[1:300], csumm$in_phy$mrca_node_number[1:300], xlab = "Time (myrs)", ylab = "Node Number", pch = as.character((csumm$in_phy$mrca_node_number-289)[1:85]), col = csumm$in_phy$reference[1:300])

plot(-csumm$in_phy$MaxAge, csumm$in_phy$mrca_node_number, xlab = "Time (myrs)", ylab = "Node Number", col = csumm$in_phy$reference, pch = 16)
```
Export as svg

```{r}
svg(filename = "../figures/node_ages_scatterplot.svg")
plot(-csumm$in_phy$MaxAge, csumm$in_phy$mrca_node_number, xlab = "Time (myrs)", ylab = "Node Number", col = csumm$in_phy$reference, pch = 16)
dev.off()

svg(filename = "../figures/fringillidae_sample_tree.svg")
ape::plot.phylo(dsumm$phylo_sdm, cex = 0.7)
dev.off()



```


```{r}
matched$phy$calibration_distribution
ls(matched$phy)
```

```{r}
phyloch:::HPDbars
data(hordeum, package = "phyloch")
hordeum
ls(hordeum)
head(hordeum$`height_95%_HPD_MAX`)
length(hordeum$`height_95%_HPD_MAX`)
hordeum$edge[,1] # the first column contains the node numbers


plot(hordeum, edge.color = "gray", tip.color = 0, 
    x.lim = c(-5.24603, 23.26244))
HPDbars(hordeum, col = "skyblue", lwd = 5, nodes = 1)
ape::nodelabels()

ot$ages_MIN <- 
ot$ages_MAX <- 
```
