---
title: "node-age-comparison"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{node-age-comparison}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
#devtools::install_github("phylotastic/datelife")
devtools::install()
```

## Data for node age comparison

We will use the `datelifeQuery`, `datelifeResult` and `datelifeSummary` objects that we created in "Fringillidae.Rmd". These objects are saved in the data folder and can be loaded with the function `data()`:

```{r}
data("dquery", package = "datelifeMS1")
data("dres", package = "datelifeMS1")
data("dsumm", package = "datelifeMS1")
```

## Extract all available calibrations

We can extract calibrations directly from the chronograms summarized from the `datelifeResult` object `dres` that are stored in `dsumm$phylo_all`. 

TODO: Create a print method for a phylo_all datelifeResult that summarizes number of studies and number of chronograms available per study.


Let's explore them:

```{r}
dsumm$phylo_all
```
We have 13 chronograms in there. Let's get the reference for the study they were originally published on:

```{r}
names(dsumm$phylo_all)
```

Reading through the references, we can see that there are some duplicated ones. This means that we have more than one chronogram for some of those studies.
We can identify them with the function `duplicated()`:

```{r}
names(dsumm$phylo_all)[duplicated(names(dsumm$phylo_all))]
```

### Getting node ages for our taxa


Now, we can use the function `datelife::extract_calibrations_phylo`.


```{r}
all_calibrations <- datelife::extract_calibrations_phylo(input = dsumm$phylo_all,
                                               each = FALSE)
class(all_calibrations)
names(all_calibrations)
nrow(all_calibrations)
```
Our `all_calibrations` object is a `data.frame` of `r nrow(all_calibrations)` rows and 6 columns: 
 - `$MRCA` 
 - `$MaxAge`
 - `$MinAge`
 - `$taxonA`
 - `$taxonB`
 - `$reference`

<!--
```{r}
length(all_calibrations$MRCA)
length(unique(all_calibrations$MRCA))
all_calibrations$MRCA[duplicated(all_calibrations$MRCA)]


calibs_each <- datelife::extract_calibrations_phylo(input = dsumm$phylo_all,
                                               each = TRUE)

# length(calibs_each)
# class(calibs_each)
# names(calibs_each)
# # sapply(calibs_each, nrow)
# sum(sapply(calibs_each, nrow))
```

`all_calibrations` and `calibs_each` have the same information.
-->

## Choose a topology

We have several options within the `datelifeSummary` object `dsumm`:

```{r}
dsumm$phylo_median
dsumm$phylo_sdm
dsumm$phylo_biggest
```

However, none of this contain all of our `r length(dquery$cleaned_names)` input taxon names. Our best shot to get a topology with most of our input taxa is getting an Open Tree of Life subtree with `rotl::tol_induced_subtree()`. We will use the Open Tree of Life Taxonomy (OTT) identifier numbers in our `datelifeQuery` object, `dquery$ott_ids` as input:

```{r, eval FALSE}
rotl::tnrs_match_names("fringillidae")
#>   search_string  unique_name approximate_match ott_id is_synonym
#> 1  fringillidae Fringillidae             FALSE 839319      FALSE
#>   flags number_matches
#> 1                    1

rotl::is_in_tree(ott_ids = 839319)
#> [1] FALSE

# rotl::tol_subtree(ott_id = "839319")

ot <- rotl::tol_induced_subtree(ott_ids = dquery$ott_ids, label_format = "name")
ot
```

## Match calibrations to a chosen topology

Now we can match the extracted calibrations unto nodes of our chosen tree, `ot`. For this we will use the function `datelife::match_all_calibrations()`:

```{r}
matched <- datelife::match_all_calibrations(phy = dsumm$phylo_sdm, calibrations = all_calibrations)

class(matched)
names(matched)
names(matched$matched_calibrations)
nrow(matched$matched_calibrations)
class(matched$matched_calibrations)
```

Write data down as a csv file for visual inspection and archiving:

```{r}
write.csv(matched$matched_calibrations, file = "../data-raw/fringillidae-all-matched-calibrations-sdm.csv")
write.csv(matched$matched_calibrations, file = "../data-raw/fringillidae-sample-matched-calibrations.csv")
```


## Summarize our `matchedCalibrations` object

We will use the function `datelife::summary.matchedCalibrations()`:

```{r}
csumm <- datelife:::summary.matchedCalibrations(matched$matched_calibrations)
names(csumm)
nrow(csumm$not_in_phy)
```
There are 12 taxon pairs that were not present in our target tree topology, as shown in `csumm$not_in_phy`.

Our calibrations of interest are in `csumm$in_phy`:
We will use the function `plot_node_ages` defined here (TODO: add it to datelifeplots) to plot the tree and the node ages from the calibrations:

```{r}
# the node position on the y axis is given by pp$yy
# pp$yy also stores the positions of the tips
# the y coordinate of nodes start at pp$yy[ape::Ntip + 1]
#' Add Node Age/Calibration Data to a Plotted Phylogeny
#' calibration_summary An output of [datelife:::summary.matchedCalibrations()]
#' color A named vector of colors. Names must correspond to study references. 
#'       If vector is not named, colors will be recycled.
plot_node_ages <- function(phy, cex_tips = 1, calibration_summary, color, pch = 16) {

  # obtain max x lim from ages
  x_max <- max(c(calibration_summary$in_phy$MinAge, calibration_summary$in_phy$MaxAge))
  # plot.phylo
  ape::plot.phylo(phy,
                  cex = cex_tips,
                  plot = FALSE)
  # get plotting x.lim
  library(ape) # we need this so it loads object .PlotPhyloEnv 
  lastPP_x.lim <- get("last_plot.phylo", envir = .PlotPhyloEnv)$x.lim[2]
  ape::plot.phylo(phy,
                  cex = cex_tips,
                  x.lim = c(-5,lastPP_x.lim))
  # get recorded plotting parameters
  lastPP <- get("last_plot.phylo", envir = .PlotPhyloEnv)
  # we use lastPP$xx positions to get plotting coordinates for node ages
  if (all(calibration_summary$in_phy$MinAge == calibration_summary$in_phy$MaxAge)) {
    x_ages <- max(lastPP$xx) - calibration_summary$in_phy$MinAge
  }
  # we use lastPP$yy positions to get plotting coordinates for node numbers
  if (all(calibration_summary$in_phy$mrca_node_number > lastPP$Ntip)) {
    # case in which mrca_node_numbers start at Ntip + 1
    # we can directly use mrca_node_numbers as index for lastPP$yy
    y_nodes <- lastPP$yy[calibration_summary$in_phy$mrca_node_number]
  }
  # using references to color points
  if (missing(color)) {
    color <- calibration_summary$in_phy$reference
  }
  points(x_ages,
         y_nodes,
         col = color,
         pch = pch)
  ape::axisPhylo(1)
  graphics::mtext("Time (myrs)", side = 1, line = 2, at = max(lastPP$xx) * 0.5)
}
svg(filename = "../figures/fringillidae_all_median_ages.svg",
    height = 20,
    width = 10)
plot_node_ages(dsumm$phylo_median,
               cex_tips = 0.7,
               calibration_summary = csumm,
               pch = 18) # a diamond
dev.off()

svg(filename = "../figures/fringillidae_all_sdm_ages.svg",
    height = 20,
    width = 10)
plot_node_ages(dsumm$phylo_sdm,
               cex_tips = 0.7,
               calibration_summary = csumm,
               pch = 18) # a diamond
dev.off()
```
