---
title: "opentree chronograms"
output: pdf_document
date: "2023-06-07"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(datelife)
data("opentree_chronograms")
```

```{r}
layouts <- c("rectangular", "ellipse", "circular", "fan", "roundrect")
i =1
for (i in seq(layouts)) {
  tree <- opentree_chronograms$trees[[i]]
  class(tree) <- "phylo" 
  ggtree::ggtree(tree, 
                 layout = layouts[i],
                 ladderize = FALSE)
  ggplot2::ggsave(filename = paste0("../figures/opentree_chronograms/chronogram_", i, ".png"),
                  device = "png")  
}
```
Some layouts do not work, maybe bc they are meant for unrooted trees:

```{r error=TRUE}
ggtree::ggtree(tree, layout = "radial")
```


```{r}
set.seed(seed = 1976452)
layouts_all <- sample(x = layouts, 
                      size = length(opentree_chronograms$trees), 
                      replace = TRUE)
```

```{r}
for (i in seq(opentree_chronograms$trees)) {
  print(i)
  tree <- opentree_chronograms$trees[[i]]
  class(tree) <- "phylo" 
  ggtree::ggtree(tree, 
                 layout = layouts_all[i],
                 ladderize = FALSE) +
    ggfun::theme_transparent()
  ggplot2::ggsave(
    filename =  paste0("../figures/opentree_chronograms/chronogram_", 
                       i, ".png"),
    device = "png",
    bg = "transparent")  
}
```

Create an image combining all, most or some sample of opentree chronograms.

Tip to use cowplot package from https://www.youtube.com/watch?v=_gfcIXxXe00.

Tip for good resolution on output image from https://stackoverflow.com/questions/50026862/the-draw-image-function-from-cowplot-results-in-blurred-pdfs.

A test:

```{r}
# install.packages("cowplot")
library(cowplot)
library(magick)
ggdraw(xlim = c(-1,1),
      ylim = c(0, 4)) +
  draw_image(image = "../figures/opentree_chronograms/chronogram_1.png",
             x = -1,
             y = 0,
             scale = 0.5) +
  draw_image(image = "../figures/opentree_chronograms/chronogram_2.png",
             x = 0,
             y = 3,
             scale = 0.7)
ggsave2(filename =  "../figures/opentree_chronograms/all_chronograms_test.png",
        plot = fig,
        device = "png",
        bg = "white",
        dpi = 1200)
```

Test adding more chronograms with a loop:

```{r}
number <- 100 # number of chronograms to plot
chronogram_pngs <- paste0("../figures/opentree_chronograms/chronogram_", 
                       1:number, ".png")
set.seed(seed = 1976452)
x_samples <- sample(x = seq(-1, 0, by = 0.01), 
                    size = number, 
                    replace = FALSE)
x_samples
y_samples <- sample(x = seq(0, 9, by = 0.08), 
                    size = number, 
                    replace = FALSE)

scale_samples <- sample(x = seq(0.2, 1.5, by = 0.01), 
                        size = number, 
                        replace = TRUE)

ggdraw(xlim = c(-1,1),
                ylim = c(0, 10)) +
  mapply(draw_image, 
         image = chronogram_pngs, 
         x = x_samples, y = y_samples, 
         scale = scale_samples) -> fig

ggsave2(filename =  "../figures/opentree_chronograms/all_chronograms_test.png",
        plot = fig,
        device = "png",
        bg = "white",
        dpi = 1200)
```
Loving the output, but still a lot of random overlap.
I think I know how I can reduce the iverlap, but need some time to think and implement it.

For now, I am going to save the individual chronograms in different sizes with `draw_image()`, and then manually arrange them in ppt.

```{r}
chronogram_pngs <- paste0("../figures/opentree_chronograms/chronogram_", 
                       1:length(opentree_chronograms$trees), ".png")
```

```{r}
# Initial idea and implementation
# 5 chr at 2
# 10 chronograms at 1.4 to 1.5
# 20 chr at 1.2 to 1.4
# 30 chr at 0.9 to 1.2
# 40 chr at 0.5 to 0.9
# 80 chr at 0.2 to 0.5
# remaining 253 - 185 = 68 chr at same minimum

# I increased scale for better resolution
scale_samples <- vector(mode = "numeric", length = length(chronogram_pngs))
scale_samples[1:5] <- 3.7
set.seed(seed = 1976452)
scale_samples[6:15] <- sample(x = seq(3, 3.5, by = 0.01), 
                        size = 10, 
                        replace = TRUE)
scale_samples[11:30] <- sample(x = seq(2.5, 2.99, by = 0.01), 
                        size = 20, 
                        replace = TRUE)
scale_samples[31:60] <- sample(x = seq(2, 2.49, by = 0.01), 
                        size = 30, 
                        replace = TRUE)
scale_samples[61:100] <- sample(x = seq(1.5, 1.99, by = 0.01), 
                        size = 40, 
                        replace = TRUE)
scale_samples[101:180] <- sample(x = seq(1.1, 1.49, by = 0.01), 
                        size = 80, 
                        replace = TRUE)
scale_samples[181:length(scale_samples)] <- 1
```

```{r}
set.seed(seed = 1976452)

# Plotting area goes from -0.4 to 0 in x axis, when xlim = c(-1,0)
x_samples <- sample(x = seq(-0.4, 0.4, by = 0.01), 
                    size = 253, 
                    replace = TRUE)

# Plotting area goes from 0.5 to 8.5 in y axis, when xlim = c(-1,0)
y_samples <- sample(x = seq(0.5, 19.0, by = 0.07), 
                    size = 253, 
                    replace = FALSE)
```

Now, rescaling and replacing on the canvas.

Testing some positions:
```{r eval = FALSE}
x_samples[5] # original value is -0.52, but that makes the chronogram go over plotting area, tried different values: anything below -0.4 does not work
x_samples[1] # original value is -0.1, and that makes the chronogram go in the middle, so we can get x coordinates up to 0.4
y_samples[6] # original value is 8.93, so max y values coordinate should be 8.5
y_samples[15] # 8.21
y_samples[28] # 8.33
y_samples[40] # 8.24
y_samples[48] # 8. 39 all these have values above 8, so cap to that
opentree_chronograms$trees[44]
# 107 and 108, and almost everything after that have very low res; and they are too small
# smaller scale should be 0.5 or 0.4 for these size of canvas
x_samples[2]
y_samples[2]
i <- 1
```


```{r}
for (i in seq(scale_samples)) {
  print(i)
  ggdraw(c(-1,1),
        ylim = c(0, 20)) +
  draw_image(image = chronogram_pngs[i], 
             x = x_samples[i],
             y = y_samples[i],
             scale = scale_samples[i]) -> fig

  ggsave2(filename =  
        paste0("../figures/opentree_chronograms/resized/chronogram", 
               i, ".png"),
        plot = fig,
        device = "png",
        bg = "transparent",
        dpi = 1200,
        width = 7.5,
        height = 10.83)
}

```

Tweaking some positions:

```{r}
x_samples[5] <- -0.14
y_samples[5] <- 4.105
i <- 5
x_samples[6] <- 0.1
y_samples[6] <- 1.05
i <- 6
x_samples[9] <- -0.39
y_samples[9] <- 1.06
i <- 9
x_samples[10] <- -0.38
y_samples[10] <- 13.17
i <- 10
x_samples[11] <- 0.4
i <- 11
y_samples[12] <- 18 # 18.91 gets cut off on the top
i <- 12
x_samples[15] <- -0.12
y_samples[15] <- 0.7
i <- 15
y_samples[16] <- 15.3
i <- 16
y_samples[17] <- 6.5 # move down
i <- 17
x_samples[18] <- -0.01 #move to the right, probably 0
i <- 18
x_samples[19] <- 0.4 # move to the right a bit
y_samples[19] <- 5.15
i <- 19
x_samples[21] # move to the center
x_samples[22] <- #move down and to the right, closer to the lower right quadrant
x_samples[23] <- # move all the way to the right and up a bit
x_samples[24] <- 
y_samples[24]
  x_samples[25] <- #move all the way to the left
x_samples[26] <- # move all the way to the left down corner
x_samples[27] <- # move to the right
```

```{r}
ggdraw(c(-1,1),
      ylim = c(0, 20)) +
draw_image(image = chronogram_pngs[i], 
           x = x_samples[i],
           y = y_samples[i],
           scale = scale_samples[i]) -> fig
ggsave2(filename =  
      paste0("../figures/opentree_chronograms/resized/chronogram", 
             i, ".png"),
      plot = fig,
      device = "png",
      bg = "transparent",
      dpi = 1200,
      width = 7.5,
      height = 10.83)
```

```{r}
ggdraw(c(-1,1),
      ylim = c(0, 20)) +
  mapply(draw_image, 
         image = chronogram_pngs, 
         x = x_samples,
         y = y_samples,
         scale = scale_samples) -> fig
ggsave2(filename = "../figures/opentree_chronograms/all_chronograms_resized.png",
      plot = fig,
      device = "png",
      bg = "transparent",
      dpi = 1200,
      width = 7.5,
      height = 10.83)
```

